---
title: "RERDDAP + NERACOOS Buoys"
format: html
editor: visual
---

## NERACOOS Buoys

Neracoos buoy data can be found at the NERACOOS ERDDAP server [http://www.neracoos.org/erddap/tabledap/index.html](http://www.neracoos.org/erddap/tabledap/index.html?page=1&itemsPerPage=1000)

```{r}
library(tidyverse)

# Package to interface with erddap
library(rerddap)

## NOAA IOOS NERACOOS - old URL
url <- "http://www.neracoos.org/erddap/"

# Buoy E
# Catalog: http://www.neracoos.org/erddap/tabledap/index.html

# check variables in dataset
info("E01_sbe37_all", url = url)$variables

# pull data
buoy_e <- tabledap(
  x = "E01_sbe37_all", 
  url = url, 
  fields = c("station", "time","depth", "temperature", "temperature_qc"))

buoy_e %>% 
  filter(temperature_qc == 0) %>% 
  ggplot(aes(time, temperature)) +
  geom_line(linewidth = 0.5, color = "gray60") +
  facet_wrap(~depth)

```

They are migrating to a newer server atm, which can be found here:\
<https://data.neracoos.org/erddap/index.html>

```{r}

# This is where the NERACOOS Server can be found
# Catalog: https://data.neracoos.org/erddap/index.html

# We then find the dataset IDs for what we want under the tabledap link

# Ex. Buoy M - 50meter, from data.neracoos.org - new URL
url <- "http://data.neracoos.org/erddap/"
buoy_m_50m <- tabledap(
  x = "M01_ocean_050m", 
  url = url, 
  fields = c("station", "time", "longitude", "latitude", "depth", "temperature"))

# Plot
ggplot(buoy_m_50m, aes(time, temperature)) +
  geom_line(linewidth = 0.5, color = "gray60")
```

### Helping with a Data Request

It is not uncommon for people to have difficulty finding the access points for data on ERDDAP or being unfamiliar with how to get the data for what they need, myself included.

This most recent data request was for:

```{r}

# If there are a number of buoys we want data for we can run them togther this way:

# This is where the NERACOOS Server can be found
# Catalog: https://data.neracoos.org/erddap/index.html

# We then find the dataset IDs for what we want under the tabledap link:
# https://data.neracoos.org/erddap/tabledap/index.html

# If we want the same data from a couple datasets we can put them in a list and reuse code
dataset_ids <-  list(
  "B" = "B01_ocean_001m",
  "E" = "E01_ocean_001m",
  "F" = "F01_ocean_001m",
  "I" = "I01_ocean_001m")

# This will pull data for each and assemble a single table
befi_surface_temps <- map_dfr(
  .x = dataset_ids,
  .f = function(dataset_id_x){
    tabledap(
      x = dataset_id_x,
      url = url,
      fields = c("station", "time", "longitude", "latitude", 
                 "depth", "temperature"), 
      # This is a constraint flag removing qc_agg values > 3
      'temperature_qc_agg<=3') %>%  
      as_tibble()
  }
)

# Plot them to check the data
befi_surface_temps %>% 
  ggplot(aes(time, temperature)) + geom_line() + facet_wrap(~station)


# Export
write_csv(befi_surface_temps, here::here("local_data/random_requests/neracoos_befi_surface_temps.csv"))

```
