---
title: "2021: Gulf of Maine's Warmest Year on Record"
author: 
  - first_name: "Adam"
    last_name: "Kemberling"
    url: https://github.com/adamkemberling
    affiliation: Gulf of Maine Research Institute
    affiliation_url: https://www.gmri.org
description: | 
  Characteristics of a record setting year for ocean temperatures
date: "Updated on: `r Sys.Date()`"
editor_options: 
  chunk_output_type: console
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}

# knitr options
knitr::opts_chunk$set(echo = FALSE, 
                      message = F, 
                      warning = F, 
                      comment = "", 
                      fig.align = "center", 
                      fig.height = 6, 
                      fig.width = 8)

####  Packages  ####
library(lubridate)
library(raster)
library(here)
library(rnaturalearth)
library(scales)
library(sf)
library(stars)
library(gmRi)
library(heatwaveR)
library(gt)
library(gtExtras)
library(patchwork)
library(tidyverse)
library(ggthemes)
library(ggstream)
library(ggforce)
library(geomtextpath)
library(ggHoriPlot)
library(directlabels)
library(tidytext)

# Support Functions
suppressPackageStartupMessages( source(here("R/oisst_support_funs.R"), verbose = FALSE) )
suppressPackageStartupMessages( source(here("R/temp_report_support.R"), verbose = FALSE) )

#box paths
oisst_path <- cs_path("res", "OISST/oisst_mainstays")


# File paths for various extents based on "apershing_gulf_of_maine"
region_paths <- get_timeseries_paths(region_group = "gmri_sst_focal_areas", 
                                     mac_os = "mojave")

# Timeseries Path
timeseries_path <- region_paths[["apershing_gulf_of_maine"]][["timeseries_path"]]

# Polygon Path
poly_path <- region_paths[["apershing_gulf_of_maine"]][["shape_path"]]
region_extent <- read_sf(poly_path)

# Pull extents for the region for cropping
crop_x <- st_bbox(region_extent)[c(1,3)]
crop_y <- st_bbox(region_extent)[c(2,4)]

# Expand the area out to see the larger patterns
crop_x_expanded <- crop_x + c(-2.25, 2.25)
crop_y_expanded <- crop_y + c(-0.75, 0.75)

# Make a new extent for cropping to larger area
sfc <- st_sfc(st_polygon(list(
  rbind(c(crop_x_expanded[[1]], crop_y_expanded[[1]]),  
        c(crop_x_expanded[[1]], crop_y_expanded[[2]]), 
        c(crop_x_expanded[[2]], crop_y_expanded[[2]]), 
        c(crop_x_expanded[[2]], crop_y_expanded[[1]]), 
        c(crop_x_expanded[[1]], crop_y_expanded[[1]])))))
sfc <- st_as_sf(sfc)


# Coastline Shapes
new_england <- ne_states("united states of america", returnclass = "sf")
canada      <- ne_states("canada", returnclass = "sf")
world_sf    <- ne_countries(returnclass = "sf")
greenland   <- ne_states(country = "greenland", returnclass = "sf")

# Adding Logo to plots
logo_path <- paste0(system.file("stylesheets", package = "gmRi"), "/gmri_logo.png")
lab_logo <- magick::image_read(logo_path)


# Box Location to save hi-res images
save_location <- cs_path("root", "GoM Seasonal Climate Update/annual_update_figures")
```

`r gmRi::use_gmri_style_rmd(css = "gmri_rmarkdown.css")`


# 2021 Sea Surface Temperature

##### An Exceptional Year

2021 was an exceptionally warm year for the Gulf of Maine. It had the warmest fall on record for the area, and the second warmest summer. Much of the year the region experienced what would be considered "heatwave conditions". After compiling sea surface temperature data for the entire year we can now conclude that 2021 was the hottest year on Record for the Gulf of Maine.

```{r load timeseries}
# Load the timeseries
region_timeseries <- read_csv(timeseries_path, col_types = cols(), guess_max = 1e6)

# Clean up the data - add labels
region_timeseries <- region_timeseries %>% 
  distinct(time, .keep_all = T) %>% 
  supplement_season_info() %>% 
  filter(year %in% c(1982:2021))


# Summarize by year to return mean annual anomalies and variance
annual_summary <- region_timeseries %>% 
  group_by(year) %>% 
  temperature_summaries() %>% 
  mutate(yr_as_dtime = as.Date(paste0(year, "-07-02")),
         anom_direction = ifelse(area_wtd_anom > 0, "Hot", "Cold"))


# Get heatwave statuses for each day:
# Uses area weighted sst by default
region_hw <- pull_heatwave_events(
  temperature_timeseries = region_timeseries, 
  threshold = 90, 
  clim_ref_period = c("1982-01-01", "2011-12-31")) %>% 
  supplement_hw_data() %>% 
  filter(doy != 366)



# # Global Temperature Anomaly Rates
global_anoms <- read_csv(
    paste0(oisst_path, "global_timeseries/global_anoms_1982to2011.csv"), 
    guess_max = 1e6,
    col_types = cols()) %>% 
  mutate(year = year(time)) %>% 
  filter(between(year, 1982, 2021))

# summarize by year again
global_summary <- global_anoms %>% 
  group_by(year) %>% 
  temperature_summaries() %>% 
  mutate(yr_as_dtime = as.Date(paste0(year, "-07-02")),
         anom_direction = ifelse(area_wtd_anom > 0, "Hot", "Cold"))


####  Run Warming Rate regressions  ####
rate_data <- list(
  "GoM All F" = get_decadal_rates(temp_df = annual_summary, 
                                temp_col = "area_wtd_f", 
                                year_col = "year", 
                                year_lim = c(1982, 2021), 
                                area_name = "GoM", 
                                degree_c = F),
  "GoM All C" = get_decadal_rates(temp_df = annual_summary, 
                                temp_col = "area_wtd_sst", 
                                year_col = "year", 
                                year_lim = c(1982, 2021), 
                                area_name = "GoM", 
                                degree_c = T),
  "GoM 15" = get_decadal_rates(temp_df = annual_summary, 
                                temp_col = "area_wtd_f", 
                                year_col = "year", 
                                year_lim = c(2007, 2021), 
                                area_name = "GoM", 
                                degree_c = F),
  "Global All" = get_decadal_rates(temp_df = global_summary, 
                                temp_col = "area_wtd_f", 
                                year_col = "year", 
                                year_lim = c(1982, 2021), 
                                area_name = "Global", 
                                degree_c = F))


```




## Departure from Historic Conditions

Thanks to some long-running temperature monitoring programs we are able to track monthly sea surface temperatures as far back as 1850. From this view we can see that recent temperatures are the warmest we've ever seen. 2021 setting the new record for the warmest temperature for the region.




```{r}
#| eval = TRUE,
#| fig.alt = "Long term temperature record for the Gulf of Maine"

# From temp_report_support.R
ersst_yr <- load_annual_ersst()

# Benchmark the averages temp and maximum temps
long_avg <- mean( filter(ersst_yr, between(yr, 1860, 2000)) %>% pull(sst) , na.rm = T)
long_max <- max( filter(ersst_yr, between(yr, 1860, 2000)) %>% pull(sst) , na.rm = T)

# Make Splines
ersst_smooth <- as.data.frame(spline(ersst_yr$yr, ersst_yr$sst))
ersst_smooth_f <- as.data.frame(spline(ersst_yr$yr, ersst_yr$sst_f))


# Fahrenheit Plot
ersst_long_term_p <- ggplot(data = ersst_yr, aes(yr, sst_f)) +
  geom_line(group = 1, alpha = 0.4, linetype = 1, color = gmri_cols("gmri blue")) +
  geom_point(size = 1, color = gmri_cols("gmri blue")) +
  geom_point(data = filter(ersst_yr, sst > long_max), aes(yr, sst_f), color = "darkred", size = 1.5) +
  geom_mark_ellipse(aes(yr, sst_f, filter = sst > long_max,
                        description = "Temperatures in 6 of the 10 Last Years Above 1951 Record",
                        label = "Recent Extremes"), label.fontsize = 9,
                    color = "darkred", label.fill = "white", label.colour = "black", con.colour = "black") +
  geom_mark_ellipse(aes(yr, sst_f, filter = sst == long_max, 
                        label = str_c(yr,": ", round(sst_f, 2), " \u00b0F"), 
                        description = "Former Hottest Temperature on Record"), 
                    label.fontsize = 9,  label.buffer = unit(0.25, "lines"),
                    color = "darkred", linetype = 2) +
  scale_color_gmri() +
  theme_gmri() +
  scale_x_continuous(expand = expansion(add = c(2,10))) +
  scale_y_continuous(expand = expansion(add = c(0.5,5)),
    labels = number_format(suffix = " \u00b0F")) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5))  +
  theme(legend.position = "bottom") +
  labs(x = "", y = "Sea Surface Temperature", 
       color = "Temperature Data Record",
       title = "Long-Term Temperature Record for the Gulf of Maine",
       subtitle = "Current temperatures above 150-year highs",
       caption = "Data Source: ERSSTv5 Monthly Sea Surface Temperature")



ersst_long_term_p
```

## Newer Technology: Higher-Quality Data

In 1981 the availability of a higher resolution temperature resource became available in the form of the Optimum Interpolation Sea Surface Temperature (OISSTv2) Satellite Data Record. Once the OISSTv2 data became available we gained access to a higher resolution record, and to the ability to validate past records through comparison. These two data sources now overlap for a period of 42 years and we can see visually that they are not far off:


```{r oisst rates}

# Get yearly averages for OISST
oisst_yr <- region_hw %>% 
  filter(between(yr, 1982, 2021)) %>% 
  group_by(yr) %>% 
  summarise(sst = mean(sst), 
            sst_anom = mean(sst_anom, na.rm = T),
            sst_f = as_fahrenheit(sst, "temperature"),
            anom_f = as_fahrenheit(sst_anom, "anomalies"),
            .groups = "drop")

# Average Distance between OISST and ERSST in GOM
temp_diffs <- ersst_yr %>% 
  rename(ersst_sst = sst,
         ersst_sst_f = sst_f,
         ersst_anom = sst_anom,
         ersst_anom_f = anom_f) %>% 
  inner_join(oisst_yr, by = "yr") %>% 
  mutate(temp_diff = ersst_sst - sst,
         temp_diff_f = ersst_sst_f -sst_f)

# Get temperature differences
temp_difference <- temp_diffs %>% 
  pull(temp_diff) %>% mean() %>% round(2)
temp_difference_f <- temp_diffs %>% 
  pull(temp_diff_f) %>% mean() %>% round(2)


# Make Splines
oisst_smooth <- as.data.frame(spline(oisst_yr$yr, oisst_yr$sst))
oisst_smooth_f <- as.data.frame(spline(oisst_yr$yr, oisst_yr$sst_f))


# Degrees F
ersst_v_oisst <- ggplot() +
  geom_line(data = ersst_smooth_f, aes(x, y, color = "ERSSTv5"),
            group = 1, alpha = 0.4, linetype = 1) +
  geom_point(data = ersst_yr, aes(yr, sst_f, color = "ERSSTv5"), 
             size = 1) +
 geom_line(data = oisst_smooth_f, aes(x, y, color = "OISSTv2"),
            group = 1, alpha = 0.4, linetype = 1) +
  geom_point(data = oisst_yr, aes(yr, sst_f, color = "OISSTv2"), 
             size = 1) +
  scale_color_gmri() +
  theme_gmri() +
   scale_y_continuous(
     #expand = expansion(add = c(0.5,5)),
     labels = number_format(suffix = " \u00b0F")) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5)) +
  facet_zoom(xlim = c(1982, 2021)) +
  theme(legend.position = "bottom")  +
  labs(x = "", y = "Sea Surface Temperature", 
       color = "Temperature Data Source",
       title = "Long-Term Temperature Record for the Gulf of Maine",
       subtitle = "OISSTv2 Data Availability Overlap with ERSSTv5")



ersst_v_oisst
```



Because the ERSST data is of a coarser resolution (monthly measurements at a 0.5 x 0.5 degree grid) and does not capture the warmer inshore dynamics (a bias to show colder temperature). The OISSTv2 temperatures in this area are on average warmer. **Across the years where data from both sources is available, ERSSTv5 is on average `r temp_difference`$^oC$ or `r temp_difference_f`$^oF$ Colder than OISSTv2 data.** 



## Placing 2021 into Context

```{r}
# Get the overall average and build labels
all_year_avgs <- annual_summary %>% 
  summarise(avg_temp_c = mean(area_wtd_sst),
            avg_temp_f = mean(area_wtd_f),
            avg_anom_c = mean(area_wtd_anom),
            avg_anom_f = mean(area_wtd_anom_f),            
            .groups = "drop") %>% 
  mutate(temp_label = str_c(round(avg_temp_c, 2), " \u00b0C  |  ", round(avg_temp_f, 2), " \u00b0F"))

```


2021 was the hottest year on record for the Gulf of Maine. It was on average `r all_year_avgs$temp_label` above the 1982-2011 average for the area, shifting the 1982-2021 warming rate to `r str_c(rate_data[["GoM All F"]][["decadal_rate"]], "\u00b0F / Decade")`. 


```{r}


# Build label for 2021
anom_2021 <- annual_summary %>% filter(year == 2021) %>% select(area_wtd_anom, area_wtd_anom_f)
anom_label <- str_c(round(anom_2021[1], 2), " \u00b0C  |  ", round(anom_2021[2], 2), " \u00b0F")


# Assemble plot highlighting 2021 characteristics
annual_temps_plot <- ggplot(annual_summary, aes(year, area_wtd_f)) +
  geom_hline(yintercept = all_year_avgs$avg_temp_f, color = "darkblue", linetype = 2) +
  geom_line(linetype = 3, size = 0.5, alpha = 0.5) +
  geom_smooth(formula = y ~ x, 
              linetype = 1,
              method = "lm", 
              color = "darkred", 
              se = FALSE) +
  geom_point(color = "gray20", size = 1, alpha = 0.8) +
  geom_mark_ellipse(
    data = data.frame(x = 2016, y = all_year_avgs$avg_temp_f),
    aes(x, y, label = all_year_avgs$temp_label), 
    description = "1982-2021 Average Temperature", 
    label.colour = "darkblue",
    color = "transparent", 
    con.colour = "darkblue", 
    label.fill = "transparent", 
    label.fontsize = 9, 
    label.buffer = unit(0.25, "cm")) +
  geom_mark_ellipse(
    aes(filter = year == 1994, 
        label = str_c(rate_data$`GoM All F`$decadal_rate, " \u00b0F / Decade")), 
    color = "transparent", 
    description = str_c("40-Year Warming Trend"), 
    label.colour = "darkred",
    con.colour =  "darkred",
    label.buffer = unit(1.35, "cm"), 
    label.fill = "transparent", 
    label.fontsize = 9) +
  geom_mark_ellipse(
    aes(filter = year == 2021, 
        label = anom_label),
    color = "darkred",
    linetype = 2,
    description = str_c("Temperature Above\n1982-2011 climatological average"), 
    label.colour = "darkred",
    con.colour =  "darkred",
    label.fill = "transparent", 
    label.buffer = unit(0.15, "cm"), 
    label.fontsize = 9) +
  scale_y_continuous(limits = c(48, 56),
                   labels =  number_format(suffix = " \u00b0F")) +
  labs(x = "Year", y = "Average Annual Temperature")


# display plot
annual_temps_plot
```




## Warming Relative to the World

While this report focuses on the Gulf of Maine, there is a recognition that this region is not unique in the fact that its temperatures are rising. The Earth as a whole is warming, and that change is being driven by anthropogenic influences to our atmosphere. The impacts of these changes however are not distributed evenly, and there are regions like ours that are warming at rates above the global average.


```{r warming rates info}

# Build Regression Equation Labels
eq_all_c  <- rate_data[["GoM All C"]][["eq_label"]]
eq_all    <- rate_data[["GoM All F"]][["eq_label"]]
eq_15     <- rate_data[["GoM 15"]][["eq_label"]]
eq_global <- rate_data[["Global All"]][["eq_label"]]



# Line color formatting
line_colors <- setNames(
  as.character(c(
    gmri_cols("gmri blue"), 
    gmri_cols("gmri green"))),  
  c(eq_all, eq_global))



# Single Line Plot
(temp_simplified <- ggplot(data = annual_summary, aes(year, area_wtd_anom_f)) +

  # Overlay yearly means
  geom_line(color = "gray10", size = 1.5, linetype = 1) +
  geom_point(color = "gray10", size = 1.5) +
  geom_smooth(
     data = mutate(global_summary, `Warming Rate` = eq_global),
     aes(color = `Warming Rate`),
    method = "lm",
    formula = y ~ x, se = F,
    alpha = 0.90,
    size = 1.5,
    linetype = 2) +
  geom_smooth(
    data = annual_summary %>% 
      mutate(`Warming Rate` = eq_all),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    color = gmri_cols("gmri blue"),
    alpha = 0.90,
    size = 1.5,
    linetype = 2) + 
   scale_color_manual(values = line_colors) +
   scale_x_continuous(limits = c(1978, 2023), expand = expansion(add = c(0,0))) +
   scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
   labs(title = "Gulf of Maine:",
        subtitle = "Annual Sea Surface Temperature Anomalies",
        x = "Year", y = "Sea Surface Temperature Anomaly",
        caption = "Anomalies calculated using 1982-2011 reference period.") +
   theme(legend.position = c(0.25, 0.85),
         legend.background = element_rect(color = "black", fill = "white"))

)




# # Don't Plot the Gulf of Maine temps as a line
# # Trend lines and points colored for both gorups:
# ggplot(data = global_summary, aes(year, area_wtd_anom_f)) +
# 
#   # Overlay yearly means
#    geom_point(data = annual_summary, aes(year, area_wtd_anom_f), color = gmri_cols("gmri blue"), size = 1) +
#   geom_smooth(
#     data = annual_summary %>% 
#       mutate(`Warming Rate` = eq_all),
#     method = "lm",
#     formula = y ~ x, se = T,
#     aes(color = `Warming Rate`, fill = `Warming Rate`),
#    alpha = 0.20,
#     size = 1,
#     linetype = 1) + 
#   geom_point(data = global_summary, aes(year, area_wtd_anom_f), color = gmri_cols("gmri green"), size = 1) +
#   geom_smooth(
#      data = mutate(global_summary, `Warming Rate` = eq_global),
#      aes(color = `Warming Rate`, fill = `Warming Rate`),
#     method = "lm",
#     formula = y ~ x, se = T,
#     alpha = 0.20, 
#     size = 1,
#     linetype = 1) +
#   scale_color_manual(values = line_colors) +
#   scale_fill_manual(values = line_colors) +
#   scale_x_continuous(limits = c(1978, 2024), expand = expansion(add = c(0,0))) +
#   scale_y_continuous(labels =  number_format(suffix = " \u00b0F"),
#                       limits = c(-1.5, 4.5)) +
#    labs(title = "Global Temperatures:",
#         subtitle = "Annual Sea Surface Temperature Anomalies",
#         x = "Year", y = "Sea Surface Temperature Anomaly",
#         caption = "Anomalies calculated using 1982-2011 reference period.") +
#    theme(legend.position = c(0.25, 0.85),
#          legend.background = element_rect(color = "black", fill = "white"))

```



```{r just gom temps, eval = F}

# Fahrenheit
(gom_f <- ggplot(data = annual_summary, aes(year, area_wtd_anom_f)) +

  # Overlay yearly means
  geom_line(color = "gray10", size = 1.5, linetype = 1) +
  geom_point(color = "gray10", size = 1.5) +
  geom_smooth(
    data = annual_summary %>% 
      mutate(`Warming Rate` = eq_all),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    alpha = 0.90,
    size = 1.5,
    linetype = 2) + 
   scale_color_gmri() +
   scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
   labs(title = "Gulf of Maine:",
        subtitle = "Annual Sea Surface Temperature Anomalies",
        x = "Year", y = "Sea Surface Temperature Anomaly",
        caption = "Anomalies calculated using 1982-2011 reference period.")  +
   theme(legend.position = c(0.25, 0.85),
         legend.background = element_rect(color = "black", fill = "white")))


# Celsius
(gom_c <- ggplot(data = annual_summary, aes(year, area_wtd_anom)) +

  # Overlay yearly means
  geom_line(color = "gray10", size = 1.5, linetype = 1) +
  geom_point(color = "gray10", size = 1.5) +
  geom_smooth(
    data = annual_summary %>% 
      mutate(`Warming Rate` = eq_all_c),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    alpha = 0.90,
    size = 1.5,
    linetype = 2) + 
   scale_color_gmri() +
   scale_y_continuous(labels =  number_format(suffix = " \u00b0C")) +
   labs(title = "Gulf of Maine:",
        subtitle = "Annual Sea Surface Temperature Anomalies",
        x = "Year", y = "Sea Surface Temperature Anomaly",
        caption = "Anomalies calculated using 1982-2011 reference period.")  +
   theme(legend.position = c(0.25, 0.85),
         legend.background = element_rect(color = "black", fill = "white")))



ggsave(plot = gom_f, filename = str_c(save_location, "gom_annual_f.png"),
       dpi = "retina", bg = "white")

# top 20% rankings
ggsave(plot = gom_c, filename = str_c(save_location, "gom_annual_c.png"),
       dpi = "retina", bg = "white")
```



## Warming at an Accelerated Pace

In recent years we've seen temperatures in the Gulf of Maine (and adjacent waters) to accelerate rapidly. Scientists believe this is partly due to changes in the behavior of the Gulf Stream Current which brings warm, salty water from the Gulf of Mexico & Florida Northward along the east coast of the US and eventually flowing Northeast past Nova Scotia.

Scientists believe a regime shift has occurred in the Northwest Atlantic as a result of a Northward shift in the Gulf Stream, and a general weakening in the Atlantic Meridional Overturning Circulation Pattern (AMOC). This Northward shift in the Gulf stream does two things: 1) It brings more warm/salty water into the Gulf of Maine, & 2) It obstructs the flow of cold and relatively fresh water from entering the Gulf of Maine from the Labrador current and Scotian Shelf Water (SSW).

These relative contribution of these two impacts may vary through time, but the result (in surface temperature terms) is similar: **a warmer Gulf of Maine.**

```{r gom full warming, eval = T}

# Line color formatting
line_colors <- setNames(
  as.character(c(
    gmri_cols("gmri blue"), 
    gmri_cols("orange"))),  
  c(eq_all, 
    eq_15))


# Gulf of Maine Warming
gom_all_trend <- annual_summary %>%
  ggplot(aes(year, area_wtd_anom_f)) +
  geom_line(color = "gray10", size = 0.75, linetype = 1) +
  geom_point(color = "gray10", size = 1) +
  geom_smooth(
    data = annual_summary %>%
      mutate(`Warming Rate` = eq_all),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    alpha = 0.90,
    size = 1.5,
    linetype = 1) +
  scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
  scale_color_manual(values = line_colors) +
  labs(title = "Gulf of Maine: Long-Term Warming Trend",
       x = "",
       y = "SST Anomaly",
       caption = "")  +
   theme(legend.position = c(0.25, 0.85),
         legend.background = element_rect(color = "black", fill = "white"))


# Gulf of Maine Warming - last 15 years
gom_recent_trend <- annual_summary %>% 
  ggplot(aes(year, area_wtd_anom_f)) +
  geom_line(color = "gray10", size = 0.75, linetype = 1) +
  geom_point(color = "gray10", size = 1) +
  # Add regression lines
  geom_smooth(
    data = filter(annual_summary, year %in% c(2007:2021)) %>% 
      mutate(`Warming Rate` = eq_15),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    alpha = 0.90,
    size = 1.5,
    linetype = 1) +
  scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
  scale_color_manual(values = line_colors) +
  labs(title = "Gulf of Maine: Rapid Recent Warming",
       x = "", 
       y = "SST Anomaly",
       caption = paste0("Anomalies calculated using 1982-2011 reference period."))  +
   theme(legend.position = c(0.25, 0.85),
         legend.background = element_rect(color = "black", fill = "white"))




# Put them together
(gom_recent_warming <- (gom_all_trend / gom_recent_trend ) + 
  plot_layout(guides = "collect")  & 
  theme(legend.position = "bottom") &
  guides(color = guide_legend(title.position = "top", ncol = 1, title.hjust = 0.5)))
```


## Concerning Temperatures in Recent Years




```{r}
####  Annual Trend Plot  ####
gom_anomaly_bars <- annual_summary %>% 
  ggplot(aes(yr_as_dtime, area_wtd_anom_f)) +
    geom_col(aes(fill = anom_direction), color = "black", size = 0.25, alpha = 0.9) +
    scale_fill_manual(values = c("Hot" = "#CA0020", "Cold" = "#0571B0")) +
    theme_gmri() +
    theme(legend.position = c(0.25, 0.85),
          legend.background = element_rect(fill = "white", color = "black"), 
          legend.key = element_rect(fill = "transparent", color = "transparent")) +
    labs(title = "Gulf of Maine: Rapid Recent Warming",
         subtitle = "Recent Rate More Than Double the 40-Year Average",
         fill = "Anomaly Type",
         x = "", 
         y = "Sea Surface Temperature Anomaly",
         caption = paste0("Anomalies calculated using 1982-2011 reference period."))



gom_anomaly_bars

```


### 2021: Record Daily Temperatures

To add context on the severity of these sustained heatwaves, many of the individual daily temperatures throughout 2021 were themselves the hottest days on record.

```{r}
# Grab the hottest day for each day of the year
hottest_days <- region_hw %>% 
  group_by(doy) %>% 
  slice_max(order_by = sst_anom, n = 1) %>% 
  ungroup() 


# Totel Number in 2021
hottest_2021 <- hottest_days %>% 
  count(yr) %>% 
  filter(yr == "2021") %>% 
  pull(n)


# Daily Record Temperatures
(hottest_days_plot <- hottest_days %>% 
  mutate(is_21 = ifelse(yr == "2021", "2021", "1982-2020"),
         flat_date = as.Date("2021-01-01") + (doy - 1)) %>% 
  ggplot(aes(x = flat_date, y = sst_f, color = is_21)) +
  #geom_segment() +
  #geom_path(group = 1) +
  geom_point() +
  scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(add = c(0,0))) +
  scale_y_continuous(labels = number_format(suffix = deg_f)) +
  scale_color_manual(values = c("2021" = "darkred", "1982-2020" = "gray80")) +
  labs(x = "", 
       y = "Sea Surface Temperature", 
       title = "Hottest Days on Record", 
       subtitle = "Gulf of Maine",
       color = "Record Setting Year")  +
   theme(legend.position = c(0.25, 0.85),
         legend.background = element_rect(color = "black", fill = "white")) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5)) 
  )

```

Of the 365 days in a year, 2021 set a new record temperature on `r hottest_2021` or `r round((hottest_2021/365) * 100, 2)`% of the year. Many of these occurred in clusters of exceptionally warm periods that happened at various points throughout the year.



```{r, fig.height=3}

# What days set records in 2021
records_21 <- hottest_days %>% 
  filter(yr == "2021") %>% #count(month)
  mutate(record_temp = TRUE) %>% 
  select(yr, time, sst, seas, sst_anom, record_temp)


#@ What percent of days in month were records:
region_hw %>% 
  filter(yr == "2021") %>% 
  full_join(records_21) %>% 
  mutate(record_temp = ifelse(record_temp == TRUE, TRUE, FALSE)) %>% 
  group_by(yr, month) %>% 
  summarise(n_days = n(),
            n_records = sum(record_temp, na.rm = T),
            .groups = "drop") %>% 
  mutate(percent_records = (n_records / n_days),
         highlight_oct = ifelse(month == "Oct", "100% Record Temp.", "filler")) %>% 
  ggplot(aes(month, percent_records, fill = highlight_oct)) +
  geom_col() +
  scale_fill_grey() +
  scale_y_continuous(labels = percent_format(), expand = expansion(add = c(0,0))) +
  theme(legend.position = "none") +
  labs(x = "", y = "Percent of Daily Temperatures", title = "2021: Record Temperatures as Percent of Each Month")
```

Six months in 2021 had record setting temperatures on atleast 50% of the days in that month. 2021 was an exceptionally warm year in October, which set record temperatures on every day during that prolonged fall heatwave event.

### 2021: Daily Temperatures and Heatwave Status

```{r}
 # Last Calendar Year
this_yr <- region_hw %>% 
  filter(year(time) == 2021)

#min/max of climate and min/max of 2021
ranges_c <- list(
  "clim_range" = round(range(this_yr$seas), 2),
  "temp_range" = round(range(this_yr$sst), 2))
ranges_f <- map(ranges_c, function(x){ round(as_fahrenheit(x, data_type = "temperature"), 2) })


# Differences in climate and temperature ranges
diffs_c <- list(
  "anom_range" = round(range(this_yr$sst_anom), 2),
  "clim_diff"  = round(diff(ranges_c$clim_range), 2),
  "temp_diff"  = round(diff(ranges_c$temp_range), 2),
  "anom_diff"  = round(diff(range(this_yr$sst_anom)), 2))
diffs_f <- map(diffs_c, function(x){ round(as_fahrenheit(x, data_type = "anomalies"), 2) })
```


The Gulf of Maine has a seasonal temperature cycle that experiences its lowest temperatures around March, and its peak summer highs in August. The difference between these two seasons is typically `r diffs_f$clim_diff`$^oF$ or `r diffs_c$clim_diff`$^oC$. In 2021 the temperature range was `r ranges_f$temp_range[1]` - `r ranges_f$temp_range[2]`$^oF$ or `r diffs_f$temp_diff`$^oF$, with temperature anomaly values ranging from `r diffs_f$anom_range[1]` - `r diffs_f$anom_range[2]`$^oF$.



```{r single year lineplot}
#| fig.alt = "Timelines of observed temperature and observed temperature anomalies for the year of 2021, and how they compare to the mean and 90th percentile of the climate reference period"


# Set colors by name
color_vals <- c(
  "Sea Surface Temperature" = "#0571B0",
  "Heatwave Event"          = "darkred",
  "MHW Threshold"           = "coral3",
  "Daily Climatology"       = "gray30")

# Set linetypes manually
linetype_key <- c(
  "Sea Surface Temperature Anomaly" = 1,
  "Heatwave Event"                  = 1,
  "MHW Threshold"                   = 3,
  "Daily Climatology"               = 2)



# Plot the last 365 days
(hw_temp_p <- ggplot(this_yr, aes(x = time)) +
    geom_segment(aes(x = time, xend = time, 
                     y = seas_f, yend = sst_f), 
                 color = "#0571B0", alpha = 0.25) +
    geom_segment(aes(x = time, xend = time, 
                     y = mhw_thresh_f, yend = hwe_f), 
                 color = "darkred", alpha = 0.25) +
    geom_line(aes(y = sst_f, color = "Sea Surface Temperature")) +
    geom_line(aes(y = hwe_f, color = "Heatwave Event")) +
    geom_line(aes(y = mhw_thresh_f, color = "MHW Threshold"), lty = 3, size = .5) +
    geom_textpath(aes(y = seas_f), color = "gray30", label = "Climatological Mean", hjust = 0.5, lty = 2) +
    scale_color_manual(values = color_vals) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(mult = c(0,0))) +
    scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
    guides(color = guide_legend(override.aes = list(linetype = linetype_key), nrow = 1)) +
    theme(legend.title = element_blank(),
          legend.position = "bottom") +
    labs(x = "", 
         y = "Sea Surface Temperature")
)
```

The highest temperatures anomalies spiked in the last days of June, capping off an already warm start to the year. A very-warm August and steady, above-average temperatures from early September through the end of the year helped solidify 2021 as the hottest year on record for the Gulf of Maine.



### 2021: Heatwave Events

```{r}

# Get the Event numbers
events_21 <-  this_yr %>% 
  drop_na(mhw_event_no) %>% 
  distinct(mhw_event_no) %>% 
  pull(mhw_event_no)

events_num <-  length(events_21)


# Pull their data, including the days before and after
events_dat <- filter(region_hw, mhw_event_no %in% as.character(events_21),
                     yr == 2021)
# Days in HW state
days_hw <- nrow(events_dat)

# How long on average
avg_duration <- round(days_hw / events_num, digits = 0)
```

During 2021 there were `r events_num` marine heatwave events,  lasting a total of `r days_hw`. These events include one that carried over from the previous year and another that has persisted into 2022. The average duration for these events was `r avg_duration` days (excluding days in 2020 & 2022).



```{r single year anom lineplot}
# Plot the last 365 days - anomaly scale
(hw_anom_p <- this_yr %>% 
  mutate(
    anom_thresh = mhw_thresh - seas,
    anom_thresh_f = mhw_thresh_f - seas_f,
    anom_hwe = hwe - seas,
    anom_hwe_f = hwe_f - seas_f) %>% 
ggplot(aes(x = time)) +
    geom_segment(aes(x = time, xend = time, 
                     y = 0, yend = anom_f), 
                 color = "#0571B0", alpha = 0.25) +
    geom_segment(aes(x = time, xend = time, 
                       y = anom_thresh_f, yend = anom_hwe_f), 
                 color = "darkred", alpha = 0.25) +
    geom_line(aes(y = anom_f, color = "Sea Surface Temperature Anomaly")) +
    geom_line(aes(y = anom_hwe_f, color = "Heatwave Event")) +
    geom_line(aes(y = anom_thresh_f, color = "MHW Threshold"), lty = 3, size = .5) +
    geom_line(aes(y = 0, color = "Daily Climatology"), lty = 2, size = .5) +
   scale_color_manual(values = color_vals) +
    scale_linetype_manual(values = linetype_key, guide = "none") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(mult = c(0,0))) +
    guides(color = guide_legend(override.aes = list(linetype = linetype_key), nrow = 1)) +
    theme(legend.title = element_blank(),
          legend.position = "bottom") +
    scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
    labs(x = "", 
         y = "Temperature Anomaly", 
         caption = paste0("Climate reference period :  1982-2011")))


# Show figure
year_timeline <- (hw_temp_p + theme(legend.position = "none")) / hw_anom_p

```




## Long-Term Heatwave Seasonality

When looking at the last ~40 years of marine heatwave events here in the Gulf of Maine. We see that the frequency, duration, and intensity of heatwave events has increased. This year was no exception with `r days_hw` days meeting the criteria for heatwave status.


```{r}
#| fig.alt = "Heatmap of heatwave temperatures and heatwave days"

# Set up the date within a year
base_date <- as.Date("2000-01-01")
grid_data <- region_hw %>% 
  mutate(year = year(time),
         yday = yday(time),
         flat_date = as.Date(yday-1, origin = base_date),
         yr_rev = factor(year),
         yr_rev = fct_rev(yr_rev))

# Prep the legend title
guide_lab <- "Sea Surface Temperature Anomaly"

# Color limit for palettes
temp_limits <- c(-5, 5)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0F")


# Assemble heatmap plot
heatwave_heatmap <- ggplot(grid_data, aes(x = flat_date, y = year)) +
  
  # background box fill for missing dates
  geom_rect(xmin = base_date, xmax = base_date + 365, 
            ymin = min(grid_data$year) - .5, ymax = max(grid_data$year) + .5, 
            fill = "gray75", color = "transparent") +
  
  # tile for sst colors
  geom_tile(aes(fill = anom_f)) +
  # points for heatwave events
  geom_point(data = filter(grid_data, mhw_event == TRUE),
             aes(x = flat_date, y = year), size = .25)  +
  scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = c(0,0)) +
  scale_y_continuous(limits = c(1980.5, 2021.5), expand = c(0,0)) +
  scale_fill_distiller(palette = "RdBu", 
                     na.value = "transparent", 
                     limit = temp_limits, 
                     oob = scales::squish,
                     breaks = temp_breaks, 
                     labels = temp_labels) +
  
  #5 inches is default rmarkdown height for barheight
  guides("fill" = guide_colorbar(title = guide_lab, 
                                 title.position = "right", 
                                 title.hjust = 0.5,
                                 barheight = unit(4.8, "inches"), 
                                 frame.colour = "black", 
                                 ticks.colour = "black")) +  
  theme(legend.title = element_text(angle = 90)) +
  labs(title = "Temperature Anomalies and Marine Heatwave Events",
       #subtitle = "Marine Heatwave Events Lasting Months Common Since 2010",
       x = "", 
       y = "",
       "\nClimate reference period : 1982-2011",
       caption = "Marine heatwave is an event hotter than 90th percentile of 1982-2011 temperatures on that day.")


# Assemble pieces
heatwave_heatmap
grid::grid.raster(lab_logo, x = 0.05, y = 0.03, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```




##  Temperature Anomaly Horizons

One way to think about the severity of these changes is to think about temperature horizons. A temperature horizon captures how long temperatures remain above certain thresholds. Each threshold is designated its own temperature, and in this way we can see how long within a year temperatures remained: 1 to 2 or as much as 4$^oC$ above normal.  



```{r horizons f, eval = T}
#| fig.alt = "Horizon plot of all years and their temperature anomaly horizons"

# Filter outliers beyond 25th and 75th quantile
cutpoints <- grid_data %>% 
  mutate(
    outlier = between(
      anom_f,
      quantile(anom_f, 0.25, na.rm = T)-
        1.5 * IQR(anom_f, na.rm = T),
      quantile(anom_f, 0.75, na.rm = T)+
        1.5 * IQR(anom_f, na.rm = T))) %>% 
  filter(outlier)

# Set origin
ori <- 0

# Manually pick cut points
sca <- seq(-5, 5, length.out = 9)

sca_bins <- str_c(sca[1:(length(sca)-1)], " to ", sca[2:length(sca)], "\u00b0F")
sca_labels <- rev(sca_bins)


# Function to plot one or more years
anom_horizon_plot <- function(grid_data, origin = ori, scale_cutpoints = sca, labels = sca_labels){
  
   ggplot(grid_data) +
  geom_horizon(aes(flat_date, 
                   sst_anom,
                   fill = ..Cutpoints..), 
               origin = ori, 
               horizonscale = sca) +
  scale_fill_hcl(palette = 'RdBu', reverse = T, labels = sca_labels) +
  facet_grid(year~.) +
  theme(
    #strip.background = element_rect(fill = "#36454F", color = "#36454F"),
    panel.grid = element_blank(),
    panel.spacing.y=unit(0.1, "lines"),
    strip.text.y = element_text(size = 7, angle = 0, hjust = 0),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.border = element_blank(),
    legend.position = "left") +
  scale_x_date(expand=c(0,0), 
               date_breaks = "1 month", 
               date_labels = "%b") +
  labs(x = '', 
       fill = "Temperature Anomaly") 
  
  
}


# Plot Horizon Plot Using All Years
gom_annual_horizons <- anom_horizon_plot(grid_data = filter(grid_data, yr <= 2021), 
                  origin = ori, 
                  scale_cutpoints = sca) +
  labs(title = "Temperature Anomaly Horizons",
       subtitle = "How deviation from the norm has become the new-normal",
       caption = "Climatatological Reference Period: 1982-2011")


# Plot it
gom_annual_horizons
grid::grid.raster(lab_logo, x = 0.05, y = 0.03, just = c('left', 'bottom'), width = unit(0.8, 'inches'))

```

## Comparing the Gulf's Two Hottest Years

If we pull the horizons of our two hottest years on record it makes it easier to contrast where either one experienced acute high temperature events, and where there were sustained periods of above average temperatures.

Early into 2021 it was apparent that the year was on-par with the previous title-holder for warmest year on record.


```{r hottest two years f, eval = T}
#| fig.alt = "Single year horizon plots comparing 2012 and 2022"

# cut out outliers and set cutpoints
cutpoints_2 <- grid_data %>% 
  filter(year %in% c("2012", "2021")) %>% 
  mutate(
    outlier = between(
      anom_f,
      quantile(anom_f, 0.25, na.rm=T)-
        1.5*IQR(anom_f, na.rm=T),
      quantile(anom_f, 0.75, na.rm=T)+
        1.5*IQR(anom_f, na.rm=T))) %>% 
  filter(outlier)



# 2012
hori_2012 <- grid_data %>% 
  filter(year == "2012") %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  facet_wrap(~year, strip.position = "top")
  
# 2021
hori_2021 <- grid_data %>% filter(year == "2021") %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  facet_wrap(~year, strip.position = "top") +
  theme(plot.caption = element_text(colour = "gray25", size = 6)) +
  labs(caption = "Climatatological Reference Period: 1982-2011")

# Put them together as one figure
comparison_horizons <- hori_2012 / hori_2021 + plot_layout(guides = "collect")

# Do a bunch of formatting
comparison_horizons <-comparison_horizons &
  theme(
    #Legend
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines")) &
   guides(fill = guide_legend(title = "Temperature Anomaly",
                              title.hjust = 0.5, 
                              title.position = "top",
                              label.position = "bottom", 
                              nrow = 1, 
                              byrow = T, reverse = T)) &
  plot_annotation(title = "Comparing the Gulf of Maine's Two Hottest Years", 
                  subtitle = "Duration of temperature anomaly horizons, colored by their strength")




# Plot it
comparison_horizons
grid::grid.raster(lab_logo, x = 0.08, y = 0.04, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
  
```


## Losing the Balance Between Hot and Cold

Another way to visualize the climate transition that we are observing is by looking at the fraction of each year spent in different temperature ranges. Under a steady climate we would expect over the long-term to spend similar amounts of the year experiencing relatively warm & cold temperatures. These periods would balance themselves out and we would on-average have experienced something similar to the long-term climate.

What we have been seeing in the Gulf of Maine recently has lost that sense of balance. Larger fractions of the year are shared by above average temperatures & cold spell events are becoming vanishingly rare.


```{r stream F, eval = T}
#| fig.alt = "Streamplot tracking fraction of each year spent in varying degrees of temperature anomalies"


# Set Factor Order for Bins
bin_order <- c(
  "Less than -2", 
  "-2 to -1",
  "Within 1",
  "1 to 2", 
  "Greater than 2")

# Set Colors
bin_colors <- RColorBrewer::brewer.pal(n = length(bin_order), 
                                       name = "RdBu")
bin_colors <- rev(bin_colors)

# Count Days in Bins
day_types <- region_hw %>% 
  mutate(
    `Less than -2`   = if_else(anom_f <= -2, TRUE, FALSE),
    `-2 to -1`     = if_else(between(anom_f, -2, -1), TRUE, FALSE),
    `Within 1`     = if_else(between(anom_f, -1, 1), TRUE, FALSE),
    `1 to 2`       = if_else(between(anom_f, 1, 2), TRUE, FALSE),
    `Greater than 2` = if_else(anom_f >= 2, TRUE, FALSE)) %>% 
  group_by(yr) %>% 
  summarise(
    `Less than -2`   = sum(`Less than -2`, na.rm = T),
     `-2 to -1`    = sum(`-2 to -1`, na.rm = T),
    `Within 1`     = sum(`Within 1`, na.rm = T),
    `1 to 2`       = sum(`1 to 2`, na.rm = T),
    `Greater than 2` = sum(`Greater than 2`, na.rm = T),
    .groups = "drop") %>% 
  pivot_longer(cols = "Less than -2":"Greater than 2", 
               names_to = "Anomaly Strength", 
               values_to = "day_total", 
               values_drop_na = FALSE) %>% 
  mutate(day_total = if_else(is.na(day_total), 0L, day_total),
         `Anomaly Strength` = factor(`Anomaly Strength`, levels = bin_order))






# Plot the Streamplot
hw_streamplot <- ggplot(day_types, aes(yr, y = day_total, fill = `Anomaly Strength`)) +
  geom_stream(type = "proportion") +
  scale_x_continuous(breaks = seq(1980,2030, by = 10), 
                     minor_breaks = seq(1975, 2030, by = 5), 
                     expand = expansion(add = c(0,0))) +
  scale_y_continuous(labels = scales::percent_format(), 
                     expand = expansion(add = c(0,0))) + 
  scale_fill_manual(values = setNames(bin_colors, bin_order)) + 
  labs(y = "Percent of Year", x = "", fill = "", 
       title = "A Warmer Gulf of Maine",
       subtitle = "An increasing fraction of each year experiencing warm temperature anomalies",
       caption = "Anomalies calculated using 1982-2011 climatology.") +
  theme(
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines"),
    plot.margin = unit(c(.5,1,.3,.5), "cm"),
    plot.title.position = "plot",
    legend.margin = margin(c(7,0,0,0)),
    legend.justification = "center") +
   guides(fill = guide_legend(title = "Anomaly Strength \u00b0F",
                             title.hjust = 0.5,
                             title.position = "left",
                             label.position = "bottom",
                             nrow = 1,
                             byrow = T))


# Show figure
hw_streamplot
grid::grid.raster(lab_logo, x = 0.08, y = 0.04, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```


## Characterizing the Hottest Years

There are a number of ways one might try to describe the extent and severity of temperature extremes. Aside from overall average temperatures there may be acute temperature surges that may have a larger impact than a prolonged warm period.




```{r year panels f, eval = T}
#| fig.alt = "4-panel plot comparing different features of hot years ranked by different metrics"

# Use all years so they can be ranked independently
year_mets <- grid_data %>% 
  filter(year != "2022") %>% 
  group_by(year) %>% 
  summarise(
    total_days                         = n(),
    `Average Temperature`              = round(mean(sst_f, na.rm = T), 2),
    `Temperature Above Average`        = round(mean(anom_f, na.rm = T), 2),
    `Cumulative Degrees Above Average` = round(sum(anom_f, na.rm = T), 0),
    hw_days                            = sum(mhw_event, na.rm = T),
    hw_events                          = n_distinct(mhw_event_no),
    peak_temp                          = max(sst_f),
    peak_anom                          = max(anom_f)) %>% 
  mutate(
    hw_day_pct = round((hw_days/total_days) * 100, 2),
    `Average Heatwave Length` = round(hw_days/hw_events,  1),
    yr_focus = ifelse(year == "2021", TRUE, FALSE)) %>% 
  pivot_longer(names_to = "Var", values_to = "Metric", 
               cols = c("Average Temperature", "Temperature Above Average", "Average Heatwave Length", "Cumulative Degrees Above Average"))




# Build Seperately:
rplot_1 <- year_mets %>% 
  filter(Var == "Average Temperature") %>% 
  group_by(Var) %>% 
  slice_max(n = 5, order_by = Metric) %>% 
  mutate(year = factor(year),
         year = fct_reorder(year, Metric, median)) %>% 
  ggplot(aes(Metric, year, fill = yr_focus, color = yr_focus)) +
    directlabels::geom_dl(aes(label = str_c(Metric, deg_sym)), 
                          method = list(directlabels::dl.trans(x = x + 1), rot = 0)) +
    geom_col(show.legend = FALSE) + 
    scale_x_continuous(expand = expansion(mult = c(0, 0.15)), labels = number_format(suffix = deg_f)) +
    scale_color_gmri() +
    scale_fill_gmri() +
    theme(panel.grid = element_blank()) +
   labs(y = "",
        x = "Average Temperatures"
        # caption = "Anomalies calculated using 1982-2011 climatology."
      )


rplot_2 <- year_mets %>% 
  filter(Var == "Temperature Above Average") %>% 
  group_by(Var) %>% 
  slice_max(n = 5, order_by = Metric) %>% 
  mutate(year = factor(year),
         year = fct_reorder(year, Metric, median)) %>% 
  ggplot(aes(Metric, year, fill = yr_focus, color = yr_focus)) +
    directlabels::geom_dl(aes(label = str_c(Metric, deg_sym)), 
                          method = list(directlabels::dl.trans(x = x + 1), rot = 0)) +
    geom_col(show.legend = FALSE) + 
    scale_x_continuous(expand = expansion(mult = c(0, 0.15)), labels = number_format(suffix = deg_f)) +
    scale_color_gmri() +
    scale_fill_gmri() +
    theme(panel.grid = element_blank()) +
   labs(y = "",
        x = "Temperature Above Average",#,
        #subtitle = "Temperatures Above Average",
        caption = "Anomalies calculated using 1982-2011 climatology.")


# Replace the facet option
hottest_yr_panels <- rplot_1 / rplot_2 + plot_annotation(title = "How 2021 Stacks Up:")

# Display plot
hottest_yr_panels
grid::grid.raster(lab_logo, x = 0.08, y = 0.025, just = c('left', 'bottom'), width = unit(0.8, 'inches'))

```


```{r year panels facets, eval = F}
# # Snipe out the top5 for each
# year_met_ranks <- year_mets %>% 
#   filter(Var %in% c("Average Temperature", "Temperature Above Average")) %>% 
#   group_by(Var) %>% 
#   slice_max(n = 5, order_by = Metric) %>% 
#   ungroup() %>%
#   mutate(year = reorder_within(year, Metric, Var)) 
# 
# # Plot the rankings
# hottest_yr_panels <- year_met_ranks %>% 
#   ggplot(aes(year, Metric, color = yr_focus)) +
#     geom_col(aes(fill = yr_focus), width = 0.8,  show.legend = F) +
#     directlabels::geom_dl(aes(label = str_c(Metric, deg_sym)), 
#                           method = list(directlabels::dl.trans(x = x + 1), rot = 0)) +
#     facet_wrap(~Var, scales = "free", ncol = 1) +
#     coord_flip() +
#     scale_x_reordered() +
#     scale_y_continuous(expand = expansion(mult = c(0, 0.15)), labels = number_format(suffix = deg_f)) +
#     scale_color_gmri() +
#     scale_fill_gmri() +
#     theme(panel.grid = element_blank()) +
#    labs(y = "",
#          x = NULL,
#          title = "How 2021 Stacks Up:",
#          caption = "Anomalies calculated using 1982-2011 climatology.",
#          subtitle = "Ranking Characteristics of Acute and Prolonged Warming Events")
# 
# 
# hottest_yr_panels
# grid::grid.raster(lab_logo, x = 0.08, y = 0.025, just = c('left', 'bottom'), width = unit(0.8, 'inches'))

```


## Mapping Record Temperatures

2021 was an exceptionally warm year for many parts of the World, not just the Gulf of Maine. By taking the average temperature for the year for all years, we can rank how this year compares to previous years in the OISST data. For many places around the world this was one of the top 3 warmest years on record.

```{r}
# Load the rankings done in python, do some checking
ranks_path <- str_c(oisst_path, 'temp_rankings/yrly_ranks_1982to2021.nc')

#  Pull this year
ranks_21 <- stack(ranks_path)[["X2021"]]

# Reclassify the Ranks from ascending cold-warm to warmest -> coldest
# Ranks are all integers so we just need to be sure that they are captured in the bounds we want
reclass_df <- c(
  # 0, 1.1, -1,  # Coldest Year
  # 1.1, 2.1, -2,  # Second Coldest
  # 2.1, 3.1, -3,  # Third Coldest
  # 3.1, 4.1, -4,  # Fourth Coldest
  # 4.1, 5.1, -5,  # Fifth Coldest
  # 3.1, 35, NA, # Things to Mask
  0, 35, NA, # Things to Mask
  35.1, 36, 5, # Fifth Warmest
  36.1, 37, 4, # Fourth Warmest
  37.1, 38, 3, # Third Warmest
  38.1, 39, 2, # Second Warmest
  39.1, 40, 1  # Warmest
  )

# make reclassification matrix
reclass_m <- matrix(reclass_df, ncol = 3, byrow = TRUE)

# reclassify - masking middle values handled with reclassification
ranks_reclass <- reclassify(ranks_21, reclass_m)

# Make Stars for ggplot
ranks_st <- st_as_stars(rotate(ranks_reclass))

# Mutate to get levels as factors for legend
ranks_st <- ranks_st %>% 
  mutate(
    temp_rank = case_when(
      # X2021 == -1 ~ "Coldest",
      # X2021 == -2 ~ "2nd Coldest",
      # X2021 == -3 ~ "3rd Coldest",
      # X2021 == -4 ~ "4th Coldest",
      # X2021 == -5 ~ "5th Coldest",
      X2021 == 5 ~ "5th Warmest",
      X2021 == 4 ~ "4th Warmest",
      X2021 == 3 ~ "3rd Warmest",
      X2021 == 2 ~ "2nd Warmest",
      X2021 == 1 ~ "Warmest",
      #TRUE ~ "Not a Record Year" # Use as center color for divergent scale
      TRUE ~ "NaN"),
    temp_rank = fct_drop(temp_rank),
    temp_rank = factor(temp_rank, levels = c(
      "Warmest", "2nd Warmest", "3rd Warmest", "4th Warmest", "5th Warmest",
      "Not a Record Year",
      "5th Coldest", "4th Coldest", "3rd Coldest", "2nd Coldest", "Coldest"))) %>% 
  select(-X2021) 


# Make map
ranks_map <- ggplot() +
  geom_stars(data = ranks_st) +
  geom_sf(data = world_sf, size = .25) +
  # scale_fill_brewer(palette = "RdBu", na.value = "transparent") +
  scale_fill_brewer(palette = "Reds", na.value = "transparent", direction = -1, na.translate = FALSE) +
  scale_x_continuous(breaks = seq(-180, 180, 90), 
                     expand = expansion(mult = c(0,0))) +
  scale_y_continuous(breaks = seq(-90, 90, 30), 
                     expand = expansion(mult = c(0,0))) +
  labs(title = "Record Setting Temperatures of 2021",
       fill = "2021 Ranking",
       caption = "Temperature rankings cover period of 1982-2021") +
  map_theme +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) 
  
ranks_map
grid::grid.raster(lab_logo, x = 0.08, y = 0.16, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```


## Record Temperatures for New England


If we look more closely at the Northeast US we can see that 2021 was an exceptional year for the area, with record temperatures covering a large portion of the area:


```{r, eval = TRUE}
# Zoom in on Northeast US
(northeast_records <- ggplot() +
  geom_stars(data = ranks_st) +
  geom_sf(data = new_england, size = .25) +
  geom_sf(data = canada, size = .25) +
  # scale_fill_brewer(palette = "RdBu", na.value = "transparent") + 
  scale_fill_brewer(palette = "Reds", na.value = "transparent", direction = -1, na.translate = FALSE) + 
  coord_sf(xlim = c(-80, -55), ylim = c(34, 49)) +
  labs(x = "", y = "", fill = "2021 Ranking",
       title = "Record Warm Temperatures",
       subtitle = "From North Carolina to Newfoundland",
       caption = "Temperature rankings cover period of 1982-2021")  +
  map_theme)

grid::grid.raster(lab_logo, x = 0.08, y = 0.02, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```

## New England Anomalies

The spatial variation in anomalies during this time was striking. The largest deviations in temperatures (unexpectedly hot temperatures) were concentrated off of Georges Bank:

```{r}

# Set time limits
annual_dates <- as.Date(c("2021-01-01", "2021-12-31"))

# Make data window for season
data_window <- data.frame(lon = c(-81, -54.5),
                          lat = c(33.5, 50),
                          time = annual_dates)

# Load data
annual_anoms <- oisst_window_load(data_window = data_window, anomalies = T, mac_os = "mojave")
annual_anoms <- stack(annual_anoms)
annual_mean <- mean(annual_anoms, na.rm = T)


# Change to F
annual_mean_f <- calc(annual_mean, function(x) as_fahrenheit(x, data_type = "anomalies"))


# Color limit for palettes
temp_limits <- c(-6, 6)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0F")


# Map it

# Make it a stars class to plot with ggplot
anoms_st <- st_as_stars(annual_mean_f)

#### 1. Map the Anomalies in Space
annual_gom_map <- ggplot() +
  geom_stars(data = anoms_st) +
  geom_sf(data = new_england, fill = "gray90", size = .25) +
  geom_sf(data = canada, fill = "gray90", size = .25) +
  geom_sf(data = greenland, fill = "gray90", size = .25) +
  geom_sf(data = region_extent, 
          color = "black", 
          linetype = 2, size = 1,
          fill = "transparent") +
  scale_y_continuous(breaks = seq(30, 50, by = 2)) +
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       labels = temp_labels,
                       oob = scales::oob_squish) +
  coord_sf(xlim = c(-78, -56.5), ylim = c(37.8, 48)) +
  labs(title = "2021: Average Temperature Anomalies",
       caption = "Anomalies calculated using 1982-2011 climatology.") +
  map_theme +
  theme(legend.position = "bottom",
        legend.title = element_text(angle = 0))+
  guides("fill" = guide_colorbar(
    title = "Average Temperature Anomaly ",
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(3, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 

# Show figure
annual_gom_map


```


## Global Temperature Anomalies


```{r}

# Set time limits
annual_dates <- as.Date(c("2021-01-01", "2021-12-31"))

# Make data window for season
global_data_window <- data.frame(lon = c(-360, 180),
                          lat = c(-90, 90),
                          time = annual_dates)

# Load data
global_annual_anoms <- oisst_window_load(data_window = global_data_window, anomalies = T, mac_os = "mojave")
global_annual_anoms <- stack(global_annual_anoms)
global_annual_anoms <- setExtent(global_annual_anoms, extent(c(0, 360, -90, 90)))
global_annual_mean <- mean(global_annual_anoms, na.rm = T)
global_annual_mean <- rotate(global_annual_mean)

# Change to F
global_annual_mean_f <- calc(global_annual_mean, function(x) as_fahrenheit(x, data_type = "anomalies"))


# Color limit for palettes
temp_limits <- c(-3, 3)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0F")


# Map it

# Make it a stars class to plot with ggplot
global_anoms_st <- st_as_stars(global_annual_mean_f)

#### 1. Map the Anomalies in Space
annual_anom_map <- ggplot() +
  geom_stars(data = global_anoms_st) +
  geom_sf(data = world_sf, fill = "gray90", size = .25) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       breaks = temp_breaks,
                       labels = temp_labels,
                       oob = scales::oob_squish) +
  labs(title = "2021: Average Temperature Anomalies",
       caption = "Anomalies calculated using 1982-2011 climatology.") +
  coord_sf(expand = FALSE) +
  map_theme +
  guides("fill" = guide_colorbar(
    title = "Average Temperature Anomaly ",
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(3, "in"),
    frame.colour = "black",
    ticks.colour = "black")) +
  theme(legend.position = "bottom",
        legend.title = element_text(angle = 0),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) 

# Show figure
annual_anom_map


```

## Worldwide Temperature Changes


If we look at the rates of change for each grid cell on Earths surface, rather than a plot of SST over time, it is possible to rank how each location on Earth is warming relative to others. By ranking those warming rates, and then taking the average ranking across the Gulf of Maine, we can obtain the average warming rank for our area compared to the rest of the globe.

Prior GMRI research found that the Gulf of Maine had been warming at a faster rate than 99% of the world's oceans. This is still true when comparing warming rates across certain timescales, such as from 2004 to 2013 (when our initial warming study took place).

Over a longer reference period, from 1982 to 2021, the warming rate of the Gulf of Maine, as a whole is still among the highest in the world  increasing at a rate of almost 0.05 C per year, and faster than 96.2% of the world's oceans.

However there can be significant spatial variability in this ranked warming rate across the Gulf of Maine. Warming rates across the region range from 0.014 C per year to as high as 0.097 C per year, meaning some locations over this longer time frame are still warming faster than 99.5% of the worlds oceans.

This variability reflects the various physical oceanographic drivers influencing SST in the region, and in particular the [significant interannual variability in the Gulf Streams influence on Georges Bank](https://eos.org/features/why-is-the-gulf-of-maine-warming-faster-than-99-of-the-ocean).


```{r load warming rates, eval = T}

# # 1. Warming Rates and Rankings
rates_path <- str_c(oisst_path, "warming_rates/annual_warming_rates1982to2021.nc")
rates_nc <- stack(rates_path, varname = "annual_warming_rate")
ranks_nc <- stack(rates_path, varname = "rate_percentile")



# Get the rank information that go with each area
ranks_masked <- mask_nc(ranks_nc, region_extent)
rates_masked <- mask_nc(rates_nc, region_extent)
region_ranks <- get_masked_vals(masked_ranks = ranks_masked, 
                                masked_rates = rates_masked, 
                                in_fahrenheit = T)


# Prep it for text input.
avg_rank <- region_ranks$`Mean Rank` * 100
avg_rate <- region_ranks$`Mean Rate`
low_rank <- region_ranks$`Min Rank` * 100
low_rate <- region_ranks$`Min Rate`
top_rank <- region_ranks$`Max Rank` * 100
top_rank <- ifelse(top_rank == 100, "greater than or equal to 99.5", top_rank)
top_rate <- region_ranks$`Max Rate`

```




```{r warming rates Fahrenheit, layout="l-body-outset", fig.width=6, fig.height=4}

# Change to F
world_rate_stars <- calc(rotate(rates_nc), function(x) as_fahrenheit(x, data_type = "anomalies")) %>% 
  st_as_stars()


# Color limit for palettes
temp_limits <- c(-.1, .1)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0F")

                     
# Northeast Shelf
global_rates_zoom <- ggplot() +
  geom_stars(data = world_rate_stars) +
  geom_sf(data = new_england, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = canada, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = greenland, color = "white", fill = "gray30", size = .05) +
  scale_y_continuous(breaks = seq(-90, 90, by = 5)) +
  scale_x_continuous(breaks = seq(-180, 180, by = 10))+
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       breaks = temp_breaks, 
                       labels = temp_labels,
                       oob = oob_squish) +
  map_theme +
  coord_sf(
    xlim = c("xmin" = -77, "xmax" = -50),
    ylim = c("ymin" = 32, "ymax" = 48),
    expand = T) +
  guides("fill" = guide_colorbar(
    title = "Annual Warming Rate: 1982-2021",
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(2.5, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# Global View
global_rates <- ggplot() +
  geom_stars(data = world_rate_stars) +
  geom_sf(data = world_sf, color = "white", fill = "gray30", size = .1) +
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       breaks = temp_breaks, 
                       labels = temp_labels,
                       oob = oob_squish) +
  map_theme +
  coord_sf(expand = F) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  guides("fill" = guide_colorbar(title = "Annual Warming Rate: 1982-2021",
                                 title.position = "top", 
                                 title.hjust = 0.5,
                                 barwidth = unit(2.5, "in"),
                                 frame.colour = "black",
                                 ticks.colour = "black"))  +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


# plot map
rates_global_figure <- (global_rates_zoom | global_rates ) + 
  plot_annotation(title = "Global SST Change", 
                  caption = "Data Source: OISSTv2") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")


# plot patchwork
rates_global_figure
```


### Warming Rate Rankings


Based on data from 1982-2021, the warming rates of Gulf of Maine have been some of the highest in the world. The area as a whole has been increasing at a rate of `r avg_rate`$^{\circ}F/year$ which is faster than `r avg_rank`% of the world's oceans.

Over that same period locations within the Gulf of Maine have been warming at rates as low as `r low_rate`$^{\circ}F/year$ and as rapidly as `r top_rate`$^{\circ}F/year$, corresponding to ranks as low as `r low_rank`% and as high as `r top_rank`%.


```{r highlighting gom warming rate, eval = TRUE, layout="l-body-outset", fig.width=6, fig.height=4}
# for an expanded view need to use the global ranks
world_rank_stars <- st_as_stars(rotate(ranks_nc))

# Reclassify raster into 10% Bins
world_ranks_binned <- world_rank_stars %>% 
  mutate(
    rank = Annual.Sea.Surface.Temperature.Warming.Rate.Rank,
    rank_bin = cut(rank, breaks = seq(0, 1, by = 0.1))) %>% 
  select(-c(rank, Annual.Sea.Surface.Temperature.Warming.Rate.Rank))

# # Define the number of colors you want
# nb.cols <- 10
# mycolors <- RColorBrewer::brewer.pal(n = 10, "Reds")


# Northeast Shelf
global_ranks_zoom <- ggplot() +
  geom_stars(data = world_ranks_binned) +
  geom_sf(data = new_england, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = canada, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = greenland, color = "white", fill = "gray30", size = .05) +
  scale_y_continuous(breaks = seq(-90, 90, by = 5)) +
  scale_x_continuous(breaks = seq(-180, 180, by = 10))+
  scale_fill_brewer(palette = "Spectral", direction = -1, na.value = "transparent") +
  # scale_fill_viridis_d(option = "viridis", direction = 1, na.value = "transparent", na.translate = FALSE) +
  map_theme +
  coord_sf(
    xlim = c("xmin" = -77, "xmax" = -50),
    ylim = c("ymin" = 32, "ymax" = 48),
    expand = T) +
  guides("fill" = guide_colorsteps(
    title = "Warming Rate Percentile: 1982-2021",
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(2.5, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# Global View
global_ranks <- ggplot() +
  geom_stars(data = world_ranks_binned) +
  geom_sf(data = world_sf, color = "white", fill = "gray30", size = .1) +
  scale_fill_brewer(palette = "Spectral", direction = -1, na.value = "transparent") +
  # scale_fill_viridis_d(option = "viridis", direction = 1, na.value = "transparent", na.translate = FALSE) +
  map_theme +
  coord_sf(expand = F) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  guides("fill" = guide_colorsteps(title = "Warming Rate Percentile: 1982-2021",
                                 title.position = "top", 
                                 title.hjust = 0.5,
                                 barwidth = unit(2.5, "in"),
                                 frame.colour = "black",
                                 ticks.colour = "black"))  +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


# plot map
ranks_global_figure <- (global_ranks_zoom | global_ranks ) + 
  plot_annotation(title = "Global SST Change Rankings", 
                  caption = "Data Source: OISSTv2") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")


# plot patchwork
ranks_global_figure

```



## Fastest Warming Regions

To get a better look at extremes we can mask values below the 80th percentile and look more closely at the differences among the fastest warming areas on Earth. Below only the fastest 20% of locations are shown, to  show more detail among the fastest warming areas.


```{r}

# Mask out where its under .8
ranks_top20 <- ranks_nc
ranks_top20[ranks_top20 < 0.8] <- NA


# Make stars, bin by .25
ranks_top_stars <- st_as_stars(rotate(ranks_top20))

ranks_top_binned <- ranks_top_stars %>% 
  mutate(
    rank = Annual.Sea.Surface.Temperature.Warming.Rate.Rank,
    rank_bin = cut(rank, breaks = seq(0.8, 1, by = 0.025))) %>% 
  select(-c(rank, Annual.Sea.Surface.Temperature.Warming.Rate.Rank))

# Remake maps


# Northeast Shelf
top_ranks_zoom <- ggplot() +
  geom_stars(data = ranks_top_binned) +
  geom_sf(data = new_england, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = canada, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = greenland, color = "white", fill = "gray30", size = .05) +
  scale_y_continuous(breaks = seq(-90, 90, by = 5)) +
  scale_x_continuous(breaks = seq(-180, 180, by = 10))+
  scale_fill_viridis_d(option = "plasma", direction = 1, na.value = "gray10") +
  map_theme +
  coord_sf(
    xlim = crop_x_expanded,
    ylim = crop_y_expanded,
    # xlim = c("xmin" = -75, "xmax" = -60),
    # ylim = c("ymin" = 38, "ymax" = 48),
    expand = T) +
  guides("fill" = guide_colorsteps(
    title = "Warming Rate Percentile: 1982-2021",
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(2.5, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# Global View
top_ranks <- ggplot() +
  geom_stars(data = ranks_top_binned) +
  geom_sf(data = world_sf, color = "white", fill = "gray30", size = .1) +
   scale_fill_viridis_d(option = "plasma", direction = 1, na.value = "gray10") +
  map_theme +
  coord_sf(expand = F) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  guides("fill" = guide_colorsteps(title = "Warming Rate Percentile: 1982-2021",
                                 title.position = "top", 
                                 title.hjust = 0.5,
                                 barwidth = unit(2.5, "in"),
                                 frame.colour = "black",
                                 ticks.colour = "black"))  +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


# plot map
top_ranks_figure <- (top_ranks_zoom | top_ranks ) + 
  plot_annotation(title = "Earth's Fastest Warming Ocean Regions", 
                  caption = "Data Source: OISSTv2, rankings below 80th percentile masked.") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")


# plot patchwork
top_ranks_figure


```





## Comparing to Large Marine Ecosystems

We are often asked when speaking to the degree of the Gulf of Maine is warming: 
> "Where is it warming faster?"

This question is challenging because the Gulf of Maine, and any area in the ocean has a unique size/shape/dynamic that makes them difficult to directly compare directly. To put the Gulf of Maine's warming into perspective against similar ecologically and oceanographically relevant units we can compare it against the [large marine ecosystems (LME)](https://www.lmehub.net/) of the world.
 
```{r lme timeseries}
# Load the LME timeseries to calculate rates of warming 
lme_names <- get_region_names("lme") # names
lme_paths <- get_timeseries_paths("lme", mac_os = "mojave") # paths

# Data
lme_oisst <- map(lme_names, ~ read_csv(lme_paths[[.x]][["timeseries_path"]], col_types = cols(), guess_max = 1e6)) 
lme_oisst <- setNames(lme_oisst, lme_names)

# Add Gulf of Maine Timeseries to the List with LME's
lme_oisst[["gulf_of_maine"]] <- region_timeseries

# Make sure dates are formatted
lme_oisst <- map(lme_oisst, ~ mutate(.x, time = as.Date(time)))


####  Warming Rates  ####
lme_8221 <- map(lme_oisst, ~ mutate(.x, year = year(time)) %>% filter(year %in% c(1982:2021)))

# Get warming rates
lme_rates <- map_dfr(lme_8221, function(lme_data){
  # Uses area weighted SST
  lme_yearly <- group_by(lme_data, year) %>% 
    summarise(sst = mean(area_wtd_sst, na.rm = T))
  yrly_warming <- lm(sst ~ year, data = lme_yearly)
  mod_details <- broom::tidy(yrly_warming)
  yrly_rate <- mod_details[2,"estimate"]
  names(yrly_rate) <- "annual_warming_rate_C"
  yrly_rate
}, .id = "Region") %>% 
  arrange(desc(annual_warming_rate_C))


# Pull out the top x
rates_ordered <- lme_rates %>% 
  mutate(`Warming Rank` = row_number(),
         `Warming Rate F` = as_fahrenheit(annual_warming_rate_C, data_type = "anomalies"),
         Region = str_replace_all(Region, "_", " "),
         Region = str_to_title(Region)) %>% 
  select(Region, `Warming Rank`, `Warming Rate C` = annual_warming_rate_C, `Warming Rate F`) 


# Make table of top 10*
lme_rank_table <- rates_ordered %>% 
  slice(1:10) %>% 
  mutate(across(where(is.numeric), round, 3)) %>% 
  gt(rowname_col = "Region") %>% 
   tab_header(
    title = md("**Warming Rates of Large Marine Ecosystems**")) %>%
  tab_stubhead(label = "Region") %>% 
  tab_style(
    style = list(
      cell_fill(color = "gray90"),
      cell_text(style = "oblique")),
    locations = cells_body(
      rows = Region == "Gulf Of Maine")) %>% 
  tab_spanner(label = md("*Annual Temperature Change*"),
              columns = c(`Warming Rate C`, `Warming Rate F`)) %>% 
 
  cols_label(
      `Warming Rate F` = "\u00b0F",
      `Warming Rate C` = "\u00b0C") %>% 
  tab_source_note(source_note = md("**Data Source:** *NOAA OISSTv2 Daily Sea Surface Temperature Data.*")) %>%                    
  tab_source_note(source_note = md("**Notes:** *Temperatures adjusted for latitudinal distotion.*") ) %>%  
  tab_footnote(
    footnote = "Gulf of Maine not a true large marine ecosystem",
    locations = cells_stub(rows = which(rates_ordered$Region == "Gulf Of Maine")) 
  )


# show table
lme_rank_table
```
 
 
 
 
```{r mapping lme rates, eval = FALSE}

# Netcdf of warming rates
rates_path <- str_c(oisst_path, "warming_rates/annual_warming_rates1982to2021.nc")
rates_nc <- stack(rates_path, varname = "annual_warming_rate")
ranks_nc <- stack(rates_path, varname = "rate_percentile")



# Get paths to the LME's and their timeseries
lme_shapes <- map(lme_names, ~  read_sf(lme_paths[[.x]][["shape_path"]]))  %>% 
  setNames(lme_names)

# Include Gulf of Maine as a shape
lme_shapes[["gulf_of_maine"]] <- region_extent



# Mask with each to get warming %
lme_rates_masked <- map(lme_shapes, ~ mask_nc(ras_obj = rates_nc, mask_shape = .x))
lme_perc_masked <- map(lme_shapes, ~ mask_nc(ras_obj = ranks_nc, mask_shape = .x))


# Put the percentiles and average rates in one table
if( all(names(lme_rates_masked) == names(lme_perc_masked)) ){
  percentile_table <- map2_dfr(lme_rates_masked, lme_perc_masked, function(rate_lme, perc_lme){
    rate_val <- cellStats(rate_lme, mean, na.rm = T)
    perc_val <- cellStats(perc_lme, mean, na.rm = T)
    df_out <- data.frame(
      "annual_rate" = rate_val,
      "warming_percentile" = perc_val)
    
    return(df_out)
    
  }, .id = "ecosystem")
}


# # Check it out:
# percentile_table %>% arrange(desc(warming_percentile)) %>% head()
# percentile_table %>% arrange(desc(annual_rate)) %>% head()


# Bizarre...

# Put LME's in one collection
lme_shape_tab <- bind_rows(lme_shapes)

# Join with the  rates/ranks
perc_tab <- percentile_table %>% 
  mutate(name = str_replace_all(ecosystem, "_", " "),
         name = str_to_title(name),
         name = str_replace(name, "Of", "of"))

# Add the percentile information, and map them
lme_sf_ranks <- left_join(lme_shape_tab, perc_tab, by = c("name")) 

ggplot() +
  geom_sf(data = lme_sf_ranks, aes(fill = warming_percentile), color = "transparent") +
  geom_sf(data = world_sf, size = 0.25) +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "transparent") +
  theme(legend.position = "bottom") +
  coord_sf(expand = FALSE) +
  map_theme +
  guides(fill = guide_colorbar(
    title.position = "top",
    title.hjust = 0.5, 
    barwidth = unit(4, "in"), 
    frame.colour = "black")) +
  labs(fill = "Warming Rate Percentile",
       title = "")


```
 




## A Note on Data Sources:

NOAA_ERSST_V5 data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/data/gridded/data.noaa.ersst.v5.html.

NOAA High Resolution SST data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html. 



```{r, results='hide'}


####  Saving Figures  ####


# ERSST timeline
ggsave(plot = ersst_long_term_p, filename = str_c(save_location, "ersst_long_term_temps.png"),
       dpi = "retina", bg = "white")


# How ERSSt and OISST are different
ggsave(plot = ersst_v_oisst, filename = str_c(save_location, "ersst_and_oisst_temps.png"),
       dpi = "retina", bg = "white")

# How 2021 compares to long-term pattern
ggsave(plot = annual_temps_plot, filename = str_c(save_location, "annual_temps_plot.png"),
       dpi = "retina", bg = "white")


# simplified temperature plot
ggsave(plot = temp_simplified, filename = str_c(save_location, "oisst_simple_trends.png"),
       dpi = "retina", bg = "white")

# accelerating warming plot
ggsave(plot = gom_recent_warming, filename = str_c(save_location, "oisst_recent_gomwarming.png"),
       dpi = "retina", bg = "white")


# 2021 anomaly bar chart
ggsave(plot = gom_anomaly_bars, filename = str_c(save_location, "oisst_gom_anom_bars.png"),
       dpi = "retina", bg = "white")


# 2021 Record daily temperatures
ggsave(plot = hottest_days_plot,
       filename = str_c(save_location, "2021_hottest_days.png"),
       dpi = "retina", bg = "white")



# Full Year Temperatures
ggsave(plot = hw_temp_p, filename = str_c(save_location, "oisst_lastyr_temps.png"),
       dpi = "retina", bg = "white")



# Full Year Anomalies
ggsave(plot = hw_anom_p, filename = str_c(save_location, "oisst_lastyr_anoms.png"),
       dpi = "retina", bg = "white")



# Full year timeline of temps and anomalies
ggsave(plot = year_timeline, filename = str_c(save_location, "oisst_lastyr_timeline.png"),
       dpi = "retina", bg = "white")


# Heatwave status heatmap
ggsave(plot = heatwave_heatmap, filename = str_c(save_location, "oisst_gom_heatmap.png"),
       dpi = "retina", bg = "white")


# All year temperature horizons
ggsave(plot = gom_annual_horizons, filename = str_c(save_location, "oisst_gom_horizons.png"),
       dpi = "retina", bg = "white")


# 2021 | 2012 horizons
ggsave(plot = comparison_horizons, filename = str_c(save_location, "gom_hottest_horizons.png"),
       dpi = "retina", bg = "white")


# streamplot
ggsave(plot = hw_streamplot, filename = str_c(save_location, "gom_anomaly_streamplot.png"),
       dpi = "retina", bg = "white")


# record years panels
ggsave(plot = hottest_yr_panels, 
       filename = str_c(save_location, "gom_hottest_year_panels.png"),
       dpi = "retina", bg = "white")

# Annual Temperature Anomalies Map(s)
ggsave(plot = annual_gom_map, 
       filename = str_c(save_location, "gom_annual_anom_map.png"),
       dpi = "retina", bg = "white")

# Global version
ggsave(plot = annual_anom_map, 
       filename = str_c(save_location, "global_annual_anom_map.png"),
       dpi = "retina", bg = "white")

# Ranks maps
ggsave(plot = ranks_map, filename = str_c(save_location, "record_years_map.png"),
       dpi = "retina", bg = "white")


# northeast warmest years
ggsave(plot = northeast_records, filename = str_c(save_location, "record_years_ne_map.png"),
       dpi = "retina", bg = "white")


# Global warming rates
ggsave(plot = rates_global_figure, filename = str_c(save_location, "global_rates_map.png"),
       dpi = "retina", bg = "white")


# global warming rankings
ggsave(plot = ranks_global_figure, filename = str_c(save_location, "global_ranks_map.png"),
       dpi = "retina", bg = "white")

# top 20% rankings
ggsave(plot = top_ranks_figure, filename = str_c(save_location, "ranks_top20.png"),
       dpi = "retina", bg = "white")



# LME Report Table
if(TRUE){
  gtsave(data = lme_rank_table, 
         filename = str_c(save_location, "lme_warming_table.png"))
}


  
```



`r insert_gmri_footer()`