---
title: "Gulf of Maine Seasonal Sea Surface Temperature Update"
author: "Adam Kemberling"
url: "https://github.com/adamkemberling"
affiliation: "Gulf of Maine Research Institute"
affiliation_url: "https://www.gmri.org"
description: | 
  Seasonal Patterns in Sea Surface Temperatures
date: "Updated on: `r Sys.Date()`"
format: 
  html:
    toc: true
    self-contained: true
execute: 
  echo: false
  warning: false
  message: false
  comment: ""
  fig.align: center
params: 
  year: 2025
  season: "Summer"
  seas: "summer"
  save_data: FALSE
  save_figures: FALSE
---

```{r}
#| label: packages
#| include: false
#| echo: false
#| warning: false
 
####  Packages  ####
{
  library(lubridate)
  library(raster)
  library(here)
  library(rnaturalearth)
  library(scales)
  library(sf)
  library(stars)
  library(gmRi)
  library(heatwaveR)
  library(gt)
  library(gtExtras)
  library(patchwork)
  library(tidyverse)
  library(ggforce)
  library(geomtextpath)
  library(showtext)
  library(gtable)
  library(ggpubr)
}
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")
conflicted::conflict_prefer("rotate", "raster")

# Support Functions
source(here("R/oisst_support_funs.R"), verbose = FALSE)
source(here("R/temp_report_support.R"), verbose = FALSE)

# Box paths
oisst_path <- cs_path("res", "OISST/oisst_mainstays")

# Polygons for mapping
new_england <- ne_states("united states of america", returnclass = "sf")
canada      <- ne_states("canada", returnclass = "sf")
world_sf    <- ne_countries(returnclass = "sf")
greenland   <- ne_states(country = "greenland", returnclass = "sf")


# File paths for various extents based on "apershing_gulf_of_maine"
region_paths <- get_timeseries_paths(
  region_group = "gmri_sst_focal_areas", 
  box_location = "cloudstorage")

# Polygon Path
poly_path <- region_paths[["apershing_gulf_of_maine"]][["shape_path"]]

# Load the bounding box for Andy's GOM to show the area we're reporting on
region_extent <- st_read(poly_path, quiet = TRUE)

# Load bathymetry for the bottom contours:
# Bathymetric Contour Data
elev_terra <- terra::rast(str_c(cs_path("res","Bathy/"), "ETOPO1/NEShelf_Etopo1_bathy.tiff")) 


```





```{r}
#| label: param-setup
#| echo: false

#####  Parameter option switches:  ####
season_deets <- season_text_details(season_op = params$season, year_op = params$year)

# Set color for the season based on the parameter:
season_col <- season_deets$season_color

# Configure edges/centers of each season for label positioning
season_cent  <- season_deets$center_date
season_left  <- season_deets$left_lim
season_right <- season_deets$right_lim
season_start <- season_deets$season_start_doy
season_end   <- season_deets$season_end_doy

# Make names for the months to pull based on the season and current year
season_months <- season_deets$month_range

# Names of start and end month for text:
season_range <- season_deets$range_label

# Months/Dates covered by each season for use as labels
month_range <- switch(params$season,
  "Winter" = str_c("(December 1, ", params$year-1," - February 28, ", params$year,")"),
  "Spring" = str_c("(March 1, ", params$year," - May 31, ", params$year,")"),
  "Summer" = str_c("(June 1, ", params$year," - August 31, ", params$year,")"),
  "Fall"   = str_c("(September 1, ", params$year," - November 30, ", params$year,")"))
```





```{r}
#| label: style-sheet
#| results: asis

# Use GMRI style
use_gmri_style_rmd()

```



```{r}
#| label: fonts-config
#| echo: false

# Path to the directory containing the font file (replace with your actual path)
font_dir <- paste0(system.file("stylesheets", package = "gmRi"), "/GMRI_fonts/Avenir/")

# Register the font
font_add(
  family = "Avenir",
  file.path(font_dir, "LTe50342.ttf"),
  bold = file.path(font_dir, "LTe50340.ttf"),
  italic = file.path(font_dir, "LTe50343.ttf"),
  bolditalic = file.path(font_dir, "LTe50347.ttf"))

# Load the font
showtext::showtext_auto()

```


```{r}
#| label: regional-timeseries-prep


# Get the timeseries path
timeseries_path <- region_paths[["apershing_gulf_of_maine"]][["timeseries_path"]]

# Load timeseries of SST for Region
region_timeseries <- read_csv(
  timeseries_path, 
  col_types = cols(), 
  guess_max = 1e5) 

# Format timeseries for seasonal group estimates
region_timeseries <- region_timeseries %>% 
  mutate(
    time = as.Date(time),
    area_wtd_f = as_fahrenheit(area_wtd_sst),
    anom_f     = as_fahrenheit(area_wtd_anom, "anomalies")) %>% 
  distinct(time, .keep_all = T) %>% 
  supplement_season_info() %>% 
  filter(year %in% c(1982:params$year))



####  Get heatwave statuses for each day:

# Uses area weighted sst by default
region_hw <- pull_heatwave_events(
  temperature_timeseries = region_timeseries,
  threshold = 90, 
  clim_ref_period = c("1991-01-01", "2020-12-31")) %>% 
  supplement_hw_data() %>% 
  filter(
    #doy != 366, # remove leap years
    time <= as.Date(str_c(params$year, season_right)) # remove dates past the seasonal update
  ) 


# Heatmap prep
base_date <- as.Date("2000-01-01")
grid_dat <- region_hw %>%
# grid_dat <- sst_new %>%
  mutate(
    year = year(time),
    yday = yday(time),
    flat_date = as.Date(yday-1, origin = base_date),
    yr_rev = factor(year),
    yr_rev = fct_rev(yr_rev))


####  Prepare Data for This Seasonal Summary

# Subset data for the season were looking at
this_season_data <- region_hw %>% 
  dplyr::filter(
    season_eng == params$season,
    # -1 Works for winter, otherwise works fine
    if(params$season == "Winter") season_yr == params$year-1 else season_yr == params$year)
            

# Set seasonal end date so weeks don't bleed over
first_day <- as.Date(min(this_season_data$time))
last_day  <- as.Date(max(this_season_data$time))

# Set week start in the data so we can merge labels in accordingly later
this_season_data <-  this_season_data %>% 
  mutate(
    wk_start = floor_date(time, unit = "week"),
    wk_start = if_else(wk_start < first_day, first_day, wk_start),
    wk_start = as.Date(wk_start))
```




```{r}
#| eval: false
# Export sst for Noa, Damian's intern

# Just sending the sst for someone
region_timeseries %>% #names()
  select(
    time, 
    sst_c = area_wtd_sst, 
    anom_c = area_wtd_anom,
    sst_f = area_wtd_f, 
    anom_f = anom_f) %>%
  # write_csv(here::here("local_data/random_requests/GMRI_OISSTv2_GulfOfMaineSST.csv"))
  ggplot(aes(time, sst_f)) +
  #ggplot(aes(time, anom_f)) +
  geom_line()

```

```{r}
#| label: seasonal-sst-summary

# 1. Get the overall all year averages

# Set the start year, use 1981 if season is winter
ssn_year_start <- ifelse(params$season == "Winter", 1981, 1982)

# Get the average temperature for all years as a baseline
all_year_season_avg <- region_hw %>% 
  filter(season_yr %in% c(ssn_year_start:params$year),
         season_eng == params$season) %>% 
  group_by(season_eng) %>% 
  summarise(avg_temp_c = mean(sst),
            avg_anom_c = mean(sst_anom),
            avg_clim_c = mean(seas),
            avg_temp_f = mean(sst_f),
            avg_clim_f = mean(seas_f),
            avg_anom_f = mean(anom_f),
            .groups = "drop") %>% 
  mutate(temp_label = str_c(round(avg_temp_c, 2), " \u00b0C  |  ", round(avg_temp_f, 2), " \u00b0F"))


# 2. Get Averages for each year individually, 
seasonal_avgs <- region_hw %>% 
  filter(
    season_eng == params$season, 
    season_yr %in% c(ssn_year_start:params$year)) %>% 
  group_by(season_yr, season) %>% 
  summarise(
    # Uses Area Weighted SST
    mean_temp = mean(sst),
    mean_clim = mean(seas),
    mean_clim = median(mean_clim), # Makes sure any odd years (leap year or incomplete) are not different
    mean_anom = mean(sst_anom),
    temp_f    = as_fahrenheit(mean_temp),
    clim_f    = as_fahrenheit(mean_clim),
    anom_f    = as_fahrenheit(mean_anom, data_type = "anomalies"),
    .groups = "drop") %>% 
  mutate(
     year = season_yr,
     yr = season_yr,
     season_center = as.Date(str_c(yr, season_cent)),
     season_llim   = as.Date(str_c(yr, season_left)),
     season_rlim   = as.Date(str_c(yr, season_right)),
     temp_c_rnd    = sprintf(signif(mean_temp, 3), fmt = '%#.2f'),
     temp_f_rnd    = sprintf(signif(temp_f, 3), fmt = '%#.2f'),
     anom_c_rnd    = sprintf(signif(mean_anom, 3), fmt = '%#.2f'),
     anom_f_rnd    = sprintf(signif(anom_f, 3), fmt = '%#.2f'),
     temp_label    = str_c(temp_f_rnd, " \u00b0F  |  ", temp_c_rnd, " \u00b0C"),
     anom_label    = str_c(anom_f_rnd, " \u00b0F  |  ", anom_c_rnd, " \u00b0C")
  )
  
# Ranking the hottest years - on seasonal temperature
ranked_hotcold <- seasonal_avgs %>% 
  mutate(
    yr = factor(yr),
    year_rank = fct_reorder(
      .f = yr, 
      .x = mean_temp, 
      .fun = median, 
      .desc = TRUE)) %>% 
  arrange(year_rank) %>% 
  mutate(rank = row_number())

# Ranking in reverse
ranked_coldhot <- seasonal_avgs %>% 
  mutate(
    yr = factor(yr),
    year_rank = fct_reorder(
      .f = yr, 
      .x = mean_temp, 
      .fun = median, 
      .desc = FALSE)) %>% 
  arrange(year_rank) %>% 
  mutate(rank = row_number())



# Special Controls for Winter
if(params$season == "Winter"){
  ranked_hotcold <- ranked_hotcold %>% 
    mutate(flag_year = ifelse(year_rank == params$year-1, "flag", "unflag"))
  
  ranked_coldhot <- ranked_coldhot %>% 
    mutate(flag_year = ifelse(year_rank == params$year-1, "flag", "unflag"))
  
} else {
  ranked_hotcold <- ranked_hotcold %>% 
    mutate(flag_year = ifelse(year_rank == params$year, "flag", "unflag"))
  
  ranked_coldhot <- ranked_coldhot %>% 
    mutate(flag_year = ifelse(year_rank == params$year, "flag", "unflag"))
}


```



```{r}
#| label: seasonal-warming-rates

#---------- Regional Rates

# Decadal Rate
yr_mod         <- lm(temp_f ~ yr, seasonal_avgs)
annual_rate    <- coef(yr_mod)[[2]]
decadal_rate_f <- round((annual_rate*10), 2)


#---------  Global Rates

# Global Temperature Anomaly Rates
# These are produced in: 
global_anoms <- read_csv(
    # paste0(oisst_path, "global_timeseries/global_anoms_1982to2011.csv"), 
    paste0(oisst_path, "global_timeseries/global_anoms_1991to2020.csv"), 
    guess_max = 1e6,
    col_types = cols()) %>% 
  mutate(year = year(time)) %>% 
  filter(between(year, ssn_year_start, params$year))  %>% 
  pull_heatwave_events(
    temp_col =  "area_weighted_sst_anom",
    clim_ref_period = c("1991-01-01", "2020-12-23"))
  

# Pull the season, summarize by year
global_season_summary <- global_anoms %>% 
  supplement_season_info() %>% 
  filter(season_eng == params$season) %>% 
  group_by(yr, season_yr, season) %>% 
  summarise(
    sst             = mean(sst, na.rm = T),
    sst_anom        = mean(sst_anom, na.rm = T), 
    sst_f           = as_fahrenheit(sst, "temperature"),
    anom_f          = as_fahrenheit(sst_anom, "anomalies"),
    .groups         = "drop") %>% 
  arrange(desc(sst)) %>%
  #temperature_summaries() %>% 
  mutate(
    yr_as_dtime = as.Date(paste0(yr, "-07-02")),
    anom_direction = ifelse(sst_anom > 0, "Hot", "Cold"))


# Get the global temperature warming rate
# Decadal Rate
global_yr_mod         <- lm(anom_f ~ yr, global_season_summary)
global_annual_rate    <- coef(global_yr_mod)[[2]]
global_decadal_rate_f <- round((global_annual_rate*10), 2)


# Join the data we need together,
# Put the description into the dataframe so we can call it using one geom call
glob_plot_dat <- bind_rows(
  list(

    # 1. Global Ocean Seasonal Data
    "Global Oceans" = select(global_season_summary, yr, season, season_yr, temp_f = sst_f, anom_f) %>% 
      mutate(
        rate = str_c(global_decadal_rate_f, " \u00b0F / Decade"),
        rate_lab = str_c(season_range,  "Global Ocean Warming")),
    
    # 2. Gulf of Maine Seasonal Data
    "Gulf of Maine" = select(seasonal_avgs, yr, season, season_yr, temp_f, anom_f)  %>% 
      mutate(
        rate = str_c(decadal_rate_f, " \u00b0F / Decade"),
        rate_lab = str_c(season_range, " Gulf of Maine Warming"))), 
  
  # Label what area the rates are for
  .id = "region")

```



# `r params$season` `r params$year` Highlights



```{r}
#| label: season-highlights

# Average Temp+Anom Stats for that season Used in Text
season_avg_temp <- round(mean(this_season_data$sst_f, na.rm = T), 2)
season_avg_anom <- round(mean(this_season_data$anom_f, na.rm = T), 2)
season_avg_clim <- round(mean(this_season_data$seas_f, na.rm = T), 2)


# What rank are we this year?
if(params$season == "Winter"){
  hot_ranking <- filter(ranked_hotcold, yr == params$year-1) %>% 
    pull(rank)
  cold_ranking <- filter(ranked_coldhot, yr == params$year-1) %>% 
    pull(rank)
  
  } else{

  hot_ranking <- filter(ranked_hotcold, yr == params$year) %>% 
    pull(rank)
  cold_ranking <- filter(ranked_coldhot, yr == params$year) %>% 
    pull(rank)
}


# # Why are they not adding? ah, its a double count
# filter(ranked_hotcold, rank <= hot_ranking) 
# filter(ranked_coldhot, rank <= cold_ranking)
pretty_numbers <- list(
  "1" = "1st",
  "2" = "2nd","3" = "3rd","4" = "4th","5" = "5th", "6" = "6th","7" = "7th","8" = "8th","9" = "9th","10" = "10th", "11" = "11th", "12" = "12th", "13" = "13th", "14" = "14th", "15" = "15th", "16" = "16th", "17" = "17th", "18" = "18th", "19" = "19th", "20" = "20th", "21" = "21st", "22" = "22nd", "23" = "23rd", "24" = "24th", "25" = "25th", "26" = "26th", "27" = "27th", "28" = "28th", "29" = "29th", "30" = "30th")


# In-Text friendly version
hot_rank_label <- pretty_numbers[[hot_ranking]]
cold_rank_label <- pretty_numbers[[cold_ranking]]


# Get max anom year and max anom
max_anom_yr <- seasonal_avgs %>% 
  arrange(desc(mean_anom)) %>% 
  slice(1) %>% 
  mutate(across(c(mean_temp:anom_f), \(x) sprintf(signif(x, 3), fmt = '%#.2f')))

```

For this seasonal report we present an analysis of sea surface temperature (SST) for `r params$seas` `r params$year`, defined as the period extending from  `r month_range`. 

For this seasonal report we present an analysis of sea surface temperature (SST) for spring 2025, defined as the period extending from March 1, 2025 through May 31, 2025.
Springtime in the Gulf of Maine is when SSTs dip to their annual lows (typically in March), followed soon-after by a climb in temperature and the pronounced spring peak in phytoplankton productivity. Vertical mixing over the winter replenishes critical nutrients to the surface waters which, together with longer durations of sunlight, fuel the spring phytoplankton bloom. This is a very active time of year for marine species looking to take advantage of the seasonal bounty of plankton.

This year, the average SST for the Gulf of Maine was `r season_avg_temp`°F. This seasonal average temperature was `r season_avg_anom`°F above the 1991–2020 climatological reference period (CRP) `r params$seas` average of `r season_avg_clim`°F.

## Weekly Temperatures

@tbl-season-temps highlights how the SST for each week this `r params$seas` compares to the 1991-2020 CRP for the area shown in @fig-map. The observed SST, long-term average SST, and SST anomalies (i.e., departures from the long-term average SST) are shown. 


```{r}
#| label: week-summary-data-prep

# Format the start/end dates for weekly subsets
# Determine bookends for weeks and trim to season bookends
week_labels <- this_season_data  %>% 
  mutate(
    wk_end = ceiling_date(time, unit = "week"),
    wk_end = if_else(wk_end >= last_day, last_day, wk_end - 1)) %>% 
  
  # Pull distinct weeks
  distinct(time, wk_start, wk_end)  %>% 
  
  # Add the month and day labels
  mutate(
    start_month = month(wk_start, label = T, abbr = T),
    end_month   = month(wk_end, label = T, abbr = T),
    start_day   = str_pad(day(wk_start), pad = "0", side = "left", width = 2),
    end_day     = str_pad(day(wk_end), pad = "0", side = "left", width = 2),
    `Week`      = str_c(start_month, " ", start_day, " - ", end_month, " ", end_day),
    `Week`      = factor(`Week`)) %>%
  
  # Ensure distinct weeks only, and arrange in chronological order
  distinct(`Week`, .keep_all = T) %>% 
  arrange(wk_start) %>% 
  mutate(week_position = row_number()) %>% 
  select(-time)
    
    
# rejoin the temperatures to get weekly summaries
week_data <- this_season_data %>%
  select(time, sst, seas, sst_anom, wk_start) %>%
  left_join(week_labels, by = "wk_start") 


# Process mean temps
week_summaries <- week_data %>%
  group_by(`Week`, week_position) %>%
  summarise(
    temp_c = mean(sst),
    clim_c = mean(seas),
    anom_c = mean(sst_anom),
    .groups = "drop") %>% 
  mutate(`Week` = fct_reorder(`Week`, .x = week_position)) %>% 
  arrange(week_position)


# Fahrenheit formatting and column labels
week_summaries <- week_summaries %>%
  mutate(
    temp_f = as_fahrenheit(temp_c),
    clim_f = as_fahrenheit(clim_c),
    anom_f = as_fahrenheit(anom_c, data_type = "anomalies")) %>% 
  mutate(
    across(
      .cols = all_of(c("temp_c", "clim_c", "anom_c","temp_f", "clim_f", "anom_f")), 
      .fns = \(x) sprintf(signif(x, 3), fmt = '%#.2f')))


# in-text values
weekly_high <-  min(as.numeric(week_summaries$anom_f))
weekly_low <- max(as.numeric(week_summaries$anom_f))

```



`r params$seas` average SST were on average `r round(mean(this_season_data$anom_f),2)`°F below normal, with weekly SST anomalies ranging from -`r weekly_low`°F to +`r weekly_high`°F; all but one week at the end of 2023 recorded SSTs that were below the 1991-2020 average.



```{r}
#| label: tbl-season-temps
#| tbl-cap-location: bottom
#| tbl-cap: "Average weekly sea surface temperatures in the Gulf of Maine"

week_summ_table <- week_summaries %>% 
  mutate(
    temp = md(str_c(temp_f, deg_f, " *(", temp_c, deg_c, ")*")),
    clim = str_c(clim_f, deg_f, " *(", clim_c, deg_c, ")*"),
    anom = str_c(anom_f, deg_f, " *(", anom_c, deg_c, ")*"),
    Week = str_c("*", Week, "*")) %>% 
  select(Week, temp, clim, anom) %>%
   gt(rowname_col = "Week") %>% 
  tab_header(
    title = md(paste0("Weekly Average Sea Surface Temperatures - ", params$season)), 
    subtitle = month_range) %>%
    tab_stubhead(label = md("One-Week Period")) %>% 
  fmt_markdown(
  columns = everything(),
  rows = everything(),
  md_engine = c("markdown", "commonmark")) %>% 
  # Change the column names to the units
    cols_label(
      temp = md("Observed Temperature"),
      clim = md("Climatological Average"),
      anom = md("Temperature Anomaly")) %>% 
    tab_source_note(md("*Data Source: NOAA OISSTv2 Daily Sea Surface Temperature Data.*") ) %>% 
    tab_source_note(md("*Climatological Reference Period: 1991-2020.*")) %>%    
    opt_table_font(font = list(google_font(name = "Raleway")))

# Show Table
week_summ_table
  

# How to spit out the raw html:
#week_summ_table %>% gt::as_raw_html()
```

## Monthly Statistics

```{r}
#| label: monthly-temps

# Recreate Table for Monthly Averages
monthly_avgs <- this_season_data %>% 
  group_by(yr, month) %>% 
  summarise(
    temp_c = mean(sst),
    clim_c = mean(seas),
    anom_c = mean(sst_anom),
    temp_f = as_fahrenheit(temp_c),
    clim_f = as_fahrenheit(clim_c),
    anom_f = as_fahrenheit(anom_c, data_type = "anomalies"),
    temp_list = list(sst_f), # list column for sparkline anomalies
    .groups = "drop") %>% 
  mutate(across(c(temp_c:anom_f), \(x) signif(x, digits = 3))) 

# Do some benchmarking, where do they all rank by anomalies
month_rankings <- region_hw %>% 
  group_by(yr = year(time), month) %>% 
  summarise(mean_anom = mean(sst_anom), .groups = "drop") %>% 
  group_by(month) %>% 
  arrange(desc(mean_anom)) %>% 
  mutate(rank = rank(-mean_anom)) %>% 
  arrange(mean_anom) %>% 
  mutate(rev_rank = rank(mean_anom)) %>% 
  ungroup() %>% 
  arrange(rank, month)

# Add rankings back
monthly_avgs <- monthly_avgs %>% left_join(
  select(month_rankings, -c(mean_anom)),
  by = join_by(yr, month))

```


@tbl-monthly-temps shows monthly average SST for `r params$seas` `r params$year`. Each month has been ranked against all previous years from warmest (1), to lowest based on average temperature. The long-term climatological average, and the difference between observed SST's and those averages are also reported.



```{r}
#| label: tbl-monthly-temps
#| tbl-cap-location: bottom
#| tbl-cap: "Average monthly sea surface temperatures in the Gulf of Maine"

# Create the Monthly Summary Table
month_names <- data.frame("month" = month.abb, "month_full" = month.name)
month_summ_table <- monthly_avgs %>% 
  left_join(month_names) %>% 
  mutate(across(c(temp_c:anom_f),\(x) sprintf(signif(x, 3), fmt = '%#.2f'))) %>% 
  mutate(
    temp = md(str_c(temp_f, deg_f, " *(", temp_c, deg_c, ")*")),
    clim = str_c(clim_f, deg_f, " *(", clim_c, deg_c, ")*"),
    anom = str_c(anom_f, deg_f, " *(", anom_c, deg_c, ")*"),
    month = str_c("*", month_full, "*")) %>% 
  select(month, rank, temp, clim, anom) %>%
   gt(rowname_col = "month") %>% 
  tab_header(
    title = md(paste0("Monthly Average Sea Surface Temperatures - ", params$season, "")), 
    subtitle = month_range) %>%
    tab_stubhead(label = md("Month")) %>% 
  fmt_markdown(
  columns = everything(),
  rows = everything(),
  md_engine = c("markdown", "commonmark")) %>% 
  cols_label( # Change the column names to the units
    rank = md("Rank (1982-2025)"),
    temp = md("Observed Temperature"),
    clim = md("Climatological Average"),
    anom = md("Temperature Anomaly")) %>% 
  tab_source_note(md("*Data Source: NOAA OISSTv2 Daily Sea Surface Temperature Data.*") ) %>% 
  tab_source_note(md("*Climatological Reference Period: 1991-2020.*")) %>% 
  opt_table_font(font = list(google_font(name = "Raleway")))

# # Display Table
month_summ_table

```


## How Does this `r params$season` Compare to the Past?

`r params$year` was the `r hot_rank_label` warmest (`r cold_rank_label` coldest) `r params$seas` season observed in the Gulf of Maine during the 42+ years we have satellite data to analyze. Its position among years ranked more-highly is displayed below in @fig-season-rank.


```{r}
#| label: fig-season-rank
#| fig.height: 6
#| fig-cap: "A ranking of the 25 warmest seasons for the Gulf of Maine in the satellite record (1982-2025). "

# usually this height
# fig.height: 6

# How far to filter:
rank_filter_depth <- ifelse(
  hot_ranking <= 10,
  20,
  DescTools::RoundTo(hot_ranking, multiple = 5, FUN = ceiling)
)

# Do some data prep
rank_plot_data <- ranked_hotcold %>% 
  filter(rank <= rank_filter_depth) %>% 
  mutate(
    year_rank = fct_reorder(year_rank, rank, .desc = T),
    rank = fct_rev(factor(rank)),
    flag_col = ifelse(flag_year == "flag", gmri_cols("lv orange"), gmri_cols("gmri blue"))) 


# Plot them
temp_rankings_p <- rank_plot_data %>% 
  mutate(
    anom_label = ifelse(
      str_sub(anom_label, 1,1) == "-", 
      anom_label, 
      str_c(" ", anom_label)),
    anom_label = str_sub(anom_label, 1, 8)) %>% 
  ggplot() +
  geom_col(aes(x = anom_f, y = rank, fill = I(flag_col))) +
  # Year labels - Out in whitespace
  geom_text(
    mapping = aes(
      x = 4,
      y = rank,
      label = year_rank,
      color = I(flag_col)), 
    fontface = "bold",
    family = "Avenir",
    hjust = -.05,
    size = 5) +
  # Temperature Labels - Outside Bars
  geom_text(
    mapping = aes(
      x = 4.6, 
      y = rank, 
      label = anom_label, 
      group = rank, 
      color = I(flag_col)),
    family = "Avenir",
    hjust = -.05, size = 4, show.legend = F) +
  scale_x_continuous(
    labels = number_format(suffix = " \u00b0F"), 
    expand = expansion(mult = c(0.015,0.25))) +
  labs(
    x = "Seasonal SST Anomaly",
    y = "Rank",
    caption = NULL,
    title = str_c("Gulf of Maine's Warmest ", params$season, " SST Anomalies")) +
  theme(
    plot.background = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    axis.line.x = element_blank(),
    legend.position = "none", 
    plot.margin = margin(t = 20, l = 10, r = 5, b = 5)) +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.875, y_npc = 1.05, logo_height = 1, height_units = "cm")


temp_rankings_p

```


## Seasonal Trends and Long-Term Global Context

The Gulf of Maine is an area of particular interest to the scientific community because of the remarkable rate of warming it has experienced in recent years, coupled with its importance as a major driver for the regional economy.

 @fig-global-rates illustrates the average `r params$seas` temperature anomaly for each year in the Gulf of Maine since the beginning of the satellite record in 1982 overlaid with long-term SST anomaly trends for the Gulf of Maine (orange) and global oceans (blue) overlaid. 
 

```{r}
#| label: fig-global-rates
#| eval: true
#| fig-cap: "Average annual winter SSTs in the Gulf of Maine from 1982 through 2023-24 (black dots). The orange line indicates the trend for the full time series for the Gulf of Maine. The blue line indicates the trend for the full time series for the global oceans."

# Compare warming rates
global_rate_plot <- ggplot(glob_plot_dat) +
  geom_line(
    data = filter(glob_plot_dat, region == "Gulf of Maine"), 
    aes(yr, anom_f), 
    color = "gray20", linetype = 3, linewidth = 0.5, alpha = 0.5) +
  geom_point(
    data = filter(glob_plot_dat, region == "Gulf of Maine"), 
    aes(yr, anom_f), 
    color = "gray20", 
    size = 2,
    alpha = 0.8) +
  # All year warming rate for GOM
  geom_smooth(
    aes(yr, anom_f, color = str_c(region, " Trend\n", rate)), 
    formula = y~x, 
    linetype = 1,
    method = "lm",
    se = FALSE) +
  scale_color_gmri() +
  # Scales
  scale_y_continuous(
    expand = expansion(add = c(1.5, 1.5)),
    labels = number_format(suffix = " \u00b0F")) +
  scale_x_continuous(limits = c(1981, params$year+3)) +
  theme(
    legend.position = "inside",
    legend.position.inside =  c(0.35, 0.85), 
    legend.background = element_rect(fill = "white", color = "black")) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5, nrow = 1)) +
  labs(title = str_c(params$season, " SST Warming Rates"),
       x = "Year",
       y = str_c("Sea Surface Temperature Anomalies\n(", season_range, ")"),
       color = "1982-2025 Warming Rates")  +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.85, y_npc = 1.1, logo_height = 1, height_units = "cm")



 global_rate_plot
```



## Heatwave and Coldspell Conditions



```{r}
#| label: season-mhw-data

# Pull this full year's data, up until end of the season
year_hw_dat <- region_hw %>% 
  filter(
    if(params$season == "Winter") {
      time >= as.Date(str_c(params$year-1, "-01-01")) + season_start-1}
    else {time >= as.Date(str_c(params$year, "-01-01"))})

# Get number of heatwave events and total heatwave days for this year
num_hw_days <- sum(year_hw_dat$mhw_event, na.rm = T)
num_cs_days <- sum(year_hw_dat$mcs_event, na.rm = T)

# Fraction of year in hw?
hw_frac <- round(sum(this_season_data$mhw_event)/nrow(this_season_data) * 100, 2) 
cs_frac <- round(sum(this_season_data$mcs_event)/nrow(this_season_data) * 100, 2) 


# Flag heatwave Events:
hw_events <- year_hw_dat %>% 
  filter(mhw_event == T) %>% 
  group_split(mhw_event_no) %>% 
  map_dfr(function(x){
    data.frame(
      id = unique(x$mhw_event_no),
      llim = min(x$time),
      rlim = max(x$time),
      ymax = max(x$sst_f),
      min_anom = min(x$anom_f),
      max_anom = max(x$anom_f),
      mean_anom = mean(x$anom_f),
      len  = nrow(x))})
  

# Flag coldspell Events:
cs_events <- year_hw_dat %>% 
  filter(mcs_event == T) %>% 
  group_split(mhw_event_no) %>% 
  map_dfr(function(x){
    data.frame(
      id = unique(x$mcs_event_no),
      llim = min(x$time),
      rlim = max(x$time),
      ymax = max(x$mhw_thresh_f),
      min_anom = min(x$anom_f),
      max_anom = max(x$anom_f),
      mean_anom = mean(x$anom_f),
      len  = nrow(x))})
  
```



The most commonly used definition of a ["marine heatwave" (MHW)](http://www.marineheatwaves.org/all-about-mhws.html) is when daily average SST's exceeded the 90th percentile of a climatological (i.e., 30-year) average for at least 5 consecutive days. Gaps of 2 days or less in this threshold do not constitute a break in the MHW event. 

Using this broadly accepted definition, **the Gulf of Maine has experienced MHW conditions for `r round(hw_frac)`% of days this `r params$seas`**. 




```{r}
#| label: fig-season-mhw-timeline
#| fig-height: 5
#| eval: true
#| fig-cap: "A timeseries of marine heatwave (MHW) conditions in the Gulf of Maine extending from January 1, 2025 through May 30, 2025. Black lines represent the long-term (i.e., 1991 – 2020) average SST, the 10th percentile, and 90th percentile for a given day in the Gulf of Maine; a solid line (red for marine heatwave or blue for a non-event) indicate the observed SST this year; red (above 90th percentile) and blue (below 90th percentile) shading illustrates how far the observed SST is from the climatological average."


# Grab the data we're using
temp_units <- "F"

# Augment Data with cold spell limit and labels
year_hw_dat <- mutate(
  year_hw_dat)

  
# Handling Temperature Units:
temp_ops <- switch (
  EXPR = temp_units,
  "C" = list(
    temp_col  = expr(sst),
    anom_col  = expr(sst_anom),
    seas_col  = expr(seas),
    hw_thresh = expr(mhw_thresh),
    cs_thresh = expr(mcs_thresh),
    hw_temp   = expr(hwe),
    temp_suff = " \u00b0C"),
  "F" = list(
    temp_col  = expr(sst_f),
    anom_col  = expr(anom_f),
    seas_col  = expr(seas_f),
    hw_thresh = expr(mhw_thresh_f),
    cs_thresh = expr(mcs_thresh_f),
    hw_temp   = expr(hwe_f),
    temp_suff = " \u00b0F"))
  
# Assign to shorter names
temp_col      <- temp_ops$temp_col
anom_col      <- temp_ops$anom_col
clim_col      <- temp_ops$seas_col
hw_thresh_col <- temp_ops$hw_thresh
cs_thresh_col <- temp_ops$cs_thresh
hw_temp_col   <- temp_ops$hw_temp

# Set colors by name
#"#B2182B" "#D6604D" "#F4A582" "#FDDBC7" "#F7F7F7" "#D1E5F0" "#92C5DE" "#4393C3" "#2166AC"
color_vals <- c(
  "SST" = "#0571B0", 
  "Marine Heatwave" = "darkred",
  "Marine Coldspell" = "#92C5DE")
  

  
  
# # Make new breaks so expanded area text doesn't ruin everything...
# Set limits that work for winter
if(params$season == "Winter"){
  hw_plot_lims <- as.Date(
    c(str_c(params$year-1, "-12-01"), 
      str_c(params$year, "-04-30")))} else {
  hw_plot_lims <- as.Date(str_c(params$year, c("-01-01", "-12-31")))
}



# Don't use textpath, make a label df
clim_labels <- year_hw_dat %>% 
  filter(time == max(time)) %>% 
  mutate(time = time+3) %>% 
  select(time, {{hw_thresh_col}}, {{clim_col}}, {{cs_thresh_col}}) %>% 
  pivot_longer(
    cols = -time, 
    names_to = "clim_thresh", 
    values_to = "temp_vals") %>% 
  mutate(labs = case_when(
    str_detect(clim_thresh, "mhw") ~ "- 90th Percentile",
    str_detect(clim_thresh, "cs") ~ "- 10th Percentile",
    str_detect(clim_thresh, "seas") ~ "- Climatological Average"
  ))





# Build the Base Plot
hw_temp_p <- ggplot(year_hw_dat) +
    geom_path(
      aes(x = time, y = {{temp_col}}, 
          color = "SST", group = 1), 
      linewidth = 1.2, key_glyph = "timeseries") +
    geom_path(
      data = filter(year_hw_dat, mcs_event),
      aes(x = time, y = {{temp_col}}, 
          color = "Marine Coldspell", group = 1), 
      linewidth = 1.2, key_glyph = "timeseries") +
    geom_path(
      data = filter(year_hw_dat, mhw_event),
      aes(x = time, y = {{temp_col}}, 
          color = "Marine Heatwave", group = 1), 
      linewidth = 1.2, key_glyph = "timeseries") +
    geom_segment(
      aes(x = time, xend = time, y = {{clim_col}}, yend = {{temp_col}}, 
          color = mhw_event_txt), alpha = 0.25,
                 show.legend = F) +
    scale_y_continuous(
      labels =  number_format(suffix = temp_ops$temp_suff),
      expand = expansion(mult = c(0.05, .2))) +
    theme(legend.position = "bottom") +
    labs(
      title = str_c("Gulf of Maine SST and MHW Events"),
      x = NULL, 
      color = "Sea Surface Temperature & Heatwave Status:",
      fill = "Difference from Average SST",
      y = "Sea Surface Temperature") +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.85, y_npc = 1.1, logo_height = 1, height_units = "cm")




# Add the climatological reference lines
hw_temp_p <- hw_temp_p +
    geom_line(aes(x = time, y = {{hw_thresh_col}}), color = "gray40", lty = 1) + 
    geom_line(aes(x = time, y = {{clim_col}}), color = "gray10", lty = 1) +
    geom_line(aes(x = time, y = {{cs_thresh_col}}), color = "gray40", lty = 1) +
    geom_text(
      data = clim_labels, 
      aes(time, temp_vals, label = labs), family = "Avenir", hjust = 0,
      size = 3) +
    scale_color_manual(values = color_vals) +
    scale_x_date(
      date_labels = "%b",
      date_breaks = "1 month",
      # expand = expansion(add = c(0,110)))
      expand = expansion(add = c(0,90)))



# This only works if there are heatwaves
if(nrow(hw_events) > 0){
  hw_temp_p <- hw_temp_p +
    geom_bracket(
      data = hw_events,
      aes(xmin = llim, xmax = rlim, y.position = ymax + 2, label = str_c(len, " Days")),
      family = "Avenir", label.size = 3)
}
if(nrow(cs_events) > 0){
  hw_temp_p <- hw_temp_p +
    geom_bracket(
      data = cs_events, 
      aes(xmin = llim, xmax = rlim, y.position = ymax + 2, label = str_c(len, " Days")),
      family = "Avenir", label.size = 3)
}


hw_temp_p
```

 
```{r}
#| label: last time no mhw?
#| eval: false

# # Checking how many mhw events have happened by this point in the year:
# # 2025, 2019, 2017, 2014, 2011 are cold years
# region_hw %>% filter(doy <= max(this_season_data$doy)) %>% 
#   group_by(yr) %>% 
#   summarise(n_hws = n_distinct(mhw_event_no, na.rm = T)) %>% 
#   arrange(desc(yr))
```



## Alternative Approaches for MHW Identification

The way extreme events are identified has become a topic of discussion, and the merits of using a fixed climatological baseline under a changing climate are being questioned.

With waters warming over time due to climate change, the use of a fixed climatological baseline can lead to a predictable pattern in hot and cold extreme events. Under warming, extreme cold events cluster at the beginning of timeseries, and extreme heat events cluster towards recent years.

There is a tradeoff in choosing between these two approaches (fixed baselines vs. adaptive thresholds). Under the fixed baseline approach there is an importance that temperatures have moved away from long-term conditions.

```{r}
#| label: run detrended heatwaves

# De-trended heatwaves?
heatwaves_detrended <- pull_heatwave_events(
    temperature_timeseries = region_timeseries,
    threshold = 90, 
    clim_ref_period = c("1991-01-01", "2020-12-31"), detrend = T) %>% 
    supplement_hw_data() %>% 
    filter(
      doy != 366, 
      time <= as.Date(str_c(params$year, season_right))) %>% 
  mutate(
    year = year(time),
    yday = yday(time),
    flat_date = as.Date(yday-1, origin = base_date),
    yr_rev = factor(year),
    yr_rev = fct_rev(yr_rev),
    season_alpha = if_else(
      between(flat_date, as.Date("2000-01-01"), as.Date("2000-02-28")) |
       between(flat_date, as.Date("2000-12-01"), as.Date("2000-12-31")),
      1,
      0.4),
    hwcs_event_alpha = if_else(mhw_event|mcs_event, 1, 0.6),
    hw_date = if_else(mhw_event, flat_date, NA),  
    cs_date = if_else(mcs_event, flat_date, NA))


# Take annual average data, with baselines added
static_annual <- region_hw %>% 
  filter(between(yr, 1982,2024),
         doy != 366) %>% 
  group_by(yr) %>% 
  summarise(
    across(
      .cols = c(sst_f, anom_f, seas_f, mhw_thresh_f, mcs_thresh_f), 
      .fns = mean),
    .groups = "drop") %>% 
  mutate(
    clim_yrs = if_else(between(yr, 1991, 2020), T, F),
    clim_size = if_else(clim_yrs, 2, 0.8),
    clim_color = if_else(clim_yrs, gmri_cols("gmri blue"), "gray20"),
    mhw = if_else(sst_f > mhw_thresh_f, T, F),
    clim_color = if_else(mhw, gmri_cols("lv orange"), clim_color),
    clim_size = if_else(mhw, 3, clim_size),
    mcs = if_else(sst_f < mcs_thresh_f, T, F),
    clim_color = if_else(mcs, "#2166AC", clim_color),
    clim_size = if_else(mcs, 3, clim_size))



# Repeat for detrended, update the thresholds to include trend
detrend_annual <- heatwaves_detrended %>% 
  filter(between(yr, 1982,2024)) %>% 
  group_by(yr) %>% 
  mutate(
    mhw_trend_thresh_f = mhw_thresh_f + trend_pred,
    mcs_trend_thresh_f = mcs_thresh_f + trend_pred) %>%  
  summarise(
    across(
      .cols = c(sst_f, anom_f, seas_f, mhw_thresh_f, mcs_thresh_f, mhw_trend_thresh_f, mcs_trend_thresh_f), 
      .fns = mean))  %>% 
  mutate(clim_yrs = if_else(between(yr, 1991, 2020), T, F))  %>% 
  mutate(
    clim_yrs = if_else(between(yr, 1991, 2020), T, F),
    clim_size = if_else(clim_yrs, 2, 0.8),
    clim_color = if_else(clim_yrs, gmri_cols("gmri blue"), "gray20"),
    mhw = if_else(sst_f > mhw_trend_thresh_f, T, F),
    clim_color = if_else(mhw, gmri_cols("lv orange"), clim_color),
    clim_size = if_else(mhw, 3, clim_size),
    mcs = if_else(sst_f < mcs_trend_thresh_f, T, F),
    clim_color = if_else(mcs, "#2166AC", clim_color),
    clim_size = if_else(mcs, 3, clim_size))


# Conceptual Plot for static baseline climatology
static_baseline_concept <- ggplot(static_annual) +
  geom_ribbon(
    aes(ymin = mcs_thresh_f, ymax = mhw_thresh_f, x = yr), 
    fill = gmri_cols("gmri blue"), color = gmri_cols("gmri blue"), alpha = 0.15) +
  geom_line(aes(yr, sst_f), linewidth = 0.6, alpha = 0.3, color = "gray20") +
  geom_line(
    data = filter(static_annual, clim_yrs),
    aes(yr, sst_f), color = gmri_cols("gmri blue"), linewidth = 1) +
  geom_point(
    aes(yr, sst_f, size = I(clim_size), color = I(clim_color))) +
  scale_color_manual(values = gmri_cols("gmri blue")) +
  scale_x_continuous(expand = expansion(add = c(1,1))) +
  scale_y_continuous(
    expand = expansion(add = c(3,3)),
    labels = label_number(suffix = deg_c)) +
  theme(
    legend.position = "none", 
    plot.margin = margin(t = 1, b = 1, r = 3, l = 3)) +
  labs(
    x = "Year",
    y = "SST",
    color = "",
    title = "Static Thresholds for Extreme Events",
    subtitle = "Climatological Reference period determines thresholds for extreme events")



# Conceptual Plot for adaptive baseline heatwaves
detrend_concept <- ggplot(detrend_annual) +
  geom_ribbon(
    aes(ymin = mcs_trend_thresh_f, ymax = mhw_trend_thresh_f, x = yr), 
    fill = gmri_cols("warm yellow"), color = gmri_cols("warm yellow"), alpha = 0.15) +
  geom_line(aes(yr, sst_f), linewidth = 0.6, alpha = 0.3, color = "gray20") +
  geom_line(
    data = filter(detrend_annual, clim_yrs),
    aes(yr, sst_f), color = gmri_cols("gmri blue"), linewidth = 1) +
  geom_point(
    aes(yr, sst_f, size = I(clim_size), color = I(clim_color))) +
  scale_color_manual(values = gmri_cols("gmri blue")) +
  scale_x_continuous(expand = expansion(add = c(1,1))) +
  scale_y_continuous(
    expand = expansion(add = c(3,3)),
    labels = label_number(suffix = deg_c)) +
  theme(
    legend.position = "none", 
    plot.margin = margin(t = 1, b = 1, r = 3, l = 3)) +
  labs(
    x = "Year",
    y = "SST",
    color = "",
    title = "Adaptive Thresholds for Extreme Events",
    subtitle = "Extreme event thresholds are adjusted for long-term warming trend")
```


```{r}
#| fig.height: 5
#| label: concept comparison

concept_p <- static_baseline_concept | detrend_concept

(static_baseline_concept + labs(x = "")) / detrend_concept

```


## Temperature Anomalies and Extreme Events

Looking at the full record of daily SST anomalies in the Gulf of Maine (@fig-mhw-heatmap), the distinct thermal regime shift beginning around 2010 is evident. 

If we use the standard MHW definition, where heatwaves and coldspells are determined using a static 30-year baseline we get one record of these extreme events. This has traditionally been the record we've reported on, and has a predictable pattern of more MHW events in recent years.

If instead, we adopt an approach which removes the long-term warming trend we get a slightly different figure. After removal of the long-term linear trend a "cold period" emerges prior to the sudden warming in 2010, immediately followed by MHW events in 2010 & 2012.




```{r}
#| eval: true
#| label: plot both heatmaps

# Does it look different with the season highlighted
hw_dat <- grid_dat %>%
  mutate(
    season_alpha = if_else(
      between(flat_date, as.Date("2000-01-01"), as.Date("2000-02-28")) |
       between(flat_date, as.Date("2000-12-01"), as.Date("2000-12-31")),
      1,
      0.4),
    hwcs_event_alpha = if_else(mhw_event|mcs_event, 1, 0.6))



#"#B2182B" "#D6604D" "#F4A582" "#FDDBC7" "#F7F7F7" "#D1E5F0" "#92C5DE" "#4393C3" "#2166AC"
fixed_hw <- hw_dat %>% 
   mutate(
     hw_date = if_else(mhw_event, flat_date, NA),  
     cs_date = if_else(mcs_event, flat_date, NA)) %>% 
    ggplot(aes(x = flat_date, y = year)) +
    # tile for sst colors
    geom_tile(
      aes(fill = {{ anom_col }}), 
      alpha = 0.8,
      color = "transparent") +
    # lines for heatwave events
    geom_line(
      aes(x = hw_date, y = year, 
          group = str_c(mhw_event_no, year)), 
      color = "white", linewidth = 1)+
    geom_line(
      aes(x = hw_date, y = year, 
          group = str_c(mhw_event_no, year)), 
      color = "#B2182B", linewidth = 0.6) + 
    geom_line(
      aes(x = cs_date, y = year, 
          group = str_c(mcs_event_no, year)), 
      color = "white", linewidth = 1)+
    geom_line(
      aes(x = cs_date, y = year, 
          group = str_c(mcs_event_no, year)), 
      color = "#2166AC", linewidth = 0.6) + 
    scale_fill_distiller(
      palette = "RdBu",
      na.value = "transparent",
      limit = c(-6, 6),
      oob = scales::squish,
      breaks = seq(-6, 6, 2),
      labels = label_number(suffix = deg_c)) +
    scale_x_date(
      date_labels = "%b", 
      date_breaks = "1 month", 
      expand = expansion(add = c(0,0))) +
    scale_y_continuous(limits = c(1982 + .5, 2025 + .5), expand = expansion(add = c(0,0))) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(angle = 0, size = 11, face = "bold"),
      plot.margin = margin(b = 0, t = 1, l = 3, r = 3),
      axis.text.x = element_text(size = 10)) +
    guides(fill = guide_colorbar(
      title = "Sea Surface Temperature Anomaly",
      title.position = "top", 
      title.hjust = 0.5, barwidth = unit(3.5, "in"), 
      barheight = unit(0.5, "cm"), 
      frame.colour = "black", 
      ticks.colour = "black")) +
    labs(
      title = "Gulf of Maine Heatwave & Coldspell Record",
      subtitle = "Fixed Climatological Baseline (1991-2020)\nNo De-Trending",
      y = "Year",
      x = "Month",
      fill = "Sea Surface Temperature Anomaly")



# Plot the detrended heatwaves
detrend_hw <- heatwaves_detrended %>% 
   ggplot(aes(x = flat_date, y = year)) +
    # tile for sst colors
    geom_tile(
      aes(fill = {{ anom_col }}), alpha = 0.8,
      color = "transparent") +
    # lines for heatwave events
    geom_line(
      aes(x = hw_date, y = year, 
          group = str_c(mhw_event_no, year)), 
      color = "white", linewidth = 1)+
    geom_line(
      aes(x = hw_date, y = year, 
          group = str_c(mhw_event_no, year)), 
      color = "#B2182B", linewidth = 0.6) + 
    geom_line(
      aes(x = cs_date, y = year, 
          group = str_c(mcs_event_no, year)), 
      color = "white", linewidth = 1)+
    geom_line(
      aes(x = cs_date, y = year, 
          group = str_c(mcs_event_no, year)), 
      color = "#2166AC", linewidth = 0.6) + 
    scale_fill_distiller(
      palette = "RdBu",
      na.value = "transparent",
      limit = c(-6, 6),
      oob = scales::squish,
      breaks = seq(-6, 6, 2),
      labels = label_number(suffix = deg_c)) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(add = c(0,0))) +
    scale_y_continuous(limits = c(1982 + .5, 2025 + .5), expand = expansion(add = c(0,0))) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(angle = 0, size = 11, face = "bold"),
      plot.margin = margin(b = 0, t = 1, l = 3, r = 3),
      axis.text.x = element_text(size = 10)) +
    guides(fill = guide_colorbar(
      title = "Sea Surface Temperature Anomaly",
      title.position = "top", 
      title.hjust = 0.5, barwidth = unit(3.5, "in"), 
      barheight = unit(0.5, "cm"), 
      frame.colour = "black", 
      ticks.colour = "black")) +
    labs(
      title = "Gulf of Maine Heatwave & Coldspell Record",
      subtitle = "Fixed Climatological Baseline (1991-2020)\nWarming Trend Removed",
      y = "Year",
      x = "Month",
      caption = "Heatwave (red) and Coldspell (blue) events have been overlayed with lines for distinction.",
      fill = "Sea Surface Temperature Anomaly")


```


```{r}
#| fig.height: 8
#| label: fig-static-baseline-heatmap
#| fig-cap: "Heat map of daily SST anomalies from a 30-year 1991-2020 climatological baseline. Marine heatwave events (red lines) and marine coldspell events (blue lines) using the standard Hobday et al. method have been overlaid."
#| eval: false


# fixed_hw + labs(subtitle = "")
```



```{r}
#| fig.height: 8
#| label: fig-shifting-baseline-heatmap
#| eval: false
#| fig-cap: "Heat map of daily SST anomalies from a 30-year 1991-2020 climatological baseline. Marine heatwave events (red lines) and marine coldspell events (blue lines) using a detrended timeseries have been overlaid."

# detrend_hw


```




```{r}
#| fig.height: 8
#| fig.width: 8
#| label: fig-mhw-heatmap
#| fig-cap: "Heat maps of daily SST anomalies from a 30-year 1991-2020 climatological baseline. Marine heatwave events (red lines) and marine coldspell events (blue lines) contrasting two competing methodologies have been overlaid."


# Combine them
heatmap_p <- (
  (fixed_hw) | 
    (detrend_hw + labs(title = "", caption = ""))) + 
  plot_layout(guides = "collect") & theme(
    legend.position = "bottom", 
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8)) 

# # Without guides
heatmap_p

# # With conceptual guides
# (heatmap_p / concept_p) + plot_layout(heights = c(5,3))


```




A comparison of the two above approaches shows that the methodology choices will ultimately impact the MHW/MCS record for the region. 

Long-term linear trends can be thought of as a ramp, with a steady rise over time. The slope of the ramp is estimated using all the change over the full time period, and a single slope is returned. This slope is then removed from the whole timeseries.

The sudden increase in SSTs which happened in 2010 and which persisted over the next ~13 years has a charactistic "stair-step" shape, not the steady sustained climb a linear trend would capture. When using the MHW approach which removes the linear-trend, the periods immediately before and after the stair-step are highlighted as extreme events. 

There is ongoing debate on the merits of either approach, and discussions on whether regional dynamics (e.g. the North Atlantic Oscillation, ENSO, Gulf Stream Position) should be accounted for when identifying marine heatwaves.

For now, I've presented both for comparison. Both approaches are valid approaches, but this side-by-side comparison should highlight how our senses of the occurrence and severity of extreme events is dependent on the approaches used.


## Spatial Distribution of Seasonal Anomalies

```{r}
#| label: fig-seasonal-anom-stack


# Set time limits
season_dates <- c(min(this_season_data$time), max(this_season_data$time))

# Spatial Extent Info
x_buffer <- c(-2.75, 3.5)
y_buffer <- c(-2, 1)
crop_x   <- st_bbox(region_extent)[c(1,3)] + x_buffer
crop_y   <- st_bbox(region_extent)[c(2,4)] + y_buffer

# Expanded area as bbox
expanded_area <- structure(
    c(crop_x, crop_y),  
    class = "bbox",
  crs = 4326) %>% 
  st_as_sfc() %>% 
  st_as_sf()


# Make data window for season
data_window <- data.frame(
  lon = crop_x,
  lat = crop_y,
  time = season_dates)

# Load data - Temperatures
season_anoms_c <- oisst_window_load(
  data_window = data_window, 
  anomalies = T, 
  climate_ref = "1991-2020",
  box_location = "cloudstorage") %>%  
  stack()

# Convert to F
season_anoms_f <- as_fahrenheit(season_anoms_c, data_type = "anomalies")





####  Records for the Text  ####

# Calculate the average for the season
season_mean <- mean(season_anoms_f, na.rm = T)

# Get the Maximum Daily Temperature Value
max_daily_actual  <- max(values(season_anoms_f), na.rm = T)
max_daily  <- round(max(values(season_anoms_f), na.rm = T), 2)

# Where was it seasonally the highest: Flag the exact location

# If we wanted it to be in box we have to use this:
season_mean_within <- crop(season_mean, region_extent)

# Get the location
idx <- which.max(crop(season_mean_within, region_extent))
pos <- as.data.frame(xyFromCell(season_mean_within, idx))

# Get the values
# Get the maximum seasonal average temperature for each pixel
max_season <- round(max(values(season_mean_within), na.rm = T), 2)

# Make some table with context
set.seed(123)
peak_temp_df <- data.frame(
  max_val = str_c(round(max_season, 2), deg_f),
  lon = pos$x,
  lat = pos$y,
  do_circle = T) %>% 
  mutate(descr = "Highest Seasonal SST Anomaly") %>% 
  bind_rows(
    data.frame(
      max_val = rep(NA, 25),
      lon = rnorm(25, mean = -68.25, sd = 2),
      lat = rnorm(25, mean = 43, sd = 2),
      do_circle = rep(F, 25),
      descr = rep("", 25)
    )
  )


```


From an aerial perspective, the Gulf of Maine experienced below-average SST's during `r params$seas` of `r params$year`. Part of a broader regional pattern.



```{r}
#| label: fig-season-anom-map
#| fig.width: 8
#| fig.height: 8
#| fig-cap: "Map of average SST anomalies for each grid cell in the satellite record for spring 2023. The box outlined by the black dashed line denotes the region of study for the analysis (see @fig-map). Darker red regions indicate warmer anomalies. Black contours have been added at 100m and 200m depth"

# Make the seasonal average a stars class to plot with ggplot
anoms_st <- st_as_stars(season_mean)


# Set Plot legend limits
temp_limits <- c(-6, 6)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0F")



#### 1. Map the Anomalies in Space
season_map <- ggplot() +
  geom_stars(data = anoms_st) +
  tidyterra::geom_spatraster_contour(
    data = elev_terra,
    breaks = seq(0,-200,-100),
    alpha = 0.75,
    linewidth = 0.25,
    color = "gray20") +
  geom_sf(data = new_england, fill = "gray90", linewidth = .25) +
  geom_sf(data = canada, fill = "gray90", linewidth = .25) +
  geom_sf(data = greenland, fill = "gray90", linewidth = .25) +
  geom_sf(data = region_extent, 
          color = "black", 
          linetype = 2, 
          linewidth = 1,
          fill = "transparent") +
  # Peak anomalies
  # geom_point(
  #   data = drop_na(peak_temp_df),
  #   aes(lon, lat), color = "black", shape = 22, fill = "transparent", size = 5) +
  # geom_mark_ellipse(
  #   data = peak_temp_df,
  #   aes(lon, lat, 
  #       filter = do_circle == T,
  #       label = max_val, description = descr), 
  #   label.family = "Avenir",
  #   color = "black", 
  #   #label.fill = "#FAFAFA90") +
  #   label.fill = "#E9E9E990") +
  scale_y_continuous(breaks = seq(30, 50, by = 2)) +
  scale_fill_distiller(
    palette = "RdBu", 
    na.value = "transparent", 
    limit = temp_limits,
    breaks = temp_breaks,
    labels = temp_labels,
    oob = scales::oob_squish) +
  map_theme(
    text = element_text(family = "Avenir"),
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 9)) +
  coord_sf(xlim = crop_x, 
           ylim = crop_y, 
           expand = F) +
  guides("fill" = guide_colorbar(
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(4.5, "in"),
    frame.colour = "black",
    ticks.colour = "black")) +
  labs(
    title = str_c("Seasonal Sea Surface Temperature Anomaly\n", params$season, " ", params$year),
    fill = "Surface Temperature Anomaly")

# Show figure
season_map

```

## Monthly Temperature Anomalies

The aerial distributions of average monthly SST anomalies are shown in @fig-monthly-avg-map. 


```{r}
#| label: fig-monthly-avg-map
#| eval: true
#| fig.width: 8
#| fig.height: 4
#| fig.align: center
#| fig-cap: "This series of maps shows the average monthly SST anomaly for December 2023, January 2025, and May 2025. The box outlined by the black dashed line denotes the region of study for the analysis. Darker red regions indicate warmer anomalies."


# Get monthly averages
season_nums    <- unique(str_sub(names(season_anoms_f), 7,8))
month_names    <- month.name[as.numeric(season_nums)]


# Pull the layers for each and average them:
display_months <- map(season_nums, function(month_idx){
  
  which_dates <- which(str_sub(names(season_anoms_f), 7,8) == month_idx)
  month_mean <- mean(season_anoms_f[[which_dates]], na.rm = T)}) %>% 
  setNames(month_names)


# Perform masking
months_masked <- map(
  display_months, 
  mask_nc, 
  expanded_area, 
  rotate = FALSE)


#------- make plots

# Set contours to show
# Show less contours b/c smaller map
contours_make <- c(-200)


# Make plots for each
month_plots <- imap(
  months_masked, 
  function(x,y){
    monthly_sst_map(
      month_avg_layer = x, 
      month_id = y,
      plot_yr = params$year,
      depth_contours = contours_make, 
      temp_lim = 6,
      convert_to_f = FALSE,
      elev_terra =  elev_terra)
  }
) 
  
                    


# assemble aggregate plot
monthly_maps <- (
  month_plots[[1]] |
    month_plots[[2]] | 
    month_plots[[3]]) + 
  plot_layout(guides = "collect")  & 
  theme(
    text = element_text(family = "Avenir"),
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 9),
    plot.margin = margin(t = 0, b = 0, r = 0.5, l = 0.5)) 

# Add a caption
monthly_maps + plot_annotation(caption = NULL)

```


### Checking Bottom Temperatures

There is a growing understanding that local oceanographic dynamics in the Gulf of Maine are influenced by the relative mixtures of water masses entering the Gulf of Maine. 

There are two primary deep-water masses entering the Gulf of Maine which enter the area through the Northeast Channel: 
  1. "Warm Slope Water" (WSW), which is positioned between the Gulf Stream and the continent shelf break 
  2. "Labrador Slope Water" (LSW) which is an extension of the labrador current and flows southwest following the scotian shelf shelf-break.
  
There is also a third off-shelf water mass, the "Gulf Stream Water" (GSW), which as its name implies closely follows the Gulf Stream position. GSW contributes less directly to the mixed water entering the Gulf of Maine, but there is evidence of it directly contributing to the NE channel inflow. 

Closer to shore, on the landward side of the LSW over the continental shelf, more cold-water is transported along Scotian Shelf (SS). SS water follows the coastline of Nova Scotia and enters the Gulf of Maine before continuing up towards the Bay of Fundy.

These water masses each have different temperature, salinity, nutrient, and plankton community characteristics which all play a role in shaping downstream Gulf of Maine characteristics.

If we follow the bottom temperatures of the Gulf of Maine we can get some proxy of whether the inflow through the Northeast Channel favoring more WSW vs. LSW.

```{r}
glorys_path <- cs_path("res", "GLORYS/GOM_Timeseries")
glorys_bt <- read_csv(str_c(glorys_path, "Glorys_surfacebottom_gom_ts.csv")) %>% select(-1)


ggplot(glorys_bt) +
  geom_line(aes(time, surface_temp, color = "Surface Temperature")) +
  geom_line(aes(time, bottom_temp, color = "Bottom Temperature")) +
  #geom_smooth(aes(time, bottom_temp), method = "lm", color = "black") +
  facet_zoom(ylim = c(3.75, 6.5), zoom.size = 3) +
  scale_color_gmri() +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = label_number(suffix = deg_c)) +
  labs(
    title = "Gulf of Maine Bottom Temperatures",
    subtitle = "GLORYS Ocean Physics Reanalysis",
    color = "Temperature Record",
    y = "Temperature",
    x = "Year")



# glorys_annual <- glorys_bt %>% 
#   filter(time < as.Date("2025-01-01")) %>% 
#   group_by(year = year(time)) %>% 
#   summarise(across(-time, mean))
# 
# 
# ggplot(glorys_annual) +
#   # geom_line(aes(year, surface_temp, color = "Surface Temperature")) +
#   geom_line(aes(year, bottom_temp, color = "Bottom Temperature")) +
#   geom_smooth(aes(year, bottom_temp), method = "lm", color = "black") +
#   scale_color_gmri() +
#   theme(legend.position = "bottom")
```



## About the Updates:

Over the past decade, scientists have led a body of research that highlights the rapid pace of warming in the Gulf of Maine. To help keep you informed, we share seasonal and annual updates about conditions in the Gulf of Maine. For a full compilation of all of GMRI’s seasonal and annual Gulf of Maine warming updates, visit: https://gmri.org/warmingupdates/.


## The Gulf of Maine Region

For analyses like these, it is important to be clear about the spatial extent that “defines” the Gulf of Maine (@fig-map), as different borders could produce different results. The spatial domain we use as the “Gulf of Maine” is outlined with the dashed line below. This area is consistent with previous seasonal and annual SST reports that scientists at GMRI have produced.



```{r}
#| label: fig-map
#| eval: true
#| fig.width: 8
#| fig.height: 8
#| fig.cap: "Spatial domain used for Gulf of Maine SST analyses. Depth contours are colored at 100 m intervals up to 600 m; deeper blues indicate deeper water depths"
#| fig.alt: "An overhead view of the Gulf of Maine region. Landmasses and political boundaries for the United States and Canada are displayed. A blue dotted line outlining a box with a transparent blue fill is shown to demark where satellite data for the analysis has been used."



# Map the Gulf of Maine with Bathymetry
gom_extent_p <- map_study_area_color(
  region_extent = region_extent,
  new_england_sf = new_england,
  canada_sf = canada,
  greenland_sf = greenland) +
  labs(
    title = "Gulf of Maine Study Area",
    caption = NULL) +
  theme(text = element_text(family = "Avenir"))


# Show figure
gom_extent_p

```


### Data Sources:


> NOAA High Resolution SST data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html.



### Citing This Work

If you would like to cite this report, please use: 

> Gulf of Maine Research Institute. 2025. Gulf of Maine Warming Update: `r params$season` 2025 

### Note About the Data: 

The figures in this report are created using remotely-sensed satellite data as part of publicly funded research efforts. Satellite SST data was obtained from NOAA’s National Center for Environmental Information (NCEI), with all maps and figures displaying NOAA’s Optimum Interpolation Sea Surface Temperature Data. 

 

```{r}
#| label: figure-saving
#| echo: false


# Folder for Comms Team Update
comm_folder <- cs_path(
  "root", "Program Communications/Climate Center Comms/Warming Updates/Visuals/")
update_folder <- str_c(comm_folder, params$season, " ", params$year, "/")

# # For testing locally
# comm_folder <- here::here("fig_testing/")


# Export Figures
if(params$save_figures){
  
  # Prevents tiny fonts when saving
  showtext::showtext_opts(dpi=300) 
  
  # Gom Extent Map
  ggsave(
    plot = gom_extent_p,
    filename = str_c(update_folder, "GOM_Extent.png"),
    height = unit(3.2, "in"),
    width = unit(3.2, "in"),
    dpi = "retina",
    bg = "transparent",
    scale = 2.5)
  
  
  # Weekly Report Table
  gtsave(data = week_summ_table %>%
    opt_table_font(
      font = list(google_font(name = "Avenir"))),
    filename = str_c(update_folder, "weekly_summary_table.png"),
    vwidth = 800,
    vheight = 400)

  # Monthly Report Table
  gtsave(
    data = month_summ_table %>%
      opt_table_font(font = list(google_font(name = "Avenir"))),
    filename = str_c(update_folder, "monthly_summary_table.png"),
    vwidth = 800,
    vheight = 100)
  
  #--------  HTML Tables  ----------------
    
  
  # Weekly Report Table html
  gtsave(
    data = week_summ_table %>%
      opt_table_font(font = list(google_font(name = "Avenir"))),
    filename = str_c(update_folder, "weekly_summary_table.html"))

  # Monthly Report Table html
  gtsave(
    data = month_summ_table %>%
      opt_table_font(font = list(google_font(name = "Avenir"))),
    filename = str_c(update_folder, "monthly_summary_table.html"))
    
  
  # For HTML
  # Add this to table tag as inline style to center:
  #  style="margin-left:auto; margin-right:auto;"
  # Or add it to the .gt_table class controls
  
  
  
  
  # Seasonal Rates Plot
  ggsave(
    plot = global_rate_plot, 
    filename = str_c(update_folder, "Seasonal_rates_plot.png"), 
    height = unit(2, "in"),
    width = unit(3.2, "in"),
    dpi = "retina", 
    bg = "white", 
    scale = 2.5)
  
  
  # Seasonal Temp Rankings
  ggsave(
    plot = temp_rankings_p, 
    filename = str_c(update_folder, "Seasonal_temp_ranks_plot.png"), 
    height = unit(2.4, "in"),
    width = unit(3.2, "in"),
    dpi = "retina", 
    bg = "white", 
    scale = 2.5)

  
  # Heatwave Temperature Plot
  ggsave(
    plot = hw_temp_p, 
    filename = str_c(update_folder, "Heatwave_temp_plot.png"), 
    height = unit(2, "in"),
    width = unit(3.2, "in"),
    dpi = "retina", 
    bg = "white", 
    scale = 2.5)
  
  
  # # Heatwave Anomaly Plot
  # ggsave(
  #   plot = hw_temp_p, 
  #   filename = str_c(update_folder, "Heatwave_anom_plot.png"), 
  #   height = unit(2, "in"),
  #   width = unit(3.2, "in"),
  #   dpi = "retina", 
  #   bg = "white", 
  #   scale = 2.5)
  
  # heatwave heatmap
  ggsave(
    plot = (fixed_hw + theme(plot.subtitle = element_blank())), 
    filename = str_c(update_folder, "Heatwave_Heatmap.png"), 
    height = unit(3.2, "in"),
    width = unit(3.2, "in"),
    dpi = "retina", 
    bg = "white", 
    scale = 2.5)
  
  
  # Season Map
  ggsave(
    plot = season_map, 
    filename = str_c(update_folder, "Seasonal_anom_map.png"), 
    height = unit(3.2, "in"),
    width = unit(3.2, "in"),
    dpi = "retina", 
    bg = "white", 
    scale = 2.5)
  
  # monthly maps
  ggsave(
    plot = monthly_maps,
    filename = str_c(update_folder, "Monthly_anom_maps.png"), 
    height = unit(2, "in"),
    width = unit(3.2, "in"),
    dpi = "retina", 
    bg = "white", 
    scale = 2.5)
  
  
}


```



```{r}
#| label: data-save
#| echo: false

####  Saving the Data  ####

if(params$save_data){
  

  # Rename so there is 0 chance of confusion
  save_season_avg <- seasonal_avgs %>% 
    select(
      year = yr,
      months = season,
      temp_c = mean_temp,
      temp_f = temp_f,
      climate_avg_c = mean_clim,
      climate_avg_f = clim_f,
      temp_anom_c = mean_anom,
      temp_anom_f = anom_f) %>% 
    mutate(season = params$season, .before = "months") %>% 
    mutate(across(where(is.numeric), round,2))
   
  
  # Save locally
  update_name <- str_c(here("R/markdown_reports/seasonal_update_data/"),
                       "Gulf-of-Maine_SST_",
                       params$season,
                       "_Averages_GMRI_",
                       str_remove_all(Sys.Date(), "[-]"),
                       ".csv")
  write_csv(save_season_avg, 
            file = update_name)
}
```

> **Note About the Data:** The figures in this report are created using remotely-sensed satellite data as part of publicly funded research efforts. Satellite SST data was obtained from NOAA's National Center for Environmental Information (NCEI), with all maps and figures displaying [NOAA's Optimum Interpolation Sea Surface Temperature Data](https://www.ncei.noaa.gov/products/optimum-interpolation-sst).

```{r}
#| results: asis
insert_gmri_footer()
```
