---
title: "2021: Gulf of Maine's Warmest Year on Record"
author: 
  - first_name: "Adam"
    last_name: "Kemberling"
    url: https://github.com/adamkemberling
    affiliation: Gulf of Maine Research Institute
    affiliation_url: https://www.gmri.org
description: | 
  Characteristics of a record setting year for ocean temperatures
date: "Updated on: `r Sys.Date()`"
editor_options: 
  chunk_output_type: console
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = FALSE, 
                      message = F, 
                      warning = F, 
                      comment = "", 
                      fig.align = "center", 
                      fig.height = 6, 
                      fig.width = 8)

####  Packages  ####
library(lubridate)
library(raster)
library(here)
library(rnaturalearth)
library(scales)
library(sf)
library(stars)
library(gmRi)
library(heatwaveR)
library(gt)
library(gtExtras)
library(patchwork)
library(tidyverse)
library(ggthemes)
library(ggstream)
library(ggforce)
library(geomtextpath)
library(ggHoriPlot)
#library(directlabels)
library(tidytext)

# Support Functions
suppressPackageStartupMessages( source(here("R/oisst_support_funs.R"), verbose = FALSE) )
suppressPackageStartupMessages( source(here("R/temp_report_support.R"), verbose = FALSE) )

#box paths
oisst_path <- cs_path("res", "OISST/oisst_mainstays")


# File paths for various extents based on "apershing_gulf_of_maine"
region_paths <- get_timeseries_paths(region_group = "gmri_sst_focal_areas", 
                                     mac_os = "mojave")

# Timeseries Path
timeseries_path <- region_paths[["apershing_gulf_of_maine"]][["timeseries_path"]]

# Polygon Path
poly_path <- region_paths[["apershing_gulf_of_maine"]][["shape_path"]]
region_extent <- read_sf(poly_path)

# Adding Logo to plot
logo_path <- paste0(system.file("stylesheets", package = "gmRi"), "/gmri_logo.png")
lab_logo <- magick::image_read(logo_path)


# Coastline Shapes
new_england <- ne_states("united states of america", returnclass = "sf")
canada      <- ne_states("canada", returnclass = "sf")
world_sf       <- ne_countries(returnclass = "sf")
greenland   <- ne_states(country = "greenland", returnclass = "sf")


# Masking Function to clip to the study area
mask_nc <- function(ras_obj, mask_shape){
  
  # Get stats from ranks
  m1 <- mask(rotate(ras_obj), mask_shape)
  m1 <- crop(m1, mask_shape)
  return(m1)
}



# Box Location to save hi-res images
save_location <- cs_path("root", "GoM Seasonal Climate Update/annual_update_figures")
```

`r gmRi::use_gmri_style_rmd(css = "gmri_rmarkdown.css")`

```{r gmri themes, echo = FALSE}

# Set ggplot theme for figures
theme_set(theme_bw() + 
            theme(
              # Titles
              plot.title = element_text(hjust = 0, face = "bold", size = 14),
              plot.subtitle = element_text(size = 9),
              plot.caption = element_text(size = 7.2, margin = margin(t = 20), color = "gray40"),
              legend.title = element_text(size = 9),
              legend.text = element_text(size = 7.5),
              # Axes
              axis.line.y = element_line(color = "black"),
              axis.ticks.y = element_line(), 
              axis.line.x = element_line(color = "black"),
              axis.ticks.x = element_line(), 
              axis.text = element_text(size = 11),
              rect = element_rect(fill = "transparent", color = "black"),
              # Facets
              strip.text = element_text(color = "white", 
                                        face = "bold",
                                        size = 11),
              strip.background = element_rect(
                color = "#00736D", 
                fill = "#00736D", 
                size = 1, 
                linetype="solid"))
          )



# Building a GMRI theme based on Wall street Journal and NYTimes theme
# base settings from {ggthemes}
theme_gmri <- function(base_size = 10, 
                       bg_color = "lightblue", 
                       base_family = "sans", 
                       title_family = "sans",
                       facet_color = "teal") {
    # Color from gmRi palette, sets background color
    #colorhex <- gmRi::gmri_cols()[bg_color]
    facet_hex <- gmri_cols()[facet_color]
    
    # Set up theme
    theme_foundation(
      base_size = base_size, 
      base_family = base_family) + 
        theme(
          
          # Major Elements
          line = element_line(linetype = 1, colour = "black"), 
          rect = element_rect(fill = "transparent", 
                              linetype = 0, 
                              colour = NA), 
          text = element_text(colour = "black"), 
          title = element_text(family = title_family, size = 12), 
          
          # Axis elements
          axis.text.x = element_text(colour = NULL), 
          axis.text.y = element_text(colour = NULL), 
          axis.ticks = element_line(colour = NULL), 
          axis.ticks.y = element_blank(), 
          axis.ticks.x = element_line(colour = NULL), 
          axis.line = element_line(), 
          axis.line.y = element_blank(), 
          axis.text = element_text(size = 11),
          
          # Legend Elements
          legend.background = element_rect(), 
          legend.position = "top", 
          legend.direction = "horizontal", 
          legend.box = "vertical", 
          legend.title = element_text(size = 9),
          legend.text = element_text(size = 7.5),
          
          # Panel/Grid Setup
          panel.grid = element_line(colour = NULL, linetype = 3, color = "gray70"), 
          panel.grid.major = element_line(colour = "black"), 
          panel.grid.major.x = element_blank(), 
          panel.grid.minor = element_blank(), 
          
          # Title and Caption Details
          plot.title = element_text(hjust = 0, face = "bold", size = 14),
          plot.subtitle = element_text(size = 9),
          plot.caption = element_text(size = 7.2, margin = margin(t = 20), color = "gray40"),
          plot.margin = unit(c(1, 1, 1, 1), "lines"), 
          
          # Facet Details
          strip.text = element_text(color = "white", face = "bold", size = 11),
          strip.background = element_rect(
            color = "white", 
            fill = facet_hex, 
            size = 1, 
            linetype="solid"))
}



# Set theme up for maps
map_theme <- list(
  theme(
    panel.border       = element_rect(color = "black", fill = NA),
    plot.background    = element_rect(color = "transparent", fill = "transparent"),
    line               = element_blank(),
    axis.title.x       = element_blank(), # turn off titles
    axis.title.y       = element_blank(),
    legend.title.align = 0.5))
```


# 2021 Sea Surface Temperature

##### An Exceptional Year

2021 was an exceptionally warm year for the Gulf of Maine. It had the warmest fall on record for the area, and the second warmest summer. Much of the year the region experienced what would be considered "heatwave conditions". After compiling sea surface temperature data for the entire year we can now conclude that 2021 was the hottest year on Record for the Gulf of Maine.

```{r load timeseries}
# Load the timeseries
region_timeseries <- read_csv(timeseries_path, col_types = cols(), guess_max = 1e6)


# Clean up the data - add labels
region_timeseries <- region_timeseries %>% 
  distinct(time, .keep_all = T) %>% 
  mutate(time = as.Date(time),
         yr = year(time),
         month_num = month(time),
         month = month(time, label = T, abbr = T),
         week = lubridate::week(time),
         doy = yday(time),
         season = metR::season(month_num, lang = "en"),
         #Set up correct year for winters, they carry across into next year
         season_eng = case_when(
           season == "SON" ~ "Fall",
           season == "DJF" ~ "Winter",
           season == "MAM" ~ "Spring",
           season == "JJA" ~ "Summer"),
         season_yr = ifelse( (season_eng == "Winter" & month_num %in% c(1,2)), yr - 1, yr)
         )

# Get heatwave statuses for each day:
# Uses area weighted sst by default
region_hw <- pull_heatwave_events(
  temperature_timeseries = region_timeseries, 
  threshold = 90, 
  clim_ref_period = c("1982-01-01", "2011-12-31"))

# Format dates and seasons and heatwave outcomes
region_hw <- region_hw %>% 
  mutate(sst_f = as_fahrenheit(sst, "temperature"),
         seas_f = as_fahrenheit(seas, "temperature"),
         anom_f = as_fahrenheit(sst_anom, "anomalies"),
         hwe_f = as_fahrenheit(hwe, "temperature"),
         mhw_thresh_f = as_fahrenheit(mhw_thresh, "temperature"),
         time = as.Date(time),
         yr = year(time),
         month_num = month(time),
         month = month(time, label = T, abbr = T),
         week = lubridate::week(time),
         doy = yday(time),
         season = metR::season(month_num, lang = "en"),
         season_eng = case_when(
           season == "SON" ~ "Fall",
           season == "DJF" ~ "Winter",
           season == "MAM" ~ "Spring",
           season == "JJA" ~ "Summer"),
         #Set up correct year for winters, they carry across into next year
         season_yr = ifelse(
           (season_eng == "Winter" & month_num %in% c(1,2)), yr - 1, yr)
  )

```




## Departure from Historic Conditions

Thanks to some long-running temperature monitoring programs we are able to track monthly sea surface temperatures as far back as 1850. From this view we can see that recent temperatures are the warmest we've ever seen. 2021 setting the new record for the warmest temperature for the region.




```{r}
#| eval = TRUE,
#| fig.alt = "Long term temperature record for the Gulf of Maine"


# Updated 2022-02-4
ersst_path <- cs_path("res", "ERSSTv5")

# The raster stack
# ersst_temp <- stack(str_c(ersst_path, "sst.mnmean_2022_02_02.nc")) #netcdf

# Timeseries for GOM
ersst_tl <- read_csv(str_c(ersst_path, "ERSSTv5_anom_apershing_gulf_of_maine.csv"),
                     guess_max = 1e6, col_types = cols()) 

# format
ersst_tl <- ersst_tl %>% 
  rename(Date = time) %>% 
  mutate(yr = year(Date),
         sst_clim = sst - sst_anom,
         area_wtd_clim = area_wtd_sst - area_wtd_sst_anom)


# Get Yearly Average
ersst_yr <- ersst_tl %>% 
  group_by(yr) %>% 
  summarise(sst = mean(sst, na.rm = T), 
            sst_anom = mean(sst_anom, na.rm = T),
            sst_f = as_fahrenheit(sst, "temperature"),
            anom_f = as_fahrenheit(sst_anom, "anomalies"),
            .groups = "drop") %>% 
  mutate(yr = as.numeric(yr)) %>% 
  filter(yr <= 2021) 

# Benchmark the averages temp and maximum temps
long_avg <- mean( filter(ersst_yr, between(yr, 1860, 2000)) %>% pull(sst) , na.rm = T)
long_max <- max( filter(ersst_yr, between(yr, 1860, 2000)) %>% pull(sst) , na.rm = T)

# Make Splines
ersst_smooth <- as.data.frame(spline(ersst_yr$yr, ersst_yr$sst))
ersst_smooth_f <- as.data.frame(spline(ersst_yr$yr, ersst_yr$sst_f))


# Fahrenheit Plot
ersst_long_term_p <- ggplot(data = ersst_yr, aes(yr, sst_f)) +
  geom_line(group = 1, alpha = 0.4, linetype = 1, color = gmri_cols("gmri blue")) +
  geom_point(size = 1, color = gmri_cols("gmri blue")) +
  geom_point(data = filter(ersst_yr, sst > long_max), aes(yr, sst_f), color = "darkred", size = 1.5) +
  geom_mark_ellipse(aes(yr, sst_f, filter = sst > long_max,
                        description = "Temperatures in 6 of the 10 Last Years Above 1951 Record",
                        label = "Recent Extremes"), label.fontsize = 9,
                    color = "darkred", label.fill = "white", label.colour = "black", con.colour = "black") +
  geom_mark_ellipse(aes(yr, sst_f, filter = sst == long_max, 
                        label = str_c(yr,": ", round(sst_f, 2), " \u00b0F"), 
                        description = "Former Hottest Temperature on Record"), 
                    label.fontsize = 9,  label.buffer = unit(0.25, "lines"),
                    color = "darkred", linetype = 2) +
  scale_color_gmri() +
  theme_gmri() +
  scale_x_continuous(expand = expansion(add = c(2,10))) +
  scale_y_continuous(expand = expansion(add = c(0.5,5)),
    labels = number_format(suffix = " \u00b0F")) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5))  +
  theme(legend.position = "bottom") +
  labs(x = "", y = "Sea Surface Temperature", 
       color = "Temperature Data Record",
       title = "Long-Term Temperature Record for the Gulf of Maine",
       subtitle = "Current temperatures above 150-year highs",
       caption = "Data Source: ERSSTv5 Monthly Sea Surface Temperature")



ersst_long_term_p
```

## Newer Technology: Higher-Quality Data

In 1981 the availability of a higher resolution temperature resource became available in the form of the Optimum Interpolation Sea Surface Temperature (OISSTv2) Satellite Data Record. Once the OISSTv2 data became available we gained access to a higher resolution record, and to the ability to validate past records through comparison. These two data sources now overlap for a period of 42 years and we can see visually that they are not far off:


```{r oisst rates}

# Get yearly averages for OISST
oisst_yr <- region_hw %>% 
  filter(between(yr, 1982, 2021)) %>% 
  group_by(yr) %>% 
  summarise(sst = mean(sst), 
            sst_anom = mean(sst_anom, na.rm = T),
            sst_f = as_fahrenheit(sst, "temperature"),
            anom_f = as_fahrenheit(sst_anom, "anomalies"),
            .groups = "drop")

# Average Distance between OISST and ERSST in GOM
temp_diffs <- ersst_yr %>% 
  rename(ersst_sst = sst,
         ersst_sst_f = sst_f,
         ersst_anom = sst_anom,
         ersst_anom_f = anom_f) %>% 
  inner_join(oisst_yr, by = "yr") %>% 
  mutate(temp_diff = ersst_sst - sst,
         temp_diff_f = ersst_sst_f -sst_f)

# Get temperature differences
temp_difference <- temp_diffs %>% 
  pull(temp_diff) %>% mean() %>% round(2)
temp_difference_f <- temp_diffs %>% 
  pull(temp_diff_f) %>% mean() %>% round(2)


# Make Splines
oisst_smooth <- as.data.frame(spline(oisst_yr$yr, oisst_yr$sst))
oisst_smooth_f <- as.data.frame(spline(oisst_yr$yr, oisst_yr$sst_f))


# Degrees F
ersst_v_oisst <- ggplot() +
  geom_line(data = ersst_smooth_f, aes(x, y, color = "ERSSTv5"),
            group = 1, alpha = 0.4, linetype = 1) +
  geom_point(data = ersst_yr, aes(yr, sst_f, color = "ERSSTv5"), 
             size = 1) +
 geom_line(data = oisst_smooth_f, aes(x, y, color = "OISSTv2"),
            group = 1, alpha = 0.4, linetype = 1) +
  geom_point(data = oisst_yr, aes(yr, sst_f, color = "OISSTv2"), 
             size = 1) +
  scale_color_gmri() +
  theme_gmri() +
   scale_y_continuous(
     #expand = expansion(add = c(0.5,5)),
     labels = number_format(suffix = " \u00b0F")) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5)) +
  facet_zoom(xlim = c(1982, 2021)) +
  theme(legend.position = "bottom")  +
  labs(x = "", y = "Sea Surface Temperature", 
       color = "Temperature Data Source",
       title = "Long-Term Temperature Record for the Gulf of Maine",
       subtitle = "OISSTv2 Data Availability Overlap with ERSSTv5")



ersst_v_oisst
```



Because the ERSST data is of a coarser resolution (monthly measurements at a 0.5 x 0.5 degree grid) and does not capture the warmer inshore dynamics (a bias to show colder temperature). The OISSTv2 temperatures in this area are on average warmer. **Across the years where data from both sources is available, ERSSTv5 is on average `r temp_difference`$^oC$ or `r temp_difference_f`$^oF$ Colder than OISSTv2 data.** 


## Warming Relative to the World

While this report focuses on the Gulf of Maine, there is a recognition that this region is not unique in the fact that its temperatures are rising. The Earth as a whole is warming, and that change is being driven by anthropogenic influences to our atmosphere. The impacts of these changes however are not distributed evenly, and there are regions like ours that are warming at rates above the global average.


```{r yearly summaries}
# Summarize by year to return mean annual anomalies and variance
annual_summary <- region_timeseries %>% 
  mutate(year = year(time)) %>% 
  filter(year %in% c(1982:2021)) %>% 
  group_by(year) %>% 
  summarise(sst = mean(sst, na.rm = T),
            sst_anom = mean(sst_anom, na.rm = T), 
            sst_f = as_fahrenheit(sst, "temperature"),
            anom_f = as_fahrenheit(sst_anom, "anomalies"),
            area_wtd_sst = mean(area_wtd_sst, na.rm = T),
            area_wtd_anom = mean(area_wtd_anom, na.rm = T),
            area_wtd_f = as_fahrenheit(area_wtd_sst, "temperature"),
            area_wtd_anom_f = as_fahrenheit(area_wtd_anom, "anomalies"),
            .groups = "drop") %>% 
  mutate(yr_as_dtime = as.Date(paste0(year, "-07-02")))


# # Global Temperature Anomaly Rates
global_anoms <- read_csv(
    paste0(oisst_path, "global_timeseries/global_anoms_1982to2011.csv"), 
    guess_max = 1e6,
    col_types = cols()) %>% 
  mutate(year = year(time))

# summarize by year again
global_summary <- global_anoms %>% 
  group_by(year) %>% 
  summarise(sst = mean(sst, na.rm = T), 
            sst_anom = mean(sst_anom, na.rm = T),
            sst_f = as_fahrenheit(sst, "temperature"),
            anom_f = as_fahrenheit(sst_anom, "anomalies"),
            area_wtd_sst = mean(area_wtd_sst, na.rm = T),
            area_wtd_anom = mean(area_wtd_anom, na.rm = T),
            area_wtd_f = as_fahrenheit(area_wtd_sst, "temperature"),
            area_wtd_anom_f = as_fahrenheit(area_wtd_anom, "anomalies"),
            .groups = "drop") %>% 
  mutate(yr_as_dtime = as.Date(paste0(year, "-07-02")))





# Build Regression Equation Labels

# This should all be a function:
decadal_rates <- function(temp_df, temp_col, year_lim = c(19823:2021)){}
# End function

  
  
  
# 1. All years 
lm_all <- lm(area_wtd_f ~ year, 
             data = filter(annual_summary, year %in% c(1982:2021))) %>% 
  coef() %>% 
  round(3)

# 2. Last 15 years
lm_15  <- lm(area_wtd_f ~ year, 
             data = filter(annual_summary, year %in% c(2007:2021))) %>% 
  coef() %>% 
  round(3)

# 3. Global - Area-weighted
lm_global <- lm(area_wtd_f ~ year, 
                data = filter(global_summary, year %in% c(1982:2021))) %>% 
  coef() %>% 
  round(3)

# Convert yearly rate to decadal
decade_all    <- lm_all['year'] * 10
decade_15     <- lm_15['year'] * 10
decade_global <- lm_global["year"] * 10

# Fahrenheit equations
eq_all    <- paste0("GoM 1982-2021: ", decade_all,    "\u00b0", "F / Decade")
eq_15     <- paste0("GoM 2007-2021: ", decade_15,     "\u00b0", "F / Decade")
eq_global <- paste0("World 1982-2021: ", decade_global, "\u00b0", "F / Decade")


# Generate a smoothed temperature line using splines
yearly_temp_smooth <-  as.data.frame(spline(annual_summary$yr_as_dtime, annual_summary$area_wtd_f)) %>% 
  mutate(x = as.Date(x, origin = "1970-01-01"))


# Line color formatting
line_colors <- setNames(
  as.character(c(
    gmri_cols("gmri blue"), 
    #gmri_cols("orange"),
    gmri_cols("green"))),  
  c(eq_all, 
    #eq_15, 
    eq_global))



# Single Line Plot
(temp_simplified <- ggplot(data = annual_summary, aes(year, area_wtd_anom_f)) +

  # Overlay yearly means
  geom_line(color = "gray10", size = 0.75, linetype = 1) +
  geom_point(color = "gray10", size = 1) +
  geom_smooth(
     data = mutate(global_summary, `Warming Rate` = eq_global),
     aes(color = `Warming Rate`),
    method = "lm",
    formula = y ~ x, se = F,
    alpha = 0.90,
    size = 1,
    linetype = 1) +
  geom_smooth(
    data = annual_summary %>% 
      mutate(`Warming Rate` = eq_all),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    color = gmri_cols("gmri blue"),
    alpha = 0.90,
    size = 1,
    linetype = 1) + 
   scale_color_manual(values = line_colors) +
   scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
   labs(title = "Gulf of Maine:",
        subtitle = "Annual Sea Surface Temperature Anomalies",
        x = "Year", y = "Sea surface Temperature Anomaly",
        caption = "Anomalies calculated using 1982-2011 reference period.") +
   theme(legend.position = c(0.825, 0.125))

)


```

## Warming at an Accelerated Pace

In recent years we've seen temperatures in the Gulf of Maine (and adjacent waters) to accelerate rapidly. Scientists believe this is partly due to changes in the behavior of the Gulf Stream Current which brings warm, salty water from the Gulf of Mexico & Florida Northward along the east coast of the US and eventually flowing Northeast past Nova Scotia.

Scientists believe a regime shift has occurred in the Northwest Atlantic as a result of a Northward shift in the Gulf Stream, and a general weakening in the Atlantic Meridional Overturning Circulation Pattern (AMOC). This Northward shift in the Gulf stream does two things: 1) It brings more warm/salty water into the Gulf of Maine, & 2) It obstructs the flow of cold and relatively fresh water from entering the Gulf of Maine from the Labrador current and Scotian Shelf Water (SSW).

These relative contribution of these two impacts may vary through time, but the result (in surface temperature terms) is similar: **a warmer Gulf of Maine.**

```{r gom full warming, eval = T}

# Line color formatting
line_colors <- setNames(
  as.character(c(
    gmri_cols("gmri blue"), 
    gmri_cols("orange"))),  
  c(eq_all, 
    eq_15))


# Gulf of Maine Warming
gom_all_trend <- annual_summary %>%
  ggplot(aes(year, area_wtd_anom_f)) +
  geom_line(color = "gray10", size = 0.75, linetype = 1) +
  geom_point(color = "gray10", size = 1) +
  # Add regression lines
  geom_smooth(
    data = annual_summary %>%
      mutate(`Warming Rate` = eq_all),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    alpha = 0.90,
    size = 1,
    linetype = 1) +
  # Axes
  scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
  scale_color_manual(values = line_colors) +

  # labels + theme
  labs(title = "Gulf of Maine: Long-Term Warming Trend",
       x = "",
       y = "SST Anomaly",
       caption = "")


# Gulf of Maine Warming - last 15 years
gom_recent_trend <- annual_summary %>% 
  ggplot(aes(year, area_wtd_anom_f)) +
  geom_line(color = "gray10", size = 0.75, linetype = 1) +
  geom_point(color = "gray10", size = 1) +
  # Add regression lines
  geom_smooth(
    data = filter(annual_summary, year %in% c(2007:2021)) %>% 
      mutate(`Warming Rate` = eq_15),
    method = "lm",
    formula = y ~ x, se = F,
    aes(color = `Warming Rate`),
    alpha = 0.90,
    size = 1,
    linetype = 1) +
  
  # Axes
  scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
  scale_color_manual(values = line_colors) +
  
  # labels + theme
  labs(title = "Gulf of Maine: Rapid Recent Warming",
       x = "", 
       y = "SST Anomaly",
       caption = paste0("Anomalies calculated using 1982-2011 reference period."))




# Put them together
(gom_recent_warming <- (gom_all_trend / gom_recent_trend ) + 
  plot_layout(guides = "collect")  & 
  theme(legend.position = "right") &
  guides(color = guide_legend(title.position = "top", ncol = 1, title.hjust = 0.5)))
```


## Concerning Temperatures in Recent Years




```{r}
####  Annual Trend Plot  ####
gom_anomaly_bars <- annual_summary %>% 
  mutate(anom_direction = ifelse(area_wtd_anom > 0, "Hot", "Cold")) %>% 
  ggplot(aes(yr_as_dtime, area_wtd_anom_f)) +
  
  # Underlying anomaly data
  
  # Yearly Bars
  geom_col(aes(fill = anom_direction), color = "black", size = 0.25, alpha = 0.9) +
  scale_fill_manual(values = c("Hot" = "#CA0020", "Cold" = "#0571B0")) +
  # Axes
   scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +

  # theme
  theme_gmri() +
  theme(#legend.title = element_blank(),
        legend.position = c(0.85, .125),
        legend.background = element_rect(fill = "white"), 
        legend.key = element_rect(fill = "transparent", color = "transparent")) +
   # labels
  labs(title = "Gulf of Maine: Rapid Recent Warming",
       subtitle = "Recent Rate More Than Double the 40-Year Average",
       fill = "Anomaly Type",
       x = "", 
       y = "Sea Surface Temperature Anomaly",
       caption = paste0("Anomalies calculated using 1982-2011 reference period."))



gom_anomaly_bars

```




## 2021: Daily Temperatures and Heatwave Status

```{r}
 # Last Calendar Year
this_yr <- region_hw %>% 
  filter(year(time) == 2021)

#min/max of climate and min/max of 2021
clim_range <- round(range(this_yr$seas), 2)
clim_diff <- round(diff(clim_range), 2)

temp_range <- round(range(this_yr$sst), 2)
temp_diff <- round(diff(temp_range), 2)

anom_range <- round(range(this_yr$sst_anom), 2)
anom_diff <- round(diff(anom_range), 2)
```


The Gulf of Maine has a seasonal temperature cycle that experiences its lowest temperatures around March, and its peak summer highs in August. The difference between these two seasons is typically `r round(as_fahrenheit(clim_diff, "anomalies"), 2)`$^oF$ or `r clim_diff`$^oC$. In 2021 the temperature range was `r temp_range[1]`-`r temp_range[2]`$^oC$ or `r temp_diff`$^oC$, with temperature anomaly values ranging from `r anom_range[1]`-`r anom_range[2]`$^oC$.

The highest temperatures anomalies spiked in the last days of June, capping off an already warm start to the year. A very-warm August and steady, above-average temperatures from early September through the end of the year helped solidify 2021 as the hottest year on record for the Gulf of Maine.

```{r single year lineplot}
#| fig.alt = "Timelines of observed temperature and observed temperature anomalies for the year of 2021, and how they compare to the mean and 90th percentile of the climate reference period"


# Set colors by name
color_vals <- c(
  "Sea Surface Temperature" = "#0571B0",
  "Heatwave Event"          = "darkred",
  "MHW Threshold"           = "coral3",
  "Daily Climatology"       = "gray30")


# Set the label with degree symbol
ylab <- "Sea Surface Temperature"



# Plot the last 365 days
hw_temp_p <- ggplot(this_yr, aes(x = time)) +
    geom_segment(aes(x = time, xend = time, 
                     y = seas_f, yend = sst_f), 
                 color = "#0571B0", alpha = 0.25) +
    geom_segment(aes(x = time, xend = time, 
                     y = mhw_thresh_f, yend = hwe_f), 
                 color = "darkred", alpha = 0.25) +
    geom_line(aes(y = sst_f, color = "Sea Surface Temperature")) +
    geom_line(aes(y = hwe_f, color = "Heatwave Event")) +
    geom_line(aes(y = mhw_thresh_f, color = "MHW Threshold"), lty = 3, size = .5) +
    geom_textpath(aes(y = seas_f,  color = "Daily Climatology"), label = "Climatology Mean", hjust = 0.5, lty = 2) +
    scale_color_manual(values = color_vals) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(mult = c(0,0))) +
    scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
    theme(legend.title = element_blank(),
          legend.position = "none") +
    labs(x = "", 
         y = ylab)



# Plot the last 365 days - anomaly scale
linetype_key <- c(
  "Sea Surface Temperature Anomaly" = 1,
  "Heatwave Event"                  = 1,
  "MHW Threshold"                   = 3,
  "Daily Climatology"               = 2)

# Same Plot as Anomalies
hw_anom_p <- this_yr %>% 
  mutate(
    anom_thresh = mhw_thresh - seas,
    anom_thresh_f = mhw_thresh_f - seas_f,
    anom_hwe = hwe - seas,
    anom_hwe_f = hwe_f - seas_f) %>% 
ggplot(aes(x = time)) +
    geom_segment(aes(x = time, xend = time, 
                     y = 0, yend = anom_f), 
                 color = "#0571B0", alpha = 0.25) +
    geom_segment(aes(x = time, xend = time, 
                       y = anom_thresh_f, yend = anom_hwe_f), 
                 color = "darkred", alpha = 0.25) +
    geom_line(aes(y = anom_f, color = "Sea Surface Temperature Anomaly")) +
    geom_line(aes(y = anom_hwe_f, color = "Heatwave Event")) +
    geom_line(aes(y = anom_thresh_f, color = "MHW Threshold"), lty = 3, size = .5) +
    geom_line(aes(y = 0, color = "Daily Climatology"), lty = 2, size = .5) +
   scale_color_manual(values = color_vals) +
    scale_linetype_manual(values = linetype_key, guide = "none") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(mult = c(0,0))) +
    guides(color = guide_legend(override.aes = list(linetype = linetype_key), nrow = 2)) +
    theme(legend.title = element_blank(),
          legend.position = "top") +
    scale_y_continuous(labels =  number_format(suffix = " \u00b0F")) +
    labs(x = "", 
         y = "Temperature Anomaly", 
         caption = paste0("Climate reference period :  1982-2011"))


# Show figure
(year_timeline <- hw_temp_p / hw_anom_p)

```



### 2021: Heatwave Events

```{r}

# Get the Event numbers
events_21 <-  this_yr %>% 
  drop_na(mhw_event_no) %>% 
  distinct(mhw_event_no) %>% 
  pull(mhw_event_no)

events_num <-  length(events_21)


# Pull their data, including the days before and after
events_dat <- filter(region_hw, mhw_event_no %in% as.character(events_21),
                     yr == 2021)
# Days in HW state
days_hw <- nrow(events_dat)

# How long on average
avg_duration <- round(days_hw / events_num, digits = 0)
```

During 2021 there were `r events_num` marine heatwave events,  lasting a total of `r days_hw`. These events include one that carried over from the previous year and another that has persisted into 2022. The average duration for these events was `r avg_duration` days (excluding days in 2020 & 2022).


### 2021: Record Daily Temperatures

To add context on the severity of these sustained heatwaves, many of the individual daily temperatures throughout 2021 were themselves the hottest days on record.

```{r}
# Grab the hottest day for each day of the year
hottest_days <- region_hw %>% 
  group_by(doy) %>% 
  slice_max(order_by = sst_anom, n = 1) %>% 
  ungroup() 

# Number in 2021
hottest_2021 <- hottest_days %>% 
  count(yr) %>% 
  filter(yr == "2021") %>% 
  pull(n)

```

Of the 365 days in a year, 2021 set a new record temperature on `r hottest_2021` or `r round((hottest_2021/365) * 100, 2)`% of the year.


## Long-Term Heatwave Seasonality

When looking at the last ~40 years of marine heatwave events here in the Gulf of Maine. We see that the frequency, duration, and intensity of heatwave events has increased. This year was no exception with `r days_hw` days meeting the criteria for heatwave status.


```{r}
#| fig.alt = "Heatmap of heatwave temperatures and heatwave days"

# Set up the date within a year
base_date <- as.Date("2000-01-01")
grid_data <- region_hw %>% 
  mutate(year = year(time),
         yday = yday(time),
         flat_date = as.Date(yday-1, origin = base_date))

# Prep the legend title
guide_lab <- "Sea Surface Temperature Anomaly \u00b0C"

# Color limit for palettes
temp_limits <- c(-5, 5)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0F")


# Assemble heatmap plot
heatwave_heatmap <- ggplot(grid_data, aes(x = flat_date, y = year)) +
  
  # background box fill for missing dates
  geom_rect(xmin = base_date, xmax = base_date + 365, 
            ymin = min(grid_data$year) - .5, ymax = max(grid_data$year) + .5, 
            fill = "gray75", color = "transparent") +
  
  # tile for sst colors
  geom_tile(aes(fill = anom_f)) +
  # points for heatwave events
  geom_point(data = filter(grid_data, mhw_event == TRUE),
             aes(x = flat_date, y = year), size = .25)  +
  scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = c(0,0)) +
  scale_y_continuous(limits = c(1980.5, 2021.5), expand = c(0,0)) +
  scale_fill_distiller(palette = "RdBu", 
                     na.value = "transparent", 
                     limit = temp_limits, 
                     oob = scales::squish,
                     breaks = temp_breaks, 
                     labels = temp_labels) +
  
  #5 inches is default rmarkdown height for barheight
  guides("fill" = guide_colorbar(title = guide_lab, 
                                 title.position = "right", 
                                 title.hjust = 0.5,
                                 barheight = unit(4.8, "inches"), 
                                 frame.colour = "black", 
                                 ticks.colour = "black")) +  
  theme(legend.title = element_text(angle = 90)) +
  labs(title = "Temperature Regime Shift",
       subtitle = "Marine Heatwave Events Lasting Months Now Frequent",
       x = "", 
       y = "",
       "\nClimate reference period : 1982-2011",
       caption = "Marine heatwave is an event hotter than 90th percentile of 1982-2011 temperatures on that day.")


# Assemble pieces
heatwave_heatmap
grid::grid.raster(lab_logo, x = 0.05, y = 0.03, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```




##  Temperature Anomaly Horizons

One way to think about the severity of these changes is to think about temperature horizons. A temperature horizon captures how long temperatures remain above certain thresholds. Each threshold is designated its own temperature, and in this way we can see how long within a year temperatures remained: 1 to 2 or as much as 4$^oC$ above normal.  

```{r}
#| fig.alt = "Horizon plot of all years and their temperature anomaly horizons"

# make slight modification to grid_data
hori_data <- grid_data %>% 
  mutate(year = factor(year),
         year = fct_rev(year))


# cut out outliers and set cutpoints
cutpoints <- hori_data %>% 
  mutate(
    outlier = between(
      sst_anom,
      quantile(sst_anom, 0.25, na.rm=T)-
        1.5*IQR(sst_anom, na.rm=T),
      quantile(sst_anom, 0.75, na.rm=T)+
        1.5*IQR(sst_anom, na.rm=T))) %>% 
  filter(outlier)

# Set origin
ori <- 0

# Manually pick cut points
sca <- seq(-4, 4, length.out = 9)

sca_bins <- str_c(sca[1:(length(sca)-1)], " to ", sca[2:length(sca)], "\u00b0C")
sca_labels <- rev(sca_bins)


# Function to plot one or more years
anom_horizon_plot <- function(hori_data, origin = ori, scale_cutpoints = sca, labels = sca_labels){
  
   ggplot(hori_data) +
  geom_horizon(aes(flat_date, 
                   sst_anom,
                   fill = ..Cutpoints..), 
               origin = ori, 
               horizonscale = sca) +
  scale_fill_hcl(palette = 'RdBu', reverse = T, labels = sca_labels) +
  facet_grid(year~.) +
  theme(
    #strip.background = element_rect(fill = "#36454F", color = "#36454F"),
    panel.grid = element_blank(),
    panel.spacing.y=unit(0.1, "lines"),
    strip.text.y = element_text(size = 7, angle = 0, hjust = 0),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.border = element_blank(),
    legend.position = "left") +
  scale_x_date(expand=c(0,0), 
               date_breaks = "1 month", 
               date_labels = "%b") +
  labs(x = '', 
       fill = "Temperature Anomaly \u00b0C") 
  
  
}


# Plot Horizon Plot Using All Years
gom_annual_horizons <- anom_horizon_plot(hori_data = filter(hori_data, yr <= 2021), 
                  origin = ori, 
                  scale_cutpoints = sca) +
  labs(title = "Temperature Anomaly Horizons",
       subtitle = "How deviation from the norm has become the new-normal",
       caption = "Climatatological Reference Period: 1982-2011")


# Plot it
gom_annual_horizons
grid::grid.raster(lab_logo, x = 0.05, y = 0.03, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```

## Comparing the Gulf's Two Hottest Years

If we pull the horizons of our two hottest years on record it makes it easier to contrast where either one experienced acute high temperature events, and where there were sustained periods of above average temperatures.

Early into 2021 it was apparent that the year was on-par with the previous title-holder for warmest year on record.

*Could have more text here*

```{r}
#| fig.alt = "Single year horizon plots comparing 2012 and 2022"

# Reset the outliers
# cut out outliers and set cutpoints
cutpoints_2 <- hori_data %>% 
  filter(year %in% c("2012", "2021")) %>% 
  mutate(
    outlier = between(
      sst_anom,
      quantile(sst_anom, 0.25, na.rm=T)-
        1.5*IQR(sst_anom, na.rm=T),
      quantile(sst_anom, 0.75, na.rm=T)+
        1.5*IQR(sst_anom, na.rm=T))) %>% 
  filter(outlier)



# 2012
hori_2012 <- hori_data %>% 
  filter(year == "2012") %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  facet_wrap(~year, strip.position = "top")
  
# 2021
hori_2021 <- hori_data %>% filter(year == "2021") %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  facet_wrap(~year, strip.position = "top") +
  theme(plot.caption = element_text(colour = "gray25", size = 6)) +
  labs(caption = "Climatatological Reference Period: 1982-2011")

# Put them together as one figure
comparison_horizons <- hori_2012 / hori_2021 + plot_layout(guides = "collect")

# Do a bunch of formatting
comparison_horizons <-comparison_horizons &
  theme(
    #Legend
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines")) &
   guides(fill = guide_legend(title = "Temperature Anomaly",
                              title.hjust = 0.5, 
                              title.position = "top",
                              label.position = "bottom", 
                              nrow = 1, 
                              byrow = T, reverse = T)) &
  plot_annotation(title = "Comparing the Gulf of Maine's Two Hottest Years", 
                  subtitle = "Duration of temperature anomaly horizons, colored by their strength")




# Plot it
comparison_horizons
grid::grid.raster(lab_logo, x = 0.08, y = 0.04, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
  
```


## Losing the Balance Between Hot and Cold

Another way to visualize the climate transition that we are observing is by looking at the fraction of each year spent in different temperature ranges. Under a steady climate we would expect over the long-term to spend similar amounts of the year experiencing relatively warm & cold temperatures. These periods would balance themselves out and we would on-average have experienced something similar to the long-term climate.

What we have been seeing in the Gulf of Maine recently has lost that sense of balance. Larger fractions of the year are shared by above average temperatures & cold spell events are becoming vanishingly rare.

```{r}
#| fig.alt = "Streamplot tracking fraction of each year spent in varying degrees of temperature anomalies"


# Set Factor Order for Bins
bin_order <- c(
  "Less than -1", 
  "-1 to -0.5",
  "Within 0.5",
  "0.5 to 1", 
  "Greater than 1")

# Set Colors
bin_colors <- RColorBrewer::brewer.pal(n = length(bin_order), 
                                       name = "RdBu")
bin_colors <- rev(bin_colors)

# Count Days in Bins
day_types <- region_hw %>% 
  mutate(
    `Less than -1`   = if_else(sst_anom <= -1, TRUE, FALSE),
    `-1 to -0.5`     = if_else(between(sst_anom, -1, -0.5), TRUE, FALSE),
    `Within 0.5`     = if_else(between(sst_anom, -0.5, 0.5), TRUE, FALSE),
    `0.5 to 1`       = if_else(between(sst_anom, 0.5, 1), TRUE, FALSE),
    `Greater than 1` = if_else(sst_anom >= 1, TRUE, FALSE)) %>% 
  group_by(yr) %>% 
  summarise(
    `Less than -1`   = sum(`Less than -1`, na.rm = T),
     `-1 to -0.5`    = sum(`-1 to -0.5`, na.rm = T),
    `Within 0.5`     = sum(`Within 0.5`, na.rm = T),
    `0.5 to 1`       = sum(`0.5 to 1`, na.rm = T),
    `Greater than 1` = sum(`Greater than 1`, na.rm = T),
    .groups = "drop") %>% 
  pivot_longer(cols = "Less than -1":"Greater than 1", names_to = "Anomaly Strength", values_to = "day_total", values_drop_na = FALSE) %>% 
  mutate(day_total = if_else(is.na(day_total), 0L, day_total),
         `Anomaly Strength` = factor(`Anomaly Strength`, levels = bin_order))


# Plot the Streamplot
hw_streamplot <- ggplot(day_types, aes(yr, y = day_total, fill = `Anomaly Strength`)) +
  geom_stream(type = "proportion") +
  scale_x_continuous(breaks = seq(1980,2030, by = 10), 
                     minor_breaks = seq(1975, 2030, by = 5), 
                     expand = c(.03,0)) +
  scale_y_continuous(labels = scales::percent_format(), expand = c(0,0)) + 
  scale_fill_manual(values = setNames(bin_colors, bin_order)) + 
  labs(y = "", x = "", fill = "", 
       title = "A Warmer Gulf of Maine",
       subtitle = "Fractions of Each Year Experienced as Hot/Cold Anomalies",
       caption = "1981 not a complete years in the data*") +
  theme_minimal(base_size = 10) +
  theme(
    axis.line.x = element_line(),
    axis.ticks.x = element_line(),
    axis.line.y = element_line(),
    axis.ticks.y = element_line(),
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines"),
    panel.grid.minor = element_blank(),
    panel.grid = element_line(size = 0.2),
    plot.margin = unit(c(.5,1,.3,.5), "cm"),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 9),
    plot.caption = element_text(size = 7.2, margin = margin(t = 20)),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 7.5),
    legend.margin = margin(c(7,0,0,0)),
    legend.justification = "center") +
   guides(fill = guide_legend(title = "Anomaly Strength \u00b0C",
                             title.hjust = 0.5, 
                             title.position = "left",
                             label.position = "bottom", 
                             nrow = 1, 
                             byrow = T))


# Show figure
hw_streamplot
grid::grid.raster(lab_logo, x = 0.08, y = 0.04, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```


## Characterizing the Hottest Years

There are a number of ways one might try to describe the extent and severity of temperature extremes. Aside from overall average temperatures there may be acute temperature surges that may have a larger impact than a prolonged warm period. To  look more closely at these differences ranked each year on the following characteristics:   
 - Average overall temperature   
 - The average temperature above normal   
 - The cumulative degrees above normal   
 - The average heatwave length   
 
 

```{r}
#| fig.alt = "4-panel plot comparing different features of hot years ranked by different metrics"

# Use all years so they can be ranked independently
year_mets <- hori_data %>% 
  filter(year != "2022") %>% 
  group_by(year) %>% 
  summarise(
    total_days                         = n(),
    `Average Temperature`              = round(mean(sst, na.rm = T), 2),
    `Temperature Above Average`        = round(mean(sst_anom, na.rm = T), 2),
    `Cumulative Degrees Above Average` = round(sum(sst_anom, na.rm = T), 0),
    hw_days                            = sum(mhw_event, na.rm = T),
    hw_events                          = n_distinct(mhw_event_no),
    peak_temp                          = max(sst),
    peak_anom                          = max(sst_anom)) %>% 
  mutate(
    hw_day_pct = round((hw_days/total_days) * 100, 2),
    `Average Heatwave Length` = round(hw_days/hw_events,  1)
  ) %>% 
  pivot_longer(names_to = "Var", values_to = "Metric", cols = `Average Temperature`:`Average Heatwave Length`)


# Snipe out the top5 for each
year_met_ranks <- year_mets %>% 
  filter(Var %in% c("Average Temperature", "Temperature Above Average", "Average Heatwave Length", "Cumulative Degrees Above Average")) %>% 
  group_by(Var) %>% 
  slice_max(n = 5, order_by = Metric) %>% 
  ungroup() %>%
  mutate(
    yr_focus = ifelse(year == "2021", TRUE, FALSE),
    year = reorder_within(year, Metric, Var)) 

# Plot the rankings
hottest_yr_panels <- year_met_ranks %>% 
  ggplot(aes(year, Metric, color = yr_focus)) +
    geom_col(aes(fill = yr_focus), show.legend = F) +
    geom_label(aes(label = Metric),  show.legend = FALSE) +
    facet_wrap(~Var, scales = "free", ncol = 2) +
    coord_flip() +
    scale_x_reordered() +
    scale_y_continuous(expand = expansion(mult = c(0, .2))) +
    labs(y = "",
         x = NULL,
         title = "How 2021 Stacks Up:",
         caption = "Temperature statistics in degrees Celsius",
         subtitle = "Ranking Characteristics of Acute and Prolonged Warming Events") +
  theme(plot.title = element_text(hjust = 0)) + 
  scale_color_gmri() +
  scale_fill_gmri() +
  theme(panel.border = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major. = element_line(linetype = 3, color = "gray50"))


hottest_yr_panels
grid::grid.raster(lab_logo, x = 0.84, y = 0.94, just = c('left', 'bottom'), width = unit(0.8, 'inches'))

```

---

## Mapping Record Temperatures

2021 was an exceptionally warm year for many parts of the World, not just the Gulf of Maine. By taking the average temperature for the year for all years, we can rank how this year compares to previous years in the OISST data. For many places around the world this was one of the top 3 warmest years on record.

```{r}
# Load the rankings done in python, do some checking
ranks_path <- str_c(oisst_path, 'temp_rankings/yrly_ranks_1982to2021.nc')

#  Pull this year
ranks_21 <- stack(ranks_path)[["X2021"]]

# Reclassify Ranks
# Ranks are all integers so we just need to be sure that they are captured in the bounds we want
reclass_df <- c(
  0, 1.1, -1,  # Coldest Year
  1.1, 2.1, -2,  # Second Coldest
  2.1, 3.1, -3,  # Third Coldest
  3.1, 4.1, -4,  # Fourth Coldest
  4.1, 5.1, -5,  # Fifth Coldest
  3.1, 35, NA, # Things to Mask
  35.1, 36, 5, # Fifth Warmest
  36.1, 37, 4, # Fourth Warmest
  37.1, 38, 3, # Third Warmest
  38.1, 39, 2, # Second Warmest
  39.1, 40, 1  # Warmest
)

# make reclassification matrix
reclass_m <- matrix(reclass_df, ncol = 3, byrow = TRUE)

# reclassify - masking middle values handled with reclassification
ranks_reclass <- reclassify(ranks_21, reclass_m)

# Make Stars for ggplot
ranks_st <- st_as_stars(rotate(ranks_reclass))

# Mutate to get levels as factors for legend
ranks_st <- ranks_st %>% 
  mutate(
    temp_rank = case_when(
      X2021 == -1 ~ "Coldest",
      X2021 == -2 ~ "2nd Coldest",
      X2021 == -3 ~ "3rd Coldest",
      X2021 == -4 ~ "4th Coldest",
      X2021 == -5 ~ "5th Coldest",
      X2021 == 5 ~ "5th Warmest",
      X2021 == 4 ~ "4th Warmest",
      X2021 == 3 ~ "3rd Warmest",
      X2021 == 2 ~ "2nd Warmest",
      X2021 == 1 ~ "Warmest",
      TRUE ~ "Not a Record Year"),
    temp_rank = fct_drop(temp_rank),
    temp_rank = factor(temp_rank, levels = c(
      "Warmest", "2nd Warmest", "3rd Warmest", "4th Warmest", "5th Warmest",
      "Not a Record Year",
      "5th Coldest", "4th Coldest", "3rd Coldest", "2nd Coldest", "Coldest"))) %>% 
  select(-X2021) 


# Make map
ranks_map <- ggplot() +
  geom_stars(data = ranks_st) +
  geom_sf(data = world_sf, size = .25) +
  scale_fill_brewer(palette = "RdBu", na.value = "transparent") +
  scale_x_continuous(breaks = seq(-180, 180, 90), 
                     expand = expansion(mult = c(0,0))) +
  scale_y_continuous(breaks = seq(-90, 90, 30), 
                     expand = expansion(mult = c(0,0))) +
  labs(title = "Record Setting Temperatures of 2021",
       fill = "",
       caption = "Temperature rankings cover period of 1982-2021") +
  map_theme +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) 
  
ranks_map
grid::grid.raster(lab_logo, x = 0.08, y = 0.16, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```



```{r, eval = TRUE}
# Zoom in on Northeast US
new_england <- ne_states("united states of america", returnclass = "sf")
canada      <- ne_states("canada", returnclass = "sf")



(northeast_records <- ggplot() +
  geom_stars(data = ranks_st) +
  geom_sf(data = new_england, size = .25) +
  geom_sf(data = canada, size = .25) +
  scale_fill_brewer(palette = "RdBu", na.value = "transparent")+ 
  coord_sf(xlim = c(-80, -55), ylim = c(34, 49)) +
  labs(x = "", y = "", fill = "2021 Temperature Rank",
       title = "Record Warm Temperatures",
       subtitle = "From North Carolina to Newfoundland",
       caption = "Temperature rankings cover period of 1982-2021")  +
  map_theme)

grid::grid.raster(lab_logo, x = 0.08, y = 0.02, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```




## Comparing Regions

We are often asked when speaking to the degree of the Gulf of Maine is warming: 
> "Where is it warming faster?"

This question is challenging because the Gulf of Maine, and any area in the ocean has a unique size/shape/dynamic that makes them difficult to directly compare directly. To put the Gulf of Maine's warming into perspective against similar ecologically and oceanographically relevant units we can compare it against the [large marine ecosystems (LME)](https://www.lmehub.net/) of the world.
 
```{r lme timeseries}
# Load the LME timeseries to calculate rates of warming 
lme_names <- get_region_names("lme") # names
lme_paths <- get_timeseries_paths("lme", mac_os = "mojave") # paths

# Data
lme_oisst <- map(lme_names, ~ read_csv(lme_paths[[.x]][["timeseries_path"]], col_types = cols(), guess_max = 1e6)) 
lme_oisst <- setNames(lme_oisst, lme_names)

# Add Gulf of Maine Timeseries to the List with LME's
lme_oisst[["gulf_of_maine"]] <- region_timeseries

# Make sure dates are formatted
lme_oisst <- map(lme_oisst, ~ mutate(.x, time = as.Date(time)))


####  Warming Rates  ####
lme_8221 <- map(lme_oisst, ~ mutate(.x, year = year(time)) %>% filter(year %in% c(1982:2021)))

# Get warming rates
lme_rates <- map_dfr(lme_8221, function(lme_data){
  # Uses area weighted SST
  lme_yearly <- group_by(lme_data, year) %>% 
    summarise(sst = mean(area_wtd_sst, na.rm = T))
  yrly_warming <- lm(sst ~ year, data = lme_yearly)
  mod_details <- broom::tidy(yrly_warming)
  yrly_rate <- mod_details[2,"estimate"]
  names(yrly_rate) <- "annual_warming_rate_C"
  yrly_rate
}, .id = "Region") %>% 
  arrange(desc(annual_warming_rate_C))


# Pull out the top 13
rates_ordered <- lme_rates %>% 
  mutate(`Warming Rank` = row_number(),
         `Warming Rate F` = as_fahrenheit(annual_warming_rate_C, data_type = "anomalies"),
         Region = str_replace_all(Region, "_", " "),
         Region = str_to_title(Region)) %>% 
  select(Region, `Warming Rank`, `Warming Rate C` = annual_warming_rate_C, `Warming Rate F`) 

# symbol for degree: \u00b0
```


```{r lme rank table}
# Make table of top 10*
lme_rank_table <- rates_ordered %>% 
  slice(1:10) %>% 
  mutate(across(where(is.numeric), round, 3)) %>% 
  gt(rowname_col = "Region") %>% 
   tab_header(
    title = md("**Warming Rates of Large Marine Ecosystems**")) %>%
  tab_stubhead(label = "Region") %>% 
  tab_style(
    style = list(
      cell_fill(color = "gray90"),
      cell_text(style = "oblique")),
    locations = cells_body(
      rows = Region == "Gulf Of Maine")) %>% 
  tab_spanner(label = md("*Annual Temperature Change*"),
              columns = c(`Warming Rate C`, `Warming Rate F`)) %>% 
 
  cols_label(
      `Warming Rate F` = "\u00b0F",
      `Warming Rate C` = "\u00b0C") %>% 
  tab_source_note(source_note = md("**Data Source:** *NOAA OISSTv2 Daily Sea Surface Temperature Data.*")) %>%                    
  tab_source_note(source_note = md("**Notes:** *Temperatures adjusted for latitudinal distotion.*") ) %>%  
  tab_footnote(
    footnote = "Gulf of Maine not a true large marine ecosystem",
    locations = cells_stub(rows = which(rates_ordered$Region == "Gulf Of Maine")) 
  )


# show table
lme_rank_table
```
 
 
```{r rates netcdf, eval = FALSE}

# Netcdf of warming rates
rates_path <- str_c(oisst_path, "warming_rates/annual_warming_rates1982to2021.nc")
rates_nc <- stack(rates_path, varname = "annual_warming_rate")
ranks_nc <- stack(rates_path, varname = "rate_percentile")



# Get paths to the LME's and their timeseries
lme_shapes <- map(lme_names, ~  read_sf(lme_paths[[.x]][["shape_path"]]))  %>% 
  setNames(lme_names)

# Include Gulf of Maine as a shape
lme_shapes[["gulf_of_maine"]] <- region_extent



# Mask with each to get warming %
lme_rates_masked <- map(lme_shapes, ~ mask_nc(ras_obj = rates_nc, mask_shape = .x))
lme_perc_masked <- map(lme_shapes, ~ mask_nc(ras_obj = ranks_nc, mask_shape = .x))


# Put the percentiles and average rates in one table
if( all(names(lme_rates_masked) == names(lme_perc_masked)) ){
  percentile_table <- map2_dfr(lme_rates_masked, lme_perc_masked, function(rate_lme, perc_lme){
    rate_val <- cellStats(rate_lme, mean, na.rm = T)
    perc_val <- cellStats(perc_lme, mean, na.rm = T)
    df_out <- data.frame(
      "annual_rate" = rate_val,
      "warming_percentile" = perc_val)
    
    return(df_out)
    
  }, .id = "ecosystem")
}


# # Check it out:
# percentile_table %>% arrange(desc(warming_percentile)) %>% head()
# percentile_table %>% arrange(desc(annual_rate)) %>% head()


# Bizarre...

# Put LME's in one collection
lme_shape_tab <- bind_rows(lme_shapes)

# Join with the  rates/ranks
perc_tab <- percentile_table %>% 
  mutate(name = str_replace_all(ecosystem, "_", " "),
         name = str_to_title(name),
         name = str_replace(name, "Of", "of"))

# Add the percentile information, and map them
lme_sf_ranks <- left_join(lme_shape_tab, perc_tab, by = c("name")) 

ggplot() +
  geom_sf(data = lme_sf_ranks, aes(fill = warming_percentile), color = "transparent") +
  geom_sf(data = world_sf, size = 0.25) +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "transparent") +
  theme(legend.position = "bottom") +
  coord_sf(expand = FALSE) +
  map_theme +
  guides(fill = guide_colorbar(
    title.position = "top",
    title.hjust = 0.5, 
    barwidth = unit(4, "in"), 
    frame.colour = "black")) +
  labs(fill = "Warming Rate Percentile",
       title = "")


```
 


## Worldwide Temperature Changes


If we look at the rates of change for each grid cell on Earths surface, rather than a plot of SST over time, it is possible to rank how each location on Earth is warming relative to others. By ranking those warming rates, and then taking the average ranking across the Gulf of Maine, we can obtain the average warming rank for our area compared to the rest of the globe.

Prior GMRI research found that the Gulf of Maine had been warming at a faster rate than 99% of the world's oceans. This is still true when comparing warming rates across certain timescales, such as from 2004 to 2013 (when our initial warming study took place).

Over a longer reference period, from 1982 to 2021, the warming rate of the Gulf of Maine, as a whole is still among the highest in the world  increasing at a rate of almost 0.05 C per year, and faster than 96.2% of the world's oceans.

However there can be significant spatial variability in this ranked warming rate across the Gulf of Maine. Warming rates across the region range from 0.014 C per year to as high as 0.097 C per year, meaning some locations over this longer time frame are still warming faster than 99.5% of the worlds oceans.

This variability reflects the various physical oceanographic drivers influencing SST in the region, and in particular the [significant interannual variability in the Gulf Streams influence on Georges Bank](https://eos.org/features/why-is-the-gulf-of-maine-warming-faster-than-99-of-the-ocean).


```{r load warming rates, eval = T}

# # 1. Warming Rates and Rankings
# Netcdf of warming rates
rates_path <- str_c(oisst_path, "warming_rates/annual_warming_rates1982to2021.nc")
rates_nc <- stack(rates_path, varname = "annual_warming_rate")
ranks_nc <- stack(rates_path, varname = "rate_percentile")

# Pull extents for the region for crop
crop_x <- st_bbox(region_extent)[c(1,3)]
crop_y <- st_bbox(region_extent)[c(2,4)]

# Expand the area out to see the larger patterns
crop_x_expanded <- crop_x + c(-2.25, 2.25)
crop_y_expanded <- crop_y + c(-0.75, 0.75)

# Make a new extent for cropping to larger area
sfc <- st_sfc(st_polygon(list(
  rbind(c(crop_x_expanded[[1]], crop_y_expanded[[1]]),  
        c(crop_x_expanded[[1]], crop_y_expanded[[2]]), 
        c(crop_x_expanded[[2]], crop_y_expanded[[2]]), 
        c(crop_x_expanded[[2]], crop_y_expanded[[1]]), 
        c(crop_x_expanded[[1]], crop_y_expanded[[1]])))))
sfc <- st_as_sf(sfc)


# Get the average warming rates for each area
get_masked_vals <- function(masked_ranks, masked_rates){
  m1 <- masked_ranks
  rank_mean <- cellStats(m1, mean)
  rank_min <- cellStats(m1, min)
  rank_max <- cellStats(m1, max)
  
  # Get stats from rates
  m2 <- masked_rates
  rate_mean <- cellStats(m2, mean)
  rate_min <- cellStats(m2, min)
  rate_max <- cellStats(m2, max)
  
  # put in table
  table_out <- tibble("Mean Rank" = rank_mean,
                "Min Rank"  = rank_min,
                "Max Rank"  = rank_max,
                "Mean Rate" = rate_mean,
                "Min Rate"  = rate_min,
                "Max Rate"  = rate_max) %>% 
    mutate_all(round, 3)
  
  # spit them out
  return(table_out)
}


# Get the rank information that go with each area
ranks_masked <- mask_nc(ranks_nc, region_extent)
rates_masked <- mask_nc(rates_nc, region_extent)
region_ranks <- get_masked_vals(ranks_masked, rates_masked)

# Prep it for text input.
avg_rank <- region_ranks$`Mean Rank` *100
avg_rate <- region_ranks$`Mean Rate`
low_rank <- region_ranks$`Min Rank` *100
low_rate <- region_ranks$`Min Rate`
top_rank <- region_ranks$`Max Rank` *100
top_rank <- ifelse(top_rank == 100, "greater than or equal to 99.5", top_rank)
top_rate <- region_ranks$`Max Rate`

```



```{r mapping warming rates, layout="l-body-outset", fig.width=6, fig.height=4}
# for an expanded view need to use the global rates
world_rate_stars <- st_as_stars(rotate(rates_nc))

# Reclassify raster into 10% Bins
world_rates_binned <- world_rate_stars %>% 
  mutate(
    rate = Annual.Sea.Surface.Temperature.Warming.Rate,
    rate_bin = cut(rate, breaks = seq(0, 1, by = 0.1))) %>% 
  select(-c(rate, Annual.Sea.Surface.Temperature.Warming.Rate))


# Color limit for palettes
temp_limits <- c(-.08, .08)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0C")


# Northeast Shelf
global_rates_zoom <- ggplot() +
  geom_stars(data = world_rate_stars) +
  geom_sf(data = new_england, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = canada, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = greenland, color = "white", fill = "gray30", size = .05) +
  scale_y_continuous(breaks = seq(-90, 90, by = 5)) +
  scale_x_continuous(breaks = seq(-180, 180, by = 10))+
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       breaks = temp_breaks, 
                       labels = temp_labels,
                       oob = oob_squish) +
  map_theme +
  coord_sf(
    xlim = c("xmin" = -77, "xmax" = -50),
    ylim = c("ymin" = 32, "ymax" = 48),
    expand = T) +
  guides("fill" = guide_colorbar(
    title = "Annual Warming Rate: 1982-2021",
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(2.5, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# Global View
global_rates <- ggplot() +
  geom_stars(data = world_rate_stars) +
  geom_sf(data = world_sf, color = "white", fill = "gray30", size = .1) +
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       breaks = temp_breaks, 
                       labels = temp_labels,
                       oob = oob_squish) +
  map_theme +
  coord_sf(expand = F) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  guides("fill" = guide_colorbar(title = "Annual Warming Rate: 1982-2021",
                                 title.position = "top", 
                                 title.hjust = 0.5,
                                 barwidth = unit(2.5, "in"),
                                 frame.colour = "black",
                                 ticks.colour = "black"))  +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


# plot map
rates_global_figure <- (global_rates_zoom | global_rates ) + 
  plot_annotation(title = "Global SST Change", 
                  caption = "Data Source: OISSTv2") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")


# plot patchwork
rates_global_figure
```




### Warming Rate Rankings


Based on data from 1982-2021, the warming rates of Gulf of Maine have been some of the highest in the world. The area as a whole has been increasing at a rate of `r avg_rate`$^{\circ}C/year$ which is faster than `r avg_rank`% of the world's oceans.

Over that same period locations within the Gulf of Maine have been warming at rates as low as `r low_rate`$^{\circ}C/year$ and as rapidly as `r top_rate`$^{\circ}C/year$, corresponding to ranks as low as `r low_rank`% and as high as `r top_rank`%.


```{r highlighting gom warming rate, eval = TRUE, layout="l-body-outset", fig.width=6, fig.height=4}
# for an expanded view need to use the global ranks
world_rank_stars <- st_as_stars(rotate(ranks_nc))

# Reclassify raster into 10% Bins
world_ranks_binned <- world_rank_stars %>% 
  mutate(
    rank = Annual.Sea.Surface.Temperature.Warming.Rate.Rank,
    rank_bin = cut(rank, breaks = seq(0, 1, by = 0.1))) %>% 
  select(-c(rank, Annual.Sea.Surface.Temperature.Warming.Rate.Rank))

# # Define the number of colors you want
# nb.cols <- 10
# mycolors <- RColorBrewer::brewer.pal(n = 10, "Reds")


# Northeast Shelf
global_ranks_zoom <- ggplot() +
  geom_stars(data = world_ranks_binned) +
  geom_sf(data = new_england, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = canada, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = greenland, color = "white", fill = "gray30", size = .05) +
  scale_y_continuous(breaks = seq(-90, 90, by = 5)) +
  scale_x_continuous(breaks = seq(-180, 180, by = 10))+
  scale_fill_brewer(palette = "Spectral", direction = -1, na.value = "transparent") +
  # scale_fill_viridis_d(option = "viridis", direction = 1, na.value = "transparent", na.translate = FALSE) +
  map_theme +
  coord_sf(
    xlim = c("xmin" = -77, "xmax" = -50),
    ylim = c("ymin" = 32, "ymax" = 48),
    expand = T) +
  guides("fill" = guide_colorsteps(
    title = "Warming Rate Percentile: 1982-2021",
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(2.5, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# Global View
global_ranks <- ggplot() +
  geom_stars(data = world_ranks_binned) +
  geom_sf(data = world_sf, color = "white", fill = "gray30", size = .1) +
  scale_fill_brewer(palette = "Spectral", direction = -1, na.value = "transparent") +
  # scale_fill_viridis_d(option = "viridis", direction = 1, na.value = "transparent", na.translate = FALSE) +
  map_theme +
  coord_sf(expand = F) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  guides("fill" = guide_colorsteps(title = "Warming Rate Percentile: 1982-2021",
                                 title.position = "top", 
                                 title.hjust = 0.5,
                                 barwidth = unit(2.5, "in"),
                                 frame.colour = "black",
                                 ticks.colour = "black"))  +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


# plot map
ranks_global_figure <- (global_ranks_zoom | global_ranks ) + 
  plot_annotation(title = "Global SST Change Rankings", 
                  caption = "Data Source: OISSTv2") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")


# plot patchwork
ranks_global_figure

```



## Fastest Warming Regions

To get a better look at extremes we can mask values below the 80th percentile and look more closely at the differences among the fastest warming areas on Earth. Below only the fastest 20% of locations are shown, to  show more detail among the fastest warming areas.


```{r}

# Mask out where its under .8
ranks_top20 <- ranks_nc
ranks_top20[ranks_top20 < 0.8] <- NA


# Make stars, bin by .25
ranks_top_stars <- st_as_stars(rotate(ranks_top20))

ranks_top_binned <- ranks_top_stars %>% 
  mutate(
    rank = Annual.Sea.Surface.Temperature.Warming.Rate.Rank,
    rank_bin = cut(rank, breaks = seq(0.8, 1, by = 0.025))) %>% 
  select(-c(rank, Annual.Sea.Surface.Temperature.Warming.Rate.Rank))

# Remake maps


# Northeast Shelf
top_ranks_zoom <- ggplot() +
  geom_stars(data = ranks_top_binned) +
  geom_sf(data = new_england, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = canada, color = "white", fill = "gray30", size = .05) +
  geom_sf(data = greenland, color = "white", fill = "gray30", size = .05) +
  scale_y_continuous(breaks = seq(-90, 90, by = 5)) +
  scale_x_continuous(breaks = seq(-180, 180, by = 10))+
  scale_fill_viridis_d(option = "plasma", direction = 1, na.value = "gray10") +
  map_theme +
  coord_sf(
    xlim = crop_x_expanded,
    ylim = crop_y_expanded,
    # xlim = c("xmin" = -75, "xmax" = -60),
    # ylim = c("ymin" = 38, "ymax" = 48),
    expand = T) +
  guides("fill" = guide_colorsteps(
    title = "Warming Rate Percentile: 1982-2021",
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(2.5, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# Global View
top_ranks <- ggplot() +
  geom_stars(data = ranks_top_binned) +
  geom_sf(data = world_sf, color = "white", fill = "gray30", size = .1) +
   scale_fill_viridis_d(option = "plasma", direction = 1, na.value = "gray10") +
  map_theme +
  coord_sf(expand = F) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  guides("fill" = guide_colorsteps(title = "Warming Rate Percentile: 1982-2021",
                                 title.position = "top", 
                                 title.hjust = 0.5,
                                 barwidth = unit(2.5, "in"),
                                 frame.colour = "black",
                                 ticks.colour = "black"))  +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())


# plot map
top_ranks_figure <- (top_ranks_zoom | top_ranks ) + 
  plot_annotation(title = "Earth's Fastest Warming Ocean Regions", 
                  caption = "Data Source: OISSTv2, rankings below 80th percentile masked.") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")


# plot patchwork
top_ranks_figure


```





## A Note on Data Sources:

NOAA_ERSST_V5 data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/data/gridded/data.noaa.ersst.v5.html.

NOAA High Resolution SST data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html. 



```{r}


####  Saving Figures  ####


# ERSST timeline
ggsave(plot = ersst_long_term_p, filename = str_c(save_location, "ersst_long_term_temps.png"),
       dpi = "retina", bg = "white")


# How ERSSt and OISST are different
ggsave(plot = ersst_v_oisst, filename = str_c(save_location, "ersst_and_oisst_temps.png"),
       dpi = "retina", bg = "white")


# simplified temperature plot
ggsave(plot = temp_simplified, filename = str_c(save_location, "oisst_simple_trends.png"),
       dpi = "retina", bg = "white")

# accelerating warming plot
ggsave(plot = gom_recent_warming, filename = str_c(save_location, "oisst_recent_gomwarming.png"),
       dpi = "retina", bg = "white")


# 2021 anomaly bar chart
ggsave(plot = gom_anomaly_bars, filename = str_c(save_location, "oisst_gom_anom_bars.png"),
       dpi = "retina", bg = "white")


# Full year timeline of temps and anomalies
ggsave(plot = year_timeline, filename = str_c(save_location, "oisst_lastyr_timeline.png"),
       dpi = "retina", bg = "white")


# Heatwave status heatmap
ggsave(plot = heatwave_heatmap, filename = str_c(save_location, "oisst_gom_heatmap.png"),
       dpi = "retina", bg = "white")



# All year temperature horizons
ggsave(plot = gom_annual_horizons, filename = str_c(save_location, "oisst_gom_horizons.png"),
       dpi = "retina", bg = "white")


# 2021 | 2012 horizons
ggsave(plot = comparison_horizons, filename = str_c(save_location, "gom_hottest_horizons.png"),
       dpi = "retina", bg = "white")


# streamplot
ggsave(plot = hw_streamplot, filename = str_c(save_location, "gom_anomaly_streamplot.png"),
       dpi = "retina", bg = "white")


# record years panels
ggsave(plot = hottest_yr_panels, 
       filename = str_c(save_location, "gom_hottest_year_panels.png"),
       dpi = "retina", bg = "white")


# Weekly Report Table
gtsave(data = lme_rank_table, 
       filename = str_c(save_location, "lme_warming_table.png"))


# Ranks maps
ggsave(plot = ranks_map, filename = str_c(save_location, "record_years_map.png"),
       dpi = "retina", bg = "white")


# northeast warmest years
ggsave(plot = northeast_records, filename = str_c(save_location, "record_years_ne_map.png"),
       dpi = "retina", bg = "white")


# Global warming rates
ggsave(plot = rates_global_figure, filename = str_c(save_location, "global_rates_map.png"),
       dpi = "retina", bg = "white")


# global warming rankings
ggsave(plot = ranks_global_figure, filename = str_c(save_location, "global_ranks_map.png"),
       dpi = "retina", bg = "white")

# top 20% rankings
ggsave(plot = top_ranks_figure, filename = str_c(save_location, "ranks_top20.png"),
       dpi = "retina", bg = "white")



  
```





`r insert_gmri_footer()`