---
title: "SST Drivers"
author: "Adam A. Kemberling"
date: "2/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SST & Climate Drivers

This will serve as the documentation of different checks at the Gulf of Maine's relationship to differrent climate drivers. The original source was the Gulf of Maine annual report, refer to that markdown if variables or data are not found:

## Relationship to Climate Patterns

To better get a sense of what kind of year 2021 was it is helpful to relate it to what we would expect from similar years. 


### ENSO Index

One large feature in global weather patterns is the El Niño Southern Oscillation (ENSO). While ENSO and the southern oscillation index are derived from temperature/pressure differences in the Pacific Ocean, they influence atmospheric behavior coming across the United States which has the potential to influence SST patterns.

 
```{r soi data, eval = TRUE}
# Source: https://www.ncdc.noaa.gov/teleconnections/enso/soi

# Does not work:
# library(rsoi)
# clim_idx <- download_soi()


# Manually copied from link above
soi_std <- tibble::tribble(#####
~"YEAR", ~"JAN", ~"FEB", ~"MAR", ~"APR", ~"MAY", ~"JUN", ~"JUL", ~"AUG", ~"SEP", ~"OCT", ~"NOV", ~"DEC",    
1951,   1.5,   0.9,  -0.1,  -0.3,  -0.7,   0.2,  -1.0,  -0.2,  -1.1,  -1.0,  -0.8,  -0.7,
1952,  -0.9,  -0.6,   0.5,  -0.2,   0.8,   0.7,   0.5,   0.1,  -0.2,   0.4,   0.0,  -1.2,
1953,   0.3,  -0.5,  -0.2,   0.2,  -1.7,   0.1,  -0.0,  -1.2,  -1.2,   0.1,  -0.3,  -0.5,
1954,   0.7,  -0.3,   0.3,   0.6,   0.5,   0.1,   0.4,   1.1,   0.2,   0.3,   0.1,   1.4,
1955,  -0.5,   1.9,   0.6,  -0.1,   1.0,   1.3,   1.6,   1.5,   1.3,   1.5,   1.2,   1.0,
1956,   1.3,   1.6,   1.3,   0.9,   1.4,   1.1,   1.1,   1.2,   0.1,   1.8,   0.2,   1.1,
1957,   0.6,  -0.1,   0.2,   0.2,  -0.7,   0.2,   0.2,  -0.5,  -0.9,   0.0,  -1.0,  -0.3,
1958,  -1.9,  -0.5,   0.3,   0.4,  -0.5,   0.3,   0.4,   0.9,  -0.3,   0.1,  -0.4,  -0.6,
1959,  -0.9,  -1.4,   1.3,   0.4,   0.5,  -0.1,  -0.3,  -0.1,   0.0,   0.5,   0.9,   0.9,
1960,   0.1,   0.1,   1.0,   0.8,   0.5,   0.1,   0.5,   0.8,   0.7,   0.1,   0.5,   0.8,
1961,  -0.3,   0.9,  -1.8,   0.8,   0.3,   0.1,   0.2,   0.2,   0.1,  -0.3,   0.5,   1.5,
1962,   2.0,  -0.3,   0.1,   0.2,   1.1,   0.7,   0.1,   0.6,   0.4,   1.0,   0.3,   0.2,
1963,   1.0,   0.6,   1.1,   0.8,   0.4,  -0.5,  -0.1,   0.0,  -0.6,  -1.2,  -0.8,  -1.2,
1964,  -0.4,   0.0,   1.1,   1.1,   0.2,   0.8,   0.6,   1.5,   1.3,   1.3,   0.2,  -0.3,
1965,  -0.4,   0.4,   0.8,  -0.5,   0.2,  -0.6,  -1.8,  -0.7,  -1.3,  -0.9,  -1.5,   0.2,
1966,  -1.3,  -0.2,  -0.9,  -0.2,  -0.4,   0.3,   0.1,   0.6,  -0.2,  -0.1,  -0.0,  -0.3,
1967,   1.7,   1.7,   1.2,  -0.0,  -0.0,   0.6,   0.2,   0.7,   0.5,   0.1,  -0.4,  -0.6,
1968,   0.5,   1.3,   0.1,   0.0,   1.2,   1.1,   0.7,   0.3,  -0.3,  -0.1,  -0.3,   0.2,
1969,  -1.5,  -0.5,   0.4,  -0.4,  -0.2,   0.2,  -0.5,  -0.1,  -1.0,  -0.9,  -0.1,   0.4,
1970,  -1.1,  -1.0,   0.6,  -0.1,   0.4,   1.0,  -0.4,   0.6,   1.2,   1.0,   1.6,   1.9,
1971,   0.4,   2.0,   2.3,   1.7,   0.9,   0.4,   0.2,   1.5,   1.4,   1.7,   0.5,   0.3,
1972,   0.5,   1.1,   0.6,  -0.1,  -1.6,  -0.5,  -1.4,  -0.5,  -1.4,  -0.9,  -0.3,  -1.3,
1973,  -0.3,  -1.4,   0.7,   0.1,   0.4,   1.1,   0.6,   1.3,   1.2,   0.8,   2.6,   1.8,
1974,   2.4,   2.1,   2.4,   0.9,   1.0,   0.4,   1.1,   0.8,   1.1,   0.9,  -0.1,   0.2,
1975,  -0.5,   0.8,   1.6,   1.2,   0.6,   1.3,   1.9,   2.0,   2.1,   1.7,   1.2,   2.1,
1976,   1.4,   1.7,   1.7,   0.3,   0.4,   0.3,  -0.9,  -0.8,  -1.1,   0.4,   0.7,  -0.3,
1977,  -0.4,   1.2,  -0.5,  -0.4,  -0.5,  -0.9,  -1.1,  -0.8,  -0.8,  -1.0,  -1.3,  -1.1,
1978,  -0.3,  -2.7,  -0.2,  -0.3,   1.4,   0.7,   0.6,   0.4,   0.1,  -0.4,  -0.0,  -0.1,
1979,  -0.4,   1.0,   0.1,  -0.1,   0.5,   0.6,   1.3,  -0.2,   0.1,  -0.1,  -0.4,  -0.7,
1980,   0.4,   0.3,  -0.4,  -0.6,  -0.0,  -0.0,  -0.0,   0.4,  -0.5,   0.0,  -0.3,  -0.1,
1981,   0.4,  -0.2,  -1.3,  -0.1,   0.8,   1.2,   0.8,   0.7,   0.3,  -0.4,   0.2,   0.5,
1982,   1.2,   0.3,   0.6,   0.1,  -0.3,  -1.0,  -1.5,  -1.7,  -1.7,  -1.7,  -2.6,  -2.2,
1983,  -3.5,  -3.6,  -2.4,  -0.9,   0.6,   0.0,  -0.6,   0.1,   0.9,   0.4,  -0.1,   0.0,
1984,   0.2,   0.9,  -0.2,   0.3,   0.2,  -0.3,   0.2,   0.4,   0.1,  -0.3,   0.3,  -0.1,
1985,  -0.3,   1.2,   0.8,   1.2,   0.4,  -0.4,  -0.1,   1.0,   0.0,  -0.4,  -0.2,   0.2,
1986,   1.0,  -1.0,   0.5,   0.3,  -0.2,   1.0,   0.3,  -0.4,  -0.5,   0.6,  -1.2,  -1.4,
1987,  -0.7,  -1.2,  -1.3,  -1.4,  -1.3,  -1.1,  -1.4,  -0.9,  -1.0,  -0.4,  -0.0,  -0.5,
1988,  -0.1,  -0.4,   0.6,   0.1,   0.9,   0.1,   1.0,   1.5,   1.8,   1.4,   1.7,   1.2,
1989,   1.5,   1.2,   1.1,   1.6,   1.2,   0.7,   0.9,  -0.3,   0.5,   0.8,  -0.2,  -0.5,
1990,  -0.1,  -1.8,  -0.4,   0.2,   1.2,   0.3,   0.5,  -0.2,  -0.7,   0.3,  -0.5,  -0.2,
1991,   0.6,   0.3,  -0.7,  -0.6,  -1.0,  -0.1,   0.0,  -0.4,  -1.5,  -1.0,  -0.7,  -1.8,
1992,  -2.9,  -0.9,  -2.0,  -1.0,   0.3,  -0.6,  -0.6,   0.4,   0.1,  -1.4,  -0.7,  -0.6,
1993,  -0.9,  -0.7,  -0.5,  -1.2,  -0.3,  -0.8,  -0.8,  -0.9,  -0.7,  -1.1,  -0.1,   0.2,
1994,  -0.1,   0.3,  -0.7,  -1.3,  -0.7,  -0.4,  -1.3,  -1.2,  -1.6,  -1.1,  -0.6,  -1.2,
1995,  -0.4,  -0.1,   0.8,  -0.7,  -0.4,   0.1,   0.4,   0.3,   0.3,   0.0,   0.0,  -0.5,
1996,   1.0,   0.3,   1.1,   0.8,   0.3,   1.2,   0.7,   0.7,   0.6,   0.6,  -0.1,   0.9,
1997,   0.5,   1.7,  -0.4,  -0.6,  -1.3,  -1.4,  -0.8,  -1.4,  -1.4,  -1.5,  -1.2,  -1.0,
1998,  -2.7,  -2.0,  -2.4,  -1.4,   0.3,   1.0,   1.2,   1.2,   1.0,   1.1,   1.0,   1.4,
1999,   1.8,   1.0,   1.3,   1.4,   0.2,   0.3,   0.5,   0.4,  -0.1,   1.0,   1.0,   1.4,
2000,   0.7,   1.7,   1.3,   1.2,   0.4,  -0.2,  -0.2,   0.7,   0.9,   1.1,   1.8,   0.8,
2001,   1.0,   1.7,   0.9,   0.2,  -0.5,   0.3,  -0.2,  -0.4,   0.2,  -0.0,   0.7,  -0.8,
2002,   0.4,   1.1,  -0.2,  -0.1,  -0.8,  -0.2,  -0.5,  -1.0,  -0.6,  -0.4,  -0.5,  -1.1,
2003,  -0.2,  -0.7,  -0.3,  -0.1,  -0.3,  -0.6,   0.3,   0.1,  -0.1,   0.0,  -0.3,   1.1,
2004,  -1.3,   1.2,   0.4,  -0.9,   1.0,  -0.8,  -0.5,  -0.3,  -0.3,  -0.1,  -0.7,  -0.8,
2005,   0.3,  -3.1,   0.3,  -0.6,  -0.8,   0.4,   0.2,  -0.3,   0.4,   1.2,  -0.2,  -0.0,
2006,   1.7,   0.1,   1.8,   1.1,  -0.5,  -0.2,  -0.6,  -1.0,  -0.6,  -1.3,   0.1,  -0.3,
2007,  -0.8,  -0.1,   0.2,  -0.1,  -0.1,   0.5,  -0.3,   0.4,   0.2,   0.7,   0.9,   1.7,
2008,   1.8,   2.6,   1.4,   0.7,  -0.1,   0.6,   0.3,   1.0,   1.2,   1.3,   1.3,   1.4,
2009,   1.1,   1.9,   0.4,   0.8,  -0.1,   0.1,   0.2,  -0.2,   0.3,  -1.2,  -0.6,  -0.7,
2010,  -1.1,  -1.5,  -0.7,   1.2,   0.9,   0.4,   1.8,   1.8,   2.2,   1.7,   1.3,   2.9,
2011,   2.3,   2.7,   2.5,   1.9,   0.4,   0.2,   1.0,   0.4,   1.0,   0.8,   1.1,   2.5,
2012,   1.1,   0.5,   0.7,  -0.3,   0.0,  -0.4,  -0.0,  -0.2,   0.2,   0.3,   0.3,  -0.6,
2013,  -0.1,  -0.2,   1.5,   0.2,   0.8,   1.2,   0.8,   0.2,   0.3,  -0.1,   0.7,   0.1,
2014,   1.4,   0.1,  -0.9,   0.8,   0.5,   0.2,  -0.2,  -0.7,  -0.7,  -0.6,  -0.9,  -0.6,
2015,  -0.8,   0.2,  -0.7,  -0.0,  -0.7,  -0.6,  -1.1,  -1.4,  -1.6,  -1.7,  -0.5,  -0.6,
2016,  -2.2,  -2.0,  -0.1,  -1.2,   0.4,   0.6,   0.4,   0.7,   1.2,  -0.3,  -0.1,   0.3,
2017,   0.2,  -0.1,   0.9,  -0.2,   0.3,  -0.4,   0.8,   0.5,   0.6,   0.9,   0.9,  -0.1,
2018,   1.1,  -0.5,   1.5,   0.5,   0.4,  -0.1,   0.2,  -0.3,  -0.9,   0.4,  -0.1,   1.0,
2019,  -0.0,  -1.4,  -0.3,   0.1,  -0.4,  -0.5,  -0.4,  -0.1,  -1.2,  -0.4,  -0.8,  -0.6,
2020,   0.2,  -0.1,  -0.1,   0.2,   0.4,  -0.4,   0.4,   1.1,   0.9,   0.5,   0.7,   1.8,
2021,   1.9,   1.5,   0.4,   0.3,   0.5,   0.4,   1.4,   0.6,   0.8,   0.7,   1.0,   1.5
  
)

#####

# Pick a date centered in each month so you can plot them as one timeline
month_centers <- data.frame(Month = toupper(month.abb),
                            Month_num = str_pad(seq(1,12,1), width = 2, pad = "0", side = "left"))
                            
# Reformat
soi_long <- soi_std %>% 
  pivot_longer(names_to = "Month", values_to = "soi_z", cols = JAN:DEC) %>% 
  left_join(month_centers, by = "Month") %>% 
  mutate(Date = as.Date(str_c(YEAR, "-", Month_num, "-15")),
         ribbon_fill = ifelse(soi_z > 0,"above", "below"),
         ribbon_above = ifelse(soi_z > 0, soi_z, NA),
         above_ymin   = ifelse(soi_z > 0, 0, NA),
         ribbon_below = ifelse(soi_z < 0, soi_z, NA),
         below_ymin = ifelse(soi_z < 0, 0, NA)
         )

# Set color scale and break points
cutpoints <- rev(seq(-4, 4, 1))
col_rdbu <- rev(RColorBrewer::brewer.pal(n = length(cutpoints), name = "RdBu"))


# fix GeomRibbon
GeomRibbon$handle_na <- function(data, params) {  data }

# Make figure
ggplot(filter(soi_long, YEAR >= 1982), aes(x = Date, y =  soi_z)) +
  geom_ribbon(aes(ymin = above_ymin, ymax = ribbon_above, fill = "La Niña"),  alpha = 0.8) +
  geom_ribbon(aes(ymin = below_ymin, ymax = ribbon_below, fill = "El Niño"), alpha = 0.8) +
  scale_fill_manual(values = c("La Niña" = col_rdbu[1], "El Niño" = col_rdbu[8])) +
  geom_line(size = 0.25, color = "gray20") + 
  labs(x = "Date", y = "SOI", title = "ENSO Index",
       caption = "Years overlapping OISST data displayed", fill = "") + 
  theme_gmri() +
  theme(legend.position = "bottom")

```
 
 
 <aside>
 
```{r enso relationship}
#| eval = TRUE,
#| fig.alt = "Scatterplot checking correlation between Gulf of Maine SST & ENSO"


# Change the month to numeric to get dates
month_key <- data.frame(
  "month_num" = str_pad(c(1:12), side = "left", pad = "0", width = 2),
  "month" = as.character(month.abb),
  stringsAsFactors = FALSE)

# Get monthly OISST Temps in format to combine with ERSST
oisst_monthly <- region_hw %>% 
  group_by(yr, month) %>% 
  summarise(sst = mean(sst),
            sst_anom = mean(sst_anom),
            .groups = "drop") %>% 
  left_join(month_key, by = "month") %>% 
  mutate(Date = as.Date(str_c(yr, "-", month_num, "-15"))) %>% 
  select(Date, yr, sst, sst_anom)


# Put them in one table
record_combined <- bind_rows(
  list("ERSSTv5" = ersst_tl,
       "OISSTv2" = oisst_monthly), 
  .id = "Temperature Data Record")


# Join the SOI info
temps_soi <- soi_long %>% 
  select(Date, soi_z) %>% 
  inner_join(record_combined, by = "Date") %>% 
  mutate(yr = str_sub(Date, 1, 4),
         decade = floor_decade(yr))


# Plot yearly averages against SOI
ggplot(temps_soi, aes(soi_z, sst_anom, color = `Temperature Data Record`)) +
  geom_point() +
  scale_color_gmri() +
  labs(x = "SOI", y = "Temperature Anomaly", title = "ENSO x SST Anomalies")

```

</aside> 
 
When we look at average temperatures for the Gulf of Maine, and how that relates to the ENSO index, we found there was **NO** relationship. Meaning that a current month's El niño/ La Nińa status did not correlate with warmer/hotter months. 

### North Atlantic Oscillation
 
A similar climate index to ENSO, with more relevance to our area is the North Atlantic Oscillation (NAO). The NAO index is based on the surface sea-level pressure difference between the Subtropical (Azores) High and the Subpolar Low.

Strong positive phases of the NAO tend to be associated with above-normal temperatures in the eastern United States and across northern Europe and below-normal temperatures in Greenland and oftentimes across southern Europe and the Middle East. 


```{r NAO index, eval = TRUE}
# NAO Data Source:
# https://www.ncdc.noaa.gov/teleconnections/nao/

nao_idx <- tribble( #####
  ~"YEAR", ~"Jan", ~"Feb", ~"Mar", ~"Apr", ~"May", ~"Jun", ~"Jul", ~"Aug", ~"Sep", ~"Oct", ~"Nov", ~"Dec",
  1950,   0.92,   0.40,  -0.36,   0.73,  -0.59,  -0.06,  -1.26,  -0.05,   0.25,   0.85,  -1.26,  -1.02,
  1951,   0.08,   0.70,  -1.02,  -0.22,  -0.59,  -1.64,   1.37,  -0.22,  -1.36,   1.87,  -0.39,   1.32,
  1952,   0.93,  -0.83,  -1.49,   1.01,  -1.12,  -0.40,  -0.09,  -0.28,  -0.54,  -0.73,  -1.13,  -0.43,
  1953,   0.33,  -0.49,  -0.04,  -1.67,  -0.66,   1.09,   0.40,  -0.71,  -0.35,   1.32,   1.04,  -0.47,
  1954,   0.37,   0.74,  -0.83,   1.34,  -0.09,  -0.25,  -0.60,  -1.90,  -0.44,   0.60,   0.40,   0.69,
  1955,  -1.84,  -1.12,  -0.53,  -0.42,  -0.34,  -1.10,   1.76,   1.07,   0.32,  -1.47,  -1.29,   0.17,
  1956,  -0.22,  -1.12,  -0.05,  -1.06,   2.21,   0.10,  -0.75,  -1.37,   0.24,   0.88,   0.51,   0.10,
  1957,   1.05,   0.11,  -1.26,   0.49,  -0.79,  -0.72,  -1.19,  -0.55,  -1.66,   1.32,   0.73,   0.12,
  1958,  -0.54,  -1.06,  -1.96,   0.37,  -0.24,  -1.38,  -1.73,  -1.56,  -0.07,   0.16,   1.64,  -0.70,
  1959,  -0.87,   0.68,  -0.15,   0.36,   0.39,   0.40,   0.74,   0.06,   0.88,   0.89,   0.41,   0.44,
  1960,  -1.29,  -1.89,  -0.50,   1.36,   0.45,  -0.21,   0.35,  -1.40,   0.39,  -1.73,  -0.51,   0.06,
  1961,   0.41,   0.45,   0.55,  -1.55,  -0.36,   0.86,  -0.39,   0.90,   1.24,   0.51,  -0.62,  -1.48,
  1962,   0.61,   0.55,  -2.47,   0.99,  -0.10,   0.16,  -2.47,   0.14,  -0.37,   0.41,  -0.23,  -1.32,
  1963,  -2.12,  -0.96,  -0.43,  -1.35,   2.16,  -0.43,  -0.77,  -0.64,   1.79,   0.94,  -1.27,  -1.92,
  1964,  -0.95,  -1.43,  -1.20,   0.36,   0.52,   1.29,   1.90,  -1.77,   0.20,   0.74,  -0.01,  -0.15,
  1965,  -0.12,  -1.55,  -1.51,   0.72,  -0.62,   0.29,   0.32,   0.45,   0.37,   0.38,  -1.66,   1.37,
  1966,  -1.74,  -1.39,   0.56,  -0.75,   0.22,   1.05,   0.32,  -1.76,  -0.45,  -0.68,  -0.04,   0.72,
  1967,  -0.89,   0.19,   1.51,   0.18,  -0.99,   1.40,   0.41,   1.44,   0.93,   0.07,   0.60,  -0.45,
  1968,   0.13,  -1.29,   0.40,  -1.08,  -1.76,   0.33,  -0.80,  -0.66,  -1.92,  -2.30,  -0.93,  -1.40,
  1969,  -0.83,  -1.55,  -1.56,   1.53,   0.55,   0.55,   0.57,  -1.45,   2.07,   0.66,  -0.96,  -0.28,
  1970,  -1.50,   0.64,  -0.96,  -1.30,   1.14,   1.55,   0.10,   0.10,  -0.09,  -0.92,  -0.60,  -1.20,
  1971,  -1.13,   0.24,  -0.84,  -0.24,   0.50,  -1.57,   0.24,   1.55,   0.39,   0.58,  -0.20,   0.60,
  1972,   0.27,   0.32,   0.72,  -0.22,   0.95,   0.88,   0.18,   1.32,  -0.12,   1.09,   0.54,   0.19,
  1973,   0.04,   0.85,   0.30,  -0.54,  -0.44,   0.39,   0.57,  -0.06,  -0.30,  -1.24,  -0.93,   0.32,
  1974,   1.34,  -0.14,  -0.03,   0.51,  -0.24,  -0.14,  -0.76,  -0.64,   0.82,   0.49,  -0.54,   1.50,
  1975,   0.58,  -0.62,  -0.61,  -1.60,  -0.52,  -0.84,   1.55,  -0.26,   1.56,  -0.54,   0.41,   0.00,
  1976,  -0.25,   0.93,   0.75,   0.26,   0.96,   0.80,  -0.32,   1.92,  -1.29,  -0.08,   0.17,  -1.60,
  1977,  -1.04,  -0.49,  -0.81,   0.65,  -0.86,  -0.57,  -0.45,  -0.28,   0.37,   0.52,  -0.07,  -1.00,
  1978,   0.66,  -2.20,   0.70,  -1.17,   1.08,   1.38,  -1.14,   0.64,   0.46,   1.93,   3.04,  -1.57,
  1979,  -1.38,  -0.67,   0.78,  -1.71,  -1.03,   1.60,   0.83,   0.96,   1.01,  -0.30,   0.53,   1.00,
  1980,  -0.75,   0.05,  -0.31,   1.29,  -1.50,  -0.37,  -0.42,  -2.24,   0.66,  -1.77,  -0.37,   0.78,
  1981,   0.37,   0.92,  -1.19,   0.36,   0.20,  -0.45,   0.05,   0.39,  -1.45,  -1.35,  -0.38,  -0.02,
  1982,  -0.89,   1.15,   1.15,   0.10,  -0.53,  -1.63,   1.15,   0.26,   1.76,  -0.74,   1.60,   1.78,
  1983,   1.59,  -0.53,   0.95,  -0.85,  -0.07,   0.99,   1.19,   1.61,  -1.12,   0.65,  -0.98,   0.29,
  1984,   1.66,   0.72,  -0.37,  -0.28,   0.54,  -0.42,  -0.07,   1.15,   0.17,  -0.07,  -0.06,   0.00,
  1985,  -1.61,  -0.49,   0.20,   0.32,  -0.49,  -0.80,   1.22,  -0.48,  -0.52,   0.90,  -0.67,   0.22,
  1986,   1.11,  -1.00,   1.71,  -0.59,   0.85,   1.22,   0.12,  -1.09,  -1.12,   1.55,   2.29,   0.99,
  1987,  -1.15,  -0.73,   0.14,   2.00,   0.98,  -1.82,   0.52,  -0.83,  -1.22,   0.14,   0.18,   0.32,
  1988,   1.02,   0.76,  -0.17,  -1.17,   0.63,   0.88,  -0.35,   0.04,  -0.99,  -1.08,  -0.34,   0.61,
  1989,   1.17,   2.00,   1.85,   0.28,   1.38,  -0.27,   0.97,   0.01,   2.05,  -0.03,   0.16,  -1.15,
  1990,   1.04,   1.41,   1.46,   2.00,  -1.53,  -0.02,   0.53,   0.97,   1.06,   0.23,  -0.24,   0.22,
  1991,   0.86,   1.04,  -0.20,   0.29,   0.08,  -0.82,  -0.49,   1.23,   0.48,  -0.19,   0.48,   0.46,
  1992,  -0.13,   1.07,   0.87,   1.86,   2.63,   0.20,   0.16,   0.85,  -0.44,  -1.76,   1.19,   0.47,
  1993,   1.60,   0.50,   0.67,   0.97,  -0.78,  -0.59,  -3.18,   0.12,  -0.57,  -0.71,   2.56,   1.56,
  1994,   1.04,   0.46,   1.26,   1.14,  -0.57,   1.52,   1.31,   0.38,  -1.32,  -0.97,   0.64,   2.02,
  1995,   0.93,   1.14,   1.25,  -0.85,  -1.49,   0.13,  -0.22,   0.69,   0.31,   0.19,  -1.38,  -1.67,
  1996,  -0.12,  -0.07,  -0.24,  -0.17,  -1.06,   0.56,   0.67,   1.02,  -0.86,  -0.33,  -0.56,  -1.41,
  1997,  -0.49,   1.70,   1.46,  -1.02,  -0.28,  -1.47,   0.34,   0.83,   0.61,  -1.70,  -0.90,  -0.96,
  1998,   0.39,  -0.11,   0.87,  -0.68,  -1.32,  -2.72,  -0.48,  -0.02,  -2.00,  -0.29,  -0.28,   0.87,
  1999,   0.77,   0.29,   0.23,  -0.95,   0.92,   1.12,  -0.90,   0.39,   0.36,   0.20,   0.65,   1.61,
  2000,   0.60,   1.70,   0.77,  -0.03,   1.58,  -0.03,  -1.03,  -0.29,  -0.21,   0.92,  -0.92,  -0.58,
  2001,   0.25,   0.45,  -1.26,   0.00,  -0.02,  -0.20,  -0.25,  -0.07,  -0.65,  -0.24,   0.63,  -0.83,
  2002,   0.44,   1.10,   0.69,   1.18,  -0.22,   0.38,   0.62,   0.38,  -0.70,  -2.28,  -0.18,  -0.94,
  2003,   0.16,   0.62,   0.32,  -0.18,   0.01,  -0.07,   0.13,  -0.07,   0.01,  -1.26,   0.86,   0.64,
  2004,  -0.29,  -0.14,   1.02,   1.15,   0.19,  -0.89,   1.13,  -0.48,   0.38,  -1.10,   0.73,   1.21,
  2005,   1.52,  -0.06,  -1.83,  -0.30,  -1.25,  -0.05,  -0.51,   0.37,   0.63,  -0.98,  -0.31,  -0.44,
  2006,   1.27,  -0.51,  -1.28,   1.24,  -1.14,   0.84,   0.90,  -1.73,  -1.62,  -2.24,   0.44,   1.34,
  2007,   0.22,  -0.47,   1.44,   0.17,   0.66,  -1.31,  -0.58,  -0.14,   0.72,   0.45,   0.58,   0.34,
  2008,   0.89,   0.73,   0.08,  -1.07,  -1.73,  -1.39,  -1.27,  -1.16,   1.02,  -0.04,  -0.32,  -0.28,
  2009,  -0.01,   0.06,   0.57,  -0.20,   1.68,  -1.21,  -2.15,  -0.19,   1.51,  -1.03,  -0.02,  -1.93,
  2010,  -1.11,  -1.98,  -0.88,  -0.72,  -1.49,  -0.82,  -0.42,  -1.22,  -0.79,  -0.93,  -1.62,  -1.85,
  2011,  -0.88,   0.70,   0.61,   2.48,  -0.06,  -1.28,  -1.51,  -1.35,   0.54,   0.39,   1.36,   2.52,
  2012,   1.17,   0.42,   1.27,   0.47,  -0.91,  -2.53,  -1.32,  -0.98,  -0.59,  -2.06,  -0.58,   0.17,
  2013,   0.35,  -0.45,  -1.61,   0.69,   0.57,   0.52,   0.67,   0.97,   0.24,  -1.28,   0.90,   0.95,
  2014,   0.29,   1.34,   0.80,   0.31,  -0.92,  -0.97,   0.18,  -1.68,   1.62,  -1.27,   0.68,   1.86,
  2015,   1.79,   1.32,   1.45,   0.73,   0.15,  -0.07,  -3.18,  -0.76,  -0.65,   0.44,   1.74,   2.24,
  2016,   0.12,   1.58,   0.73,   0.38,  -0.77,  -0.43,  -1.76,  -1.65,   0.61,   0.41,  -0.16,   0.48,
  2017,   0.48,   1.00,   0.74,   1.73,  -1.91,   0.05,   1.26,  -1.10,  -0.61,   0.19,  -0.00,   0.88,
  2018,   1.44,   1.58,  -0.93,   1.24,   2.12,   1.09,   1.39,   1.97,   1.67,   0.93,  -0.11,   0.61,
  2019,   0.59,   0.29,   1.23,   0.47,  -2.62,  -1.09,  -1.43,  -1.17,  -0.16,  -1.41,   0.28,   1.20,
  2020,   1.34,   1.26,   1.01,  -1.02,  -0.41,  -0.15,  -1.23,   0.12,   0.98,  -0.65,   2.54,  -0.30,
  2021,  -1.11,   0.14,   0.73,  -1.43,  -1.24,   0.77,   0.03,  -0.28,  -0.21,  -2.29,  -0.18,   0.29
)
#####

# Reformat and add a center date
month_centers <- data.frame(Month = month.abb,
                            Month_num = str_pad(seq(1,12,1), width = 2, pad = "0", side = "left"))
                            
# Reformat
nao_long <- nao_idx %>% 
  pivot_longer(names_to = "Month", values_to = "nao_z", cols = Jan:Dec) %>% 
  left_join(month_centers, by = "Month") %>% 
  mutate(Date = as.Date(str_c(YEAR, "-", Month_num, "-15")),
         ribbon_fill  = ifelse(nao_z > 0,"above", "below"),
         ribbon_above = ifelse(nao_z > 0, nao_z, NA),
         above_ymin   = ifelse(nao_z > 0, 0, NA),
         ribbon_below = ifelse(nao_z < 0, nao_z, NA),
         below_ymin   = ifelse(nao_z < 0, 0, NA)
         )

# Make figure
ggplot(filter(nao_long, YEAR >= 1982), 
       aes(x = Date, y =  nao_z)) +
  geom_ribbon(aes(ymin = above_ymin, ymax = ribbon_above, fill = "Positive Forcing"),  alpha = 0.8) +
  geom_ribbon(aes(ymin = below_ymin, ymax = ribbon_below, fill = "Negative Forcing"), alpha = 0.8) +
  scale_fill_manual(values = c("Positive Forcing" = col_rdbu[1], "Negative Forcing" = col_rdbu[8])) +
  geom_line(size = 0.25, color = "gray20") + 
  labs(x = "Date", y = "NAO", title = "NAO Index",
       caption = "Years overlapping OISST data displayed", fill = "") + 
  theme_gmri() +
  theme(legend.position = "bottom")
```





<aside>
```{r, eval = TRUE}
#| eval = TRUE,
#| fig.alt = "Scatterplot checking correlation between Gulf of Maine SST & NAO Index"


# Join the SOI info
temps_nao <- nao_long %>% 
  select(Date, nao_z) %>% 
  inner_join(temps_soi, by = "Date") 


# Plot yearly averages against SOI
ggplot(temps_nao, aes(nao_z, sst_anom, color = `Temperature Data Record`)) +
  geom_point() +
  scale_color_gmri() +
  labs(x = "NAO", y = "Temperature Anomaly", title = "NAO x SST Anomalies")
```
</aside>

Looking at the NAO index, we found there was also not a strong relationship to our region's Sea surface temperature anomalies. This lends strength to the argument that recent changes are independent of normal climate oscillations, but is not the end of the story.


