---
title: "2021: Gulf of Maine's Warmest Year on Record"
author: 
  - first_name: "Adam"
    last_name: "Kemberling"
    url: https://github.com/adamkemberling
    affiliation: Gulf of Maine Research Institute
    affiliation_url: https://www.gmri.org
description: | 
  Characteristics of a record setting year for ocean temperatures
date: "Updated on: `r Sys.Date()`"
editor_options: 
  chunk_output_type: console
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = FALSE, 
                      message = F, 
                      warning = F, 
                      comment = "", 
                      fig.align = "center", 
                      fig.height = 6, 
                      fig.width = 8)

####  Packages  ####
library(lubridate)
library(raster)
library(here)
library(rnaturalearth)
library(scales)
library(sf)
library(stars)
library(gmRi)
library(heatwaveR)
library(gt)
library(gtExtras)
library(patchwork)
library(tidyverse)
library(ggstream)
library(ggforce)
library(geomtextpath)
library(ggHoriPlot)
library(ggradar)


# Support Functions
source(here("R/oisst_support_funs.R"), verbose = FALSE)
source(here("R/temp_report_support.R"), verbose = FALSE)

#box paths
oisst_path <- cs_path("res", "OISST/oisst_mainstays")


# Set ggplot theme for figures
theme_set(theme_bw() + 
            theme(
              # Titles
              plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
              plot.subtitle = element_text(size = 9),
              plot.caption = element_text(size = 7.2, margin = margin(t = 20)),
              legend.title = element_text(size = 9),
              legend.text = element_text(size = 7.5),
              # Axes
              axis.line.y = element_line(),
              axis.ticks.y = element_line(), 
              axis.line.x = element_line(),
              axis.ticks.x = element_line(), 
              axis.text = element_text(size = 11),
              # Facets
              strip.text = element_text(color = "white", 
                                        face = "bold",
                                        size = 11),
              strip.background = element_rect(
                color = "white", 
                fill = "#36454F", 
                size = 1, 
                linetype="solid"))
          )


# File paths for various extents based on "apershing_gulf_of_maine"
region_paths <- get_timeseries_paths(region_group = "gmri_sst_focal_areas", 
                                     mac_os = "mojave")

# Timeseries Path
timeseries_path <- region_paths[["apershing_gulf_of_maine"]][["timeseries_path"]]

# Polygon Path
poly_path <- region_paths[["apershing_gulf_of_maine"]][["shape_path"]]

# Adding Logo to plot
logo_path <- paste0(system.file("stylesheets", package = "gmRi"), "/gmri_logo.png")
lab_logo <- magick::image_read(logo_path)

# Add using grid::grid.raster()
# grid::grid.raster(lab_logo, x = 0.08, y = 0.03, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```

`r gmRi::use_gmri_style_rmd(css = "gmri_rmarkdown.css")`

# 2021 Sea Surface Temperature

#### An exceptional year

2021 was an exceptionally warm year for the Gulf of Maine. It had the warmest fall on record for the area, and the second warmest summer. Much of the year the region experienced what would be considered "heatwave conditions".

```{r}
# Load the timeseries
region_timeseries <- read_csv(timeseries_path, col_types = cols(), guess_max = 1e6)


# Clean up the data - add labels
region_timeseries <- region_timeseries %>% 
  distinct(time, .keep_all = T) %>% 
  mutate(time = as.Date(time),
         yr = year(time),
         month_num = month(time),
         month = month(time, label = T, abbr = T),
         week = lubridate::week(time),
         doy = yday(time),
         season = metR::season(month_num, lang = "en"),
         #Set up correct year for winters, they carry across into next year
         season_eng = case_when(
           season == "SON" ~ "Fall",
           season == "DJF" ~ "Winter",
           season == "MAM" ~ "Spring",
           season == "JJA" ~ "Summer"),
         season_yr = ifelse( (season_eng == "Winter" & month_num %in% c(1,2)), yr - 1, yr)
         )
```


```{r}
# Get heatwave statuses for each day:
# Uses area weighted sst by default
region_hw <- pull_heatwave_events(temperature_timeseries = region_timeseries)

# Format dates and seasons and heatwave outcomes
region_hw <- region_hw %>% 
  mutate(time = as.Date(time),
         yr = year(time),
         month_num = month(time),
         month = month(time, label = T, abbr = T),
         week = lubridate::week(time),
         doy = yday(time),
         season = metR::season(month_num, lang = "en"),
         season_eng = case_when(
           season == "SON" ~ "Fall",
           season == "DJF" ~ "Winter",
           season == "MAM" ~ "Spring",
           season == "JJA" ~ "Summer"),
         #Set up correct year for winters, they carry across into next year
         season_yr = ifelse(
           (season_eng == "Winter" & month_num %in% c(1,2)), yr - 1, yr)
         )

# Set up heatwave status outcomes for fancy table features:
region_hw <- region_hw %>% 
  mutate(hw_outcomes = case_when(
    mhw_event == TRUE ~ 1,
    mcs_event == TRUE ~ 0,
    TRUE ~ 0.5))

```


## Climate Patterns

To better get a sense of what kind of year 2021 was it is helpful to relate it to what we would expect from similar years. One large factor in global weather patterns is the El Niño Southern Oscillation., here is where the year fell in regards to the El nino southern oscillation.

 - El nino status
 - NAO status
 
 When we look at average temperatures and how that relates to these patterns, there is a general relationship of El nino years correlating well with ___, and la nina years correlating more with __. 
 
 
 

## Temperature and Anomaly Patterns


```{r}

 # Last Calendar Year
this_yr <- region_hw %>% 
  filter(year(time) == 2021)


# Set colors by name
color_vals <- c(
  "Sea Surface Temperature" = "royalblue",
  "Heatwave Event"          = "darkred",
  "MHW Threshold"           = "coral3",
  "Daily Climatology"        = "gray30")


# Set the label with degree symbol
ylab <- "Sea Surface Temperature"



# Plot the last 365 days
hw_temp_p <- ggplot(this_yr, aes(x = time)) +
    geom_segment(aes(x = time, xend = time, y = seas, yend = sst), 
                 color = "royalblue", alpha = 0.25) +
    geom_segment(aes(x = time, xend = time, y = mhw_thresh, yend = hwe), 
                 color = "darkred", alpha = 0.25) +
    geom_line(aes(y = sst, color = "Sea Surface Temperature")) +
    geom_line(aes(y = hwe, color = "Heatwave Event")) +
    geom_line(aes(y = mhw_thresh, color = "MHW Threshold"), lty = 3, size = .5) +
    geom_line(aes(y = seas, color = "Daily Climatology"), lty = 2, size = .5) +
    scale_color_manual(values = color_vals) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month") +
    scale_y_continuous(sec.axis = sec_axis(trans = ~as_fahrenheit(., data_type = "temperature"),
                                         labels =  number_format(suffix = " \u00b0F")),
                     labels = number_format(suffix = " \u00b0C")) +
    theme(legend.title = element_blank(),
          legend.position = "none") +
    labs(x = "", 
         y = ylab)

```


```{r heatwave anomaly timeline}

# Plot the last 365 days - anomaly scale
linetype_key <- c(
  "Sea Surface Temperature Anomaly" = 1,
  "Heatwave Event"                  = 1,
  "MHW Threshold"                   = 3,
  "Daily Climatology"               = 2)

# Same Plot as Anomalies
hw_anom_p <- this_yr %>% 
  mutate(
    sst = sst,
    seas = seas,
    sst_anom = sst_anom,
    mhw_thresh = mhw_thresh,
    anom_thresh = mhw_thresh - seas,
    anom_hwe = hwe - seas) %>% 
ggplot(aes(x = time)) +
    geom_segment(aes(x = time, xend = time, 
                     y = 0, yend = sst_anom), 
                 color = "royalblue", alpha = 0.25) +
    geom_segment(aes(x = time, xend = time, 
                       y = anom_thresh, yend = anom_hwe), 
                 color = "darkred", alpha = 0.25) +
    geom_line(aes(y = sst_anom, color = "Sea Surface Temperature Anomaly")) +
    geom_line(aes(y = anom_hwe, color = "Heatwave Event")) +
    geom_line(aes(y = anom_thresh, color = "MHW Threshold"), lty = 3, size = .5) +
    geom_line(aes(y = 0, color = "Daily Climatology"), lty = 2, size = .5) +
    scale_color_manual(values = color_vals) +
    scale_linetype_manual(values = linetype_key, guide = "none") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month") +
    guides(color = guide_legend(override.aes = list(linetype = linetype_key), nrow = 2)) +
    theme(legend.title = element_blank(),
          legend.position = "top") +
    scale_y_continuous(sec.axis = sec_axis(trans = ~as_fahrenheit(., data_type = "anomalies"),
                                           labels =  number_format(suffix = " \u00b0F")),
                       labels = number_format(suffix = " \u00b0C")) +
    labs(x = "", 
         y = "Temperature Anomaly", 
         caption = paste0("Climate reference period :  1982-2011"))
```


```{r}
# Show figure
hw_temp_p / hw_anom_p

```


## Heatwave Days

```{r}

#Use gg hori plot instead of heatmap
# use facet zoom to pull out this year:

# Prep the legend title
guide_lab <- "Sea Surface Temperature Anomaly \u00b0C"

# Set new axis dimensions, y = year, x = day within year
# use a flate_date so that they don't stair step
base_date <- as.Date("2000-01-01")
grid_data <- region_hw %>% 
  mutate(year = year(time),
         yday = yday(time),
         flat_date = as.Date(yday-1, origin = base_date))


# Set palette limits to center it on 0 with scale_fill_distiller
limit <- max(abs(grid_data$sst_anom)) *c(-1,1)



```


## Heatwave Events

```{r}

# Isolate 2021
hw_2021 <- filter(region_hw, yr == "2021")

# Get Event numbers
events_21 <-  hw_2021 %>% drop_na(mhw_event_no) %>% distinct(mhw_event_no) %>% pull(mhw_event_no)
events_num <-  length(events_21)


# Pull their data
events_dat <- filter(region_hw, mhw_event_no %in% as.character(events_21))
  

# How long on average
avg_duration <- round(nrow(events_dat) / events_num, digits = 0)
```

During 2021 there were three main marine heatwave events, including one that began on the new year. The average duration for each heatwave event was `r avg_duration` days, or over three months long.


```{r}
#| alt
```



## Heatwave Seasonality

```{r}
# Polar plot comparing when each year experienced peak heatwave conditions
```



## Heatwave Day Heatmap

```{r}
# Assemble heatmap plot
heatwave_heatmap <- ggplot(grid_data, aes(x = flat_date, y = year)) +
  
  # background box fill for missing dates
  geom_rect(xmin = base_date, xmax = base_date + 365, 
            ymin = min(grid_data$year) - .5, ymax = max(grid_data$year) + .5, 
            fill = "gray75", color = "transparent") +
  
  # tile for sst colors
  geom_tile(aes(fill = sst_anom)) +
  
  # points for heatwave events
  geom_point(data = filter(grid_data, mhw_event == TRUE),
             aes(x = flat_date, y = year), size = .25)  +
  scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = c(0,0)) +
  scale_y_continuous(limits = c(1980.5, 2021.5), expand = c(0,0)) +
  labs(x = "", 
       y = "",
       "\nClimate reference period : 1982-2011") +
  
  #scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  scale_fill_distiller(palette = "RdYlBu", na.value = "transparent", 
                       limit = limit) +
  
  #5 inches is default rmarkdown height for barheight
  guides("fill" = guide_colorbar(title = guide_lab, 
                                 title.position = "right", 
                                 title.hjust = 0.5,
                                 barheight = unit(4.8, "inches"), 
                                 frame.colour = "black", 
                                 ticks.colour = "black")) +  
  theme_classic() +
  theme(legend.title = element_text(angle = 90))





# Assemble pieces
heatwave_heatmap
```


## Following Each Heatwave

Do some breakdowns of how each heatwave looked:
 - where was heat concentrated
 - how long did it last
 - was anything unique about it?



##  Temperature Horizons

```{r}
hori_data <- grid_data %>% 
  mutate(year = factor(year),
         year = fct_rev(year))


# cut out outliers and set cutpoints
cutpoints <- hori_data %>% 
  mutate(
    outlier = between(
      sst_anom,
      quantile(sst_anom, 0.25, na.rm=T)-
        1.5*IQR(sst_anom, na.rm=T),
      quantile(sst_anom, 0.75, na.rm=T)+
        1.5*IQR(sst_anom, na.rm=T))) %>% 
  filter(outlier)

# Set origin
#ori <- sum(range(cutpoints$sst_anom)) / 2
ori <- 0

# Set horizon scale cutpoints
# sca <- seq(range(cutpoints$sst_anom)[1], 
#            range(cutpoints$sst_anom)[2], 
#            length.out = 7)[-4]

# Manually pick them
sca <- seq(-4, 4, length.out = 9)

sca_bins <- str_c(sca[1:(length(sca)-1)], " to ", sca[2:length(sca)], "\u00b0C")
sca_labels <- rev(sca_bins
                  )


# Function to plot one or more years
anom_horizon_plot <- function(hori_data, origin = ori, scale_cutpoints = sca, labels = sca_labels){
  
   ggplot(hori_data) +
  geom_horizon(aes(flat_date, 
                   sst_anom,
                   fill = ..Cutpoints..), 
               origin = ori, 
               horizonscale = sca) +
  scale_fill_hcl(palette = 'RdBu', reverse = T, labels = sca_labels) +
  #scale_fill_manual(values = setNames(sca_colors, sca_names)) +
  facet_grid(year~.) +
  theme_classic() +
  theme(
    panel.spacing.y=unit(0, "lines"),
    strip.text.y = element_text(size = 7, angle = 0, hjust = 0),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.border = element_blank(),
    legend.position = "left"
    ) +
  scale_x_date(expand=c(0,0), 
               date_breaks = "1 month", 
               date_labels = "%b") +
  labs(x = '', 
       fill = "Temperature Anomaly \u00b0C") 
  
  
}
```


```{r}
# All Years
anom_horizon_plot(hori_data = hori_data, origin = ori, scale_cutpoints = sca)

```

## Comparing the Gulf's Two Hottest Years


Early into 2021 it was apparent that the year was on-par with the previous title-holder for warmest year on record.

```{r}
# Reset the outliers
# cut out outliers and set cutpoints
cutpoints_2 <- hori_data %>% 
  filter(year %in% c("2012", "2021")) %>% 
  mutate(
    outlier = between(
      sst_anom,
      quantile(sst_anom, 0.25, na.rm=T)-
        1.5*IQR(sst_anom, na.rm=T),
      quantile(sst_anom, 0.75, na.rm=T)+
        1.5*IQR(sst_anom, na.rm=T))) %>% 
  filter(outlier)



# 2012
hori_2012 <- hori_data %>% 
  filter(year == "2012") %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  theme_minimal(base_size = 10) +
  facet_wrap(~year, strip.position = "top")
  
# 2021
hori_2021 <- hori_data %>% filter(year == "2021") %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  theme_minimal(base_size = 10) +
  facet_wrap(~year, strip.position = "top") +
  theme(plot.caption = element_text(colour = "gray25", size = 6)) +
  labs(caption = "Climatatological Reference Period: 1982-2011")

# Put them together as one figure
comparison_horizons <- hori_2012 / hori_2021 + plot_layout(guides = "collect")

# Do a bunch of formatting
comparison_horizons &
  theme(
    # axis text/ticks
    axis.title.y = element_blank(),
    axis.line.x = element_line(),
    axis.ticks.x = element_line(),
    axis.line.y = element_line(),
    axis.text.y = element_blank(),
    # Legend 
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines"),
    # Gridlines
    panel.grid.minor = element_blank(),
    panel.grid = element_line(size = 0.2),
    # Plot margin
    plot.margin = unit(c(.5,1,.3,.5), "cm"),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(size = 9),
    plot.caption = element_text(size = 7.2, margin = margin(t = 20)),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 7.5),
    legend.margin = margin(c(7,0,0,0)),
    legend.justification = "center",
    # Facet Design
    strip.text = element_text(
      color = "white", 
      face = "bold",
      size = 11),
    strip.background = element_rect(
      color = "white", 
      fill = "#36454F", 
      size = 1, 
      linetype="solid")) &
   guides(fill = guide_legend(title = "Temperature Anomaly",
                             title.hjust = 0.5, 
                             title.position = "top",
                             label.position = "bottom", 
                             nrow = 1, 
                             byrow = T, reverse = T)) &
  plot_annotation(title = "Comparing the Gulf of Maine's Two Hottest Years", 
                  subtitle = "Duration of temperature anomaly horizons, colored by their strength")
  grid::grid.raster(lab_logo, x = 0.08, y = 0.04, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
  
```


## Streamplot


```{r}

# Set Factor Order for Bins
bin_order <- c(
  "Less than -1", 
  "-1 to -0.5",
  "Within 0.5",
  "0.5 to 1", 
  "Greater than 1")

# Set Colors
bin_colors <- RColorBrewer::brewer.pal(n = length(bin_order), 
                                       name = "RdBu")
bin_colors <- rev(bin_colors)

# Count Days in Bins
day_types <- region_hw %>% 
  mutate(
    `Less than -1`   = if_else(sst_anom <= -1, TRUE, FALSE),
    `-1 to -0.5`     = if_else(between(sst_anom, -1, -0.5), TRUE, FALSE),
    `Within 0.5`     = if_else(between(sst_anom, -0.5, 0.5), TRUE, FALSE),
    `0.5 to 1`       = if_else(between(sst_anom, 0.5, 1), TRUE, FALSE),
    `Greater than 1` = if_else(sst_anom >= 1, TRUE, FALSE)) %>% 
  group_by(yr) %>% 
  summarise(
    `Less than -1`   = sum(`Less than -1`, na.rm = T),
     `-1 to -0.5`    = sum(`-1 to -0.5`, na.rm = T),
    `Within 0.5`     = sum(`Within 0.5`, na.rm = T),
    `0.5 to 1`       = sum(`0.5 to 1`, na.rm = T),
    `Greater than 1` = sum(`Greater than 1`, na.rm = T),
    .groups = "drop") %>% 
  pivot_longer(cols = "Less than -1":"Greater than 1", names_to = "Anomaly Strength", values_to = "day_total", values_drop_na = FALSE) %>% 
  mutate(day_total = if_else(is.na(day_total), 0L, day_total),
         `Anomaly Strength` = factor(`Anomaly Strength`, levels = bin_order))


# Plot the Streamplot
hw_streamplot <- ggplot(day_types, aes(yr, y = day_total, fill = `Anomaly Strength`)) +
  geom_stream(type = "proportion") +
  #geom_segment(y = .5, yend = 0.5, linetype = 3, color = "gray50", x = 1981, xend = 2021, size = 0.25) +
  scale_x_continuous(breaks = seq(1980,2030, by = 10), 
                     minor_breaks = seq(1975, 2030, by = 5), 
                     expand = c(.03,0)) +
  scale_y_continuous(labels = scales::percent_format(), expand = c(0,0)) + 
  scale_fill_manual(values = setNames(bin_colors, bin_order)) + 
  labs(y = "", x = "", fill = "", 
       title = "A Warmer Gulf of Maine",
       subtitle = "Fractions of Each Year Experienced as Hot/Cold Anomalies",
       caption = "1981 not a complete years in the data*") +
  theme_minimal(base_size = 10) +
  theme(
    axis.line.x = element_line(),
    axis.ticks.x = element_line(),
    axis.line.y = element_line(),
    axis.ticks.y = element_line(),
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines"),
    panel.grid.minor = element_blank(),
    panel.grid = element_line(size = 0.2),
    plot.margin = unit(c(.5,1,.3,.5), "cm"),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 9),
    plot.caption = element_text(size = 7.2, margin = margin(t = 20)),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 7.5),
    legend.margin = margin(c(7,0,0,0)),
    legend.justification = "center") +
   guides(fill = guide_legend(title = "Anomaly Strength \u00b0C",
                             title.hjust = 0.5, 
                             title.position = "left",
                             label.position = "bottom", 
                             nrow = 1, 
                             byrow = T))


# Show figure
hw_streamplot
grid::grid.raster(lab_logo, x = 0.08, y = 0.04, just = c('left', 'bottom'), width = unit(0.8, 'inches'))
```


## Characteristics of the Hottest Years

Multivariate figure comparing characteristics of both years: 2012 & 2021


```{r}

# # Which were hottest
# top_5_years <- hori_data %>% 
#   filter(year != "1981") %>% 
#   group_by(year) %>% 
#   summarise(mean_temp = mean(sst), .groups = "drop") %>% 
#   slice_max(n = 5, order_by = mean_temp) %>% 
#   pull(year)
# 
# # Pull data for 5 years
# top_years_dat <- filter(hori_data, year %in% top_5_years)


# Use all years so they can be ranked independently

# Metrics?
year_mets <- hori_data %>% 
  group_by(year) %>% 
  summarise(
    total_days = n(),
    `Average Temperature` = mean(sst, na.rm = T),
    `Temperature Above Average` = mean(sst_anom, na.rm = T),
    `Cumulative Degrees Above Average` = sum(sst_anom, na.rm = T),
    hw_days = sum(mhw_event, na.rm = T),
    hw_events = n_distinct(mhw_event_no),
    peak_temp = max(sst),
    peak_anom = max(sst_anom)) %>% 
  mutate(
    hw_day_pct = (hw_days/total_days) * 100,
    `Average Heatwave Length` = hw_days/hw_events
  ) %>% 
  pivot_longer(names_to = "Var", values_to = "Metric", cols = `Average Temperature`:`Average Heatwave Length`)


# Snipe out the top5 for each
library(tidytext)
year_met_ranks <- year_mets %>% 
  filter(Var %in% c("Average Temperature", "Temperature Above Average", "Average Heatwave Length", "Cumulative Degrees Above Average")) %>% 
  group_by(Var) %>% 
  slice_max(n = 5, order_by = Metric) %>% 
  ungroup() %>%
  mutate(
    yr_focus = ifelse(year == "2021", TRUE, FALSE),
    year = reorder_within(year, Metric, Var)) 


year_met_ranks %>% 
  ggplot(aes(year, Metric, fill = yr_focus)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~Var, scales = "free", ncol = 2) +
    coord_flip() +
    scale_x_reordered() +
    scale_y_continuous() +
    labs(y = "",
         x = NULL,
         title = "How 2021 Stacks Up:",
         subtitle = "Ranking Characteristics of Acute and Prolonged Warming Events") +
  theme(plot.title = element_text(hjust = 0)) + 
  scale_fill_gmri() +
  theme(panel.border = element_blank())

```




## Biological Events

 - whale sharks
 - hammerheads
 - blue crab
 - 




`r insert_gmri_footer()`