---
title: "Gulf of Maine Warming Update: 2022 the Regions Second Warmest Year"
author: "Adam Kemberling"
url: "https://github.com/adamkemberling"
affiliation: "Gulf of Maine Research Institute"
affiliation_url: "https://www.gmri.org"
description: | 
  Documenting one more year in a decade of above-average regional temperatures
date: "Updated on: `r Sys.Date()`"
format: 
  html:
    toc: true
    self-contained: true
execute: 
  echo: false
  message: false
  warning: false
  comment: ""
  fig.align: center
params: 
  report_year: 2022
  in_celsius: FALSE
  save_figures: 
    label: "Save Figures?" 
    value: TRUE
---

```{r}
#| label: code-setup
#| include: false

####  Packages  ####
{
library(lubridate)
library(raster)
library(here)
library(rnaturalearth)
library(rgeoboundaries)
library(scales)
library(sf)
library(stars)
library(gmRi)
library(heatwaveR)
library(gt)
library(gtExtras)
library(patchwork)
library(tidyverse)
library(ggthemes)
library(ggstream)
library(ggforce)
library(geomtextpath)
library(ggHoriPlot)
# library(tidytext)
}

#---------- Support Functions
suppressPackageStartupMessages( source(here("R/oisst_support_funs.R"), verbose = FALSE) )
suppressPackageStartupMessages( source(here("R/temp_report_support.R"), verbose = FALSE) )


#----------- Paths
oisst_path <- cs_path("res", "OISST/oisst_mainstays")

# File paths for various extents based on "apershing_gulf_of_maine"
region_paths <- get_timeseries_paths(region_group = "gmri_sst_focal_areas", 
                                     box_location = "cloudstorage")

# Folder for Comms Team Update
comm_folder <- cs_path("root", "Program Communications/Climate Center — Program Communications/Warming Updates/Visuals/")

# Folder based on year
save_location <- str_c(comm_folder, "Annual ", params$report_year,"/")


#----------- Parameters

# Set the year for the report:
report_yr <- params$report_year

# # Box Location to save hi-res images
# save_location <- cs_path("root", "GoM Seasonal Climate Update/annual_update_figures")


#----------- Components

# Timeseries Path
timeseries_path <- region_paths[["apershing_gulf_of_maine"]][["timeseries_path"]]

# Polygon Path
poly_path <- region_paths[["apershing_gulf_of_maine"]][["shape_path"]]
region_extent <- read_sf(poly_path)

# Pull extents for the region for cropping
crop_x <- st_bbox(region_extent)[c(1,3)]
crop_y <- st_bbox(region_extent)[c(2,4)]

# Expand the area out to see the larger patterns, Used as plot limits
crop_x_expanded <- crop_x + c(-2.25, 2.25)
crop_y_expanded <- crop_y + c(-0.75, 0.75)

# Make a new extent for cropping to larger area
sfc <- make_cropbox(xlims = crop_x_expanded, ylims = crop_y_expanded)

# Projection to use for everything
master_crs <- st_crs("EPSG:4326")

# Coastline Shapes
canada      <- ne_states("canada", returnclass = "sf") %>% st_transform(master_crs)
new_england <- gb_adm1(country = "usa", type = "sscgs")  %>% st_transform(master_crs)
world_sf    <- ne_countries(returnclass = "sf")  %>% st_transform(master_crs)
greenland   <- ne_states(country = "greenland", returnclass = "sf") %>% st_transform(master_crs)

# Adding Logo to plots
logo_path <- paste0(system.file("stylesheets", package = "gmRi"), "/gmri_logo.png")
lab_logo <- magick::image_read(logo_path)


```



```{r}
#| label: code-add-css
#| results: asis
#| echo: false

gmRi::use_gmri_style_rmd(css = "gmri_rmarkdown.css")
```



```{r}
#| label: code-load-timeseries

# Load the timeseries
region_timeseries <- read_csv(timeseries_path, 
                              col_types = cols(), 
                              guess_max = 1e6)


# Clean up the data - add labels
region_timeseries <- region_timeseries %>% 
  mutate(time = as.Date(time)) %>% 
  distinct(time, .keep_all = T) %>% 
  supplement_season_info() %>% 
  filter(year %in% c(1982:report_yr))



# Summarize by year to return mean annual anomalies and variance
annual_summary <- region_timeseries %>% 
  group_by(year) %>% 
  temperature_summaries() %>% 
  mutate(yr_as_dtime = as.Date(paste0(year, "-07-02")),
         anom_direction = ifelse(area_wtd_anom > 0, "Hot", "Cold"))


# Get heatwave statuses for each day:
# Uses area weighted sst by default
region_hw <- pull_heatwave_events(
  temperature_timeseries = region_timeseries, 
  threshold = 90, 
  clim_ref_period = c("1982-01-01", "2011-12-31")) %>% 
  supplement_hw_data() %>% 
  filter(doy != 366)


# Data for the Report Year Only
report_yr_only <- region_hw %>% 
  filter(year(time) == report_yr)

# Set up the date within a year for Heatmap
base_date <- as.Date("2000-01-01")
region_hw <- region_hw %>% 
  mutate(year = year(time),
         yday = yday(time),
         flat_date = as.Date(yday-1, origin = base_date),
         yr_rev = factor(year),
         yr_rev = fct_rev(yr_rev))


# # Global Temperature Anomalies
global_anoms <- read_csv(
    paste0(oisst_path, "global_timeseries/global_anoms_1982to2011.csv"), 
    guess_max = 1e6,
    col_types = cols()) %>% 
  mutate(year = year(time)) %>% 
  filter(between(year, 1982, report_yr)) %>% 
  supplement_season_info()

# summarize by year again
global_summary <- global_anoms %>% 
  group_by(year) %>% 
  temperature_summaries() %>% 
  mutate(yr_as_dtime = as.Date(paste0(year, "-07-02")),
         anom_direction = ifelse(area_wtd_anom > 0, "Hot", "Cold"))
```




```{r}
#| label: code-all-data-summary

# Get the overall average and build labels
all_year_avgs <- annual_summary %>% 
  summarise(avg_temp_c = mean(area_wtd_sst),
            avg_temp_f = mean(area_wtd_f),
            avg_anom_c = mean(area_wtd_anom),
            avg_anom_f = mean(area_wtd_anom_f),            
            .groups = "drop") %>% 
  mutate(temp_label = str_c(round(avg_temp_c, 2), " \u00b0C  |  ", round(avg_temp_f, 2), " \u00b0F"))

```



```{r}
#| label: txt-report-yr-records

# Values I need up front for text

# average annual sst, its anomaly value:
report_annual_temp <- round(mean(report_yr_only$sst_f, na.rm = T), 2)
report_annual_anom <- round(mean(report_yr_only$anom_f, na.rm = T), 2)

# What rank this year was
report_rank <- filter(annual_summary, year == report_yr) %>% pull(sst_rank)
global_report_rank <- filter(global_summary, year == report_yr) %>% pull(sst_rank)

# Record holder year's temperature
record_year_temp <- filter(annual_summary, sst_rank == 1) %>% pull(sst_f)

# Note on how the seasons ranked:
season_ranks <- region_timeseries %>% 
  supplement_season_info() %>% 
  split(.$season_eng) %>% 
  map_dfr(~ .x %>% 
    group_by(year, season_eng) %>% 
    temperature_summaries()) %>% 
    filter(year == report_yr)


# How many days in this year meet mhw criteria
# add some acknowledgement of the extreme characteristics:
hw_criteria <- report_yr_only %>% pull(mhw_event) %>% sum()
```




With an annual average sea surface temperature (SST) of `r report_annual_temp`°F — more than `r report_annual_anom`°F above normal — the Gulf of Maine experienced its `r report_rank`nd warmest year on record in `r report_yr`. The region ranked in the top 3 warmest temperatures across all seasons, with the most extreme temperatures occurring in November.



## Annual Sea Surface Temperatures



```{r}
#| label: code-warming-rates


####  Run Warming Rate regressions  ####
rate_data <- list(
  "GoM All F" = get_decadal_rates(
    temp_df = annual_summary, 
    temp_col = "area_wtd_f", 
    year_col = "year", 
    year_lim = c(1982, report_yr), 
    area_name = "GoM", 
    degree_c = F),
  "Global All F" = get_decadal_rates(
    temp_df = global_summary, 
    temp_col = "area_wtd_f", 
    year_col = "year", 
    year_lim = c(1982, report_yr), 
    area_name = "Global Ocean", 
    degree_c = F))



# Build Regression Equation Labels
eq_all_f    <- rate_data[["GoM All F"]][["eq_label"]]
eq_global_f <- rate_data[["Global All F"]][["eq_label"]]

```


The annual average SST in the Gulf of Maine for `r report_yr` was `r report_annual_temp`°F; this was `r report_annual_anom`°F above the long-term (i.e., 1982-2011) average. This year fell short of the previous warmest year on record — 2021 — `r round(record_year_temp-report_annual_temp, 2)`°F. Since the early 1980s, the rate of warming in the Gulf of Maine `r str_c(rate_data[["GoM All F"]][["decadal_rate"]], "\u00b0F / Decade")` has been more than triple that of the world’s oceans (`r str_c(rate_data[["Global All F"]][["decadal_rate"]], "\u00b0F / Decade")` @fig-global-comparison). 





```{r}
#| label: fig-global-comparison
#| fig-cap: "A timeseries of annual average sea surface temperature anomalies for the Gulf of Maine (solid black line) from 1982 through `r report_yr`, illustrating that `r report_yr` was the 2nd warmest year on record. Long-term trendlines for the Gulf of Maine (blue) and the entire globe (green) show how much more quickly the Gulf of Maine is warming compared to the rest of the world."


# Fahrenheit plot
temp_simplified <- global_rate_comparison(
  annual_summary_dat = annual_summary, 
  global_summary_dat = global_summary, 
  eq_all = str_remove(eq_all_f, "1982-2022"), 
  eq_global = str_remove(eq_global_f, "1982-2022"), 
  temp_units = "F",
  region_label = "Gulf of Maine") +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.115, y_npc = -0.175, logo_height = 1, height_units = "cm")

# add logo
temp_simplified 

```

The rate of warming in the region vary throughout the year. Comparisons across the four seasons (@fig-seasons-comparison) reveal that the Gulf of Maine is warming fastest during the July-September (nearly 4x the global ocean average), more than twice the rate seen during January-March (~3x the global ocean average).



```{r}
#| label: fig-seasons-comparison
#| fig-cap: "Four timeseries panels of the annual averages for each season (solid black lines) for the Gulf of Maine. Decadal trends are overlaid to compare decadal trends in the Gulf of Maine (blue) and the entire globe (green). Rates are much faster in the Gulf of Maine across all seasons, with the smallest difference in rates during the spring."
#| fig.width: 8
#| fig.height: 7


# Pull the data out by season: annual_summary steps but for each season
season_summary <- region_timeseries %>% 
  mutate(
    month_groups = case_when(
      month %in% c("Jan", "Feb", "Ma")   ~ "January - March",
      month %in% c("Apr", "May", "Jun")  ~ "April - June",
      month %in% c("Jul", "Aug", "Sep")  ~ "July - September",
      month %in% c("Oct", "Nov", "Dec")  ~ "October - December")) %>% 
  group_by(yr, month_groups) %>% 
  temperature_summaries()

global_season_summary <- global_anoms %>% 
  mutate(
    month_groups = case_when(
      month %in% c("Jan", "Feb", "Ma")   ~ "January - March",
      month %in% c("Apr", "May", "Jun")  ~ "April - June",
      month %in% c("Jul", "Aug", "Sep")  ~ "July - September",
      month %in% c("Oct", "Nov", "Dec")  ~ "October - December")) %>% 
  group_by(yr, month_groups) %>% 
  temperature_summaries() 




# Think we can Facet, but then the stupid rates won't be there...
season_avgs <- bind_rows(
  list(
    "Gulf of Maine" = season_summary,
    "Global Ocean" = global_season_summary), 
  .id = "Region") %>% 
  mutate(month_groups = factor(month_groups,
    levels = c("January - March", "April - June", "July - September", "October - December")))


# Decadal Rates Function
rate_calc <- function(.data){
   annual_lm <- lm(anom_f ~ yr, data = .data)
    rate_x <- round(coef(annual_lm)["yr"] * 10, 2)
    eq_lab <- paste0(": ", rate_x, deg_f, " / Decade")
    mutate(.data, rate_eq = eq_lab)
}


# Run the numbers to make the label
season_avgs <- season_avgs %>% 
  group_by(Region, month_groups) %>% 
  group_map( ~rate_calc(.), .keep = TRUE) %>% 
  bind_rows() %>% 
  mutate(rate_eq = str_c(Region, rate_eq))



# Make the four panels:
season_panels <- season_avgs %>% 
  split(.$month_groups) %>% 
  imap(function(x, y){
    
    season_p <- ggplot(x, aes(yr, anom_f)) +
      geom_line( data = filter(x, Region == "Gulf of Maine"), linewidth = 1, linetype = 1) +
      geom_point(data = filter(x, Region == "Gulf of Maine"), size = 1.5) +
      stat_smooth(
        aes(color = rate_eq), 
        method = "lm", 
        formula = y~x, 
        linetype = 1, 
        linewidth = 1.5, 
        alpha = 0.8,
        se = FALSE) +
      facet_wrap(~month_groups, ncol = 2) +
      scale_color_manual(values = c(gmri_cols("green"), gmri_cols("blue"))) +
      scale_y_continuous(labels = number_format(suffix = deg_f)) +
      theme(
        legend.position = c(0.35, 0.8),
        legend.background = element_rect(color = "black", fill = "white")) +
      labs(
        x = "Year", 
        y = "SST Anomaly",
        color = str_c(y, " Trend (1982-2022)"))
    
    if(y == "Winter"){
      season_p <- season_p +
        geom_point(data = filter(x, yr == 2022, Region == "Gulf of Maine"), aes(yr, anom_f),
                   shape = 21, color = "black", fill = "white", size = 1.5)}
    
    # Return the plot
    season_p
      
  })


# Assemble the plot
panel_plot <- (season_panels$`January - March` |season_panels$`April - June`) / (season_panels$`July - September` | season_panels$`October - December`) + plot_annotation(caption = "") &
  theme(plot.margin = margin(t = 1, b = 1, r = 1, l = 1))

panel_plot
grid::grid.raster(lab_logo, x = 0.08, y = 0.01, just = c('left', 'bottom'), height = unit(1, "cm"))
```


There can be significant spatial variability in annually-averaged SST patterns. The Gulf of Maine in `r report_yr` illustrated this fact with the region southeast of Georges Bank experiencing the highest SST anomalies (@fig-new-england-map). This area is more susceptible than other areas of the Gulf of Maine to large-scale variability in major oceanic currents, such as the relative influence of the Gulf Stream versus the Labrador Current.

```{r}
#| label: bathymetry-details

# These work but the detail takes forever with ggplot...
# Add the bottom contours:
bathy <- raster(str_c(cs_path("res","Bathy/"), "ETOPO1/NEShelf_Etopo1_bathy.tiff")) 

# # Contours for geom_contour()
# bathy_df <- as.data.frame(raster::coordinates(bathy))
# bathy_df$depth <- raster::extract(bathy, bathy_df)
# bathy_df$depth <- bathy_df$depth * -1
# contours_make <- c(200)

# Or make terra for ggplot using tidyterra
elev_terra <- terra::rast(bathy)

# ggplot() +
#   tidyterra::geom_spatraster_contour_filled(
#     data = elev_terra,
#     breaks = c(seq(0,-400,-100), -Inf),
#     show.legend = F) +
#   # tidyterra::geom_spatraster_contour(
#   #   data = elev_terra, 
#   #   breaks = seq(0,-400,-100), 
#   #   alpha = 0.75) +
#   geom_sf(data = new_england, fill = "gray90", linewidth = .15) +
#   geom_sf(data = canada, fill = "gray90", linewidth = .15) +
#   geom_sf(data = greenland, fill = "gray90", linewidth = .15) +
#   geom_sf(data = region_extent,
#           color = "black",
#           linetype = 2, linewidth = 0.5,
#           fill = "transparent") +
#   scale_fill_brewer(palette = "Blues", direction = 1) +
#   scale_y_continuous(breaks = seq(30, 50, by = 2)) +
#   coord_sf(xlim = c(-78, -56.5), ylim = c(37.8, 48))

```




```{r}
#| label: fig-new-england-map
#| fig-cap: "Map of annual average sea surface temperature anomalies in 2021. The box outlined by the black dashed line denotes the region of study for the analysis presented throughout this report."
#| fig.height: 6


# Set time limits to get data
annual_dates <- as.Date(
  c(str_c(report_yr,"-01-01"), str_c(report_yr,"-12-31")))

# Make data window for the year
data_window <- data.frame(lon = c(-81, -54.5),
                          lat = c(33.5, 50),
                          time = annual_dates)

# Load data, stack and tidy
ne_anoms <- oisst_window_load(
    data_window = data_window,
    anomalies = T,
    box_location = "cloudstorage") %>%  
  stack()

# Convert to F
ne_anoms_f <- calc(ne_anoms, function(x) as_fahrenheit(x, data_type = "anomalies")) %>% 
  setNames(names(ne_anoms))

# Process the yearly average
ne_annual_anom_f <- mean(ne_anoms_f, na.rm = T)



# Color limit for palettes
temp_limits <- c(-6, 6)
temp_breaks <- c(temp_limits[1], temp_limits[1]/2,  0, temp_limits[2]/2, temp_limits[2])
temp_labels <- str_c(c(str_c("< ", temp_limits[1]), temp_limits[1]/2, 0, temp_limits[2]/2, str_c("> ", temp_limits[2])), "\u00b0F")


# Make it a stars class to plot with ggplot
ne_anoms_st <- st_as_stars(ne_annual_anom_f)



#### 1. Map the Anomalies in Space
annual_gom_map <- ggplot() +
  geom_stars(data = ne_anoms_st) +
  tidyterra::geom_spatraster_contour(
    data = elev_terra,
    breaks = seq(0,-200,-100),
    alpha = 0.75) +
  geom_sf(data = new_england, fill = "gray90", linewidth = .15) +
  geom_sf(data = canada, fill = "gray90", linewidth = .15) +
  geom_sf(data = greenland, fill = "gray90", linewidth = .15) +
  geom_sf(data = region_extent,
          color = "black",
          linetype = 2, linewidth = 1,
          fill = "transparent") +
  scale_y_continuous(breaks = seq(30, 50, by = 2)) +
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       labels = temp_labels,
                       oob = scales::oob_squish) +
  coord_sf(xlim = c(-78, -56.5), ylim = c(37.8, 48)) +
  labs(title = str_c(report_yr, ": Average Temperature Anomalies"),
       caption = "Anomalies calculated using 1982-2011 climatology.\n100m & 200m contours overlaid to show bathymetric details.") +
  map_theme(
    legend.position = "bottom",
    legend.title = element_text(angle = 0))+
  guides("fill" = guide_colorbar(
    title = "Average Temperature Anomaly ",
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(3, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 

# Show figure
annual_gom_map
grid::grid.raster(lab_logo, x = 0.1, y = 0.06, just = c('left', 'bottom'), height = unit(1, "cm"))


```

Looking at these spatial patterns in 3-month averages (@fig-four-season-maps) shows just how “patchy” these pockets of above-average temperatures can be. During January-March, the areas to the south and southeast of the Gulf of Maine exhibit a strong mix of both above and below average surface temperature. During April-June, we can see that the water within the Gulf of Maine itself and the deeper waters just south of Georges Bank experienced warmer temperatures than the surrounding areas. In the second-half of the year, the exceptional warmth of the Gulf Stream is evident to the east and south of the area of study (dashed line), while the Gulf of Maine, itself, remains several degrees above the long-term average.



```{r}
#| label: fig-four-season-maps
#| fig-cap: "Maps of the  3-month averages in sea surface temperature anomalies in 2022."
#| fig.width: 8
#| fig.height: 6

# Four Season Panels

# Process the stack into the four 3-month averages
# Set time limits
season_dates <- list(
  "January - March" = distinct(filter(report_yr_only, month %in% c("Jan", "Feb", "Mar")), time) %>% pull(time),
  "April - June" = distinct(filter(report_yr_only, month %in% c("Apr", "May", "Jun")), time) %>% pull(time),
  "July - September"   = distinct(filter(report_yr_only, month %in% c("Jul", "Aug", "Sep")), time) %>% pull(time),
  "October - December" = distinct(filter(report_yr_only, month %in% c("Oct", "Nov", "Dec")), time) %>% pull(time)
)

# Grab the dates for each, process the mean 
season_stacks <- map(season_dates, function(x){
  dates_x <- str_c("X", str_replace_all(x, "-", "."))
  which_dates <- which(names(ne_anoms_f)  %in% dates_x)
  month_mean <- mean(ne_anoms_f[[which_dates]], na.rm = T)
  
})


# Assemble Panels Again:
season_panel_maps <- imap(season_stacks, function(x, seasn){
  
  season_df <- data.frame("season" = seasn)
  ne_season_st <- st_as_stars(x)
  season_map <- ggplot(data = season_df) +
    geom_stars(data = ne_season_st) +
    tidyterra::geom_spatraster_contour(
      data = elev_terra,
      breaks = seq(0,-200,-100),
      alpha = 0.75) +
    geom_sf(data = new_england, fill = "gray90", 
            linewidth = .25) +
    geom_sf(data = canada, fill = "gray90", 
            linewidth = .25) +
    geom_sf(data = greenland, fill = "gray90", 
            linewidth = .25) +
    geom_sf(data = region_extent, 
            color = "black", 
            linetype = 2, 
            linewidth = 1,
            fill = "transparent") +
    scale_y_continuous(breaks = seq(30, 50, by = 2)) +
    scale_fill_distiller(palette = "RdBu", 
                         na.value = "transparent", 
                         limit = temp_limits, 
                         labels = temp_labels,
                         oob = scales::oob_squish) +
    coord_sf(xlim = c(-78, -56.5), ylim = c(37.8, 48)) +
    facet_wrap(~season) +
    map_theme(
      legend.position = "bottom",
      legend.title = element_text(angle = 0))+
    guides("fill" = guide_colorbar(
      title = "Average Temperature Anomaly ",
      title.position = "top",
      title.hjust = 0.5,
      barwidth = unit(3, "in"),
      frame.colour = "black",
      ticks.colour = "black")) 

# Show figure
season_map
})



# Put them together and format them
season_panels_all <- (season_panel_maps$`January - March` | season_panel_maps$`April - June`)/(season_panel_maps$`July - September` | season_panel_maps$`October - December`) +
  plot_layout(guides = "collect") & 
  theme(legend.position = 'bottom', 
        plot.margin = margin(t = 1, b = 1, r = 1, l = 1)) 

# Display them
season_panels_all + plot_annotation(caption = "100m & 200m contours overlaid to show bathymetric details.")
grid::grid.raster(lab_logo, x = 0.08, y = 0.04, just = c('left', 'bottom'), height = unit(1, "cm"))

```


When we compare the annual average SST for `r report_yr` (`r str_c(report_annual_temp, deg_f)`) to other unusually warm years in the Gulf of Maine, we see that it was ~0.4`r deg_f` cooler than 2021, the warmest year on record(`r str_c(round(record_year_temp,2), deg_f)`). When we look at the deviation from the long-term average SST (i.e., the annual SST anomaly), the last decade stands out for its exceptional warmth (@fig-yr-ranks). `r report_yr` continues a pattern that began in 2010 of sustained above-average temperatures. With the exception of 2019 (ranked 13th), all of the last ten years remain in the top 10 for SST.

```{r}
#| label: fig-yr-ranks
#| fig-cap: "A ranking of the top 10 annual SST values [bars] and those years' respective SST anomalies [x-axis]."




# Plot the top 5 warmest years
temp_rankings_p <- annual_summary %>% 
  filter(sst_rank <= 10) %>% 
  mutate(flag_year = ifelse(year == report_yr, "flag", "unflag"),
         sst_rank =  fct_rev(factor(sst_rank))) %>% 
  ggplot() +
  geom_col(aes(x = anom_f, y = sst_rank, fill = flag_year)) +
  # Year Labels
  geom_text(
    mapping = aes(x = 0, y = sst_rank, label = str_c(sst_rank, ".  ", year)),
    hjust = -.1,
    color = 'white',
    fontface = "bold") +
  # Temperature Labels
  geom_text(
    mapping = aes(x = anom_f, y = sst_rank, label = str_c(round(sst_f,2), deg_f), group = sst_rank),
    hjust = 1.05,
    color = 'white',
    fontface = "bold") +
  scale_x_continuous(
    labels = number_format(suffix = deg_f), 
    expand = expansion(mult = c(0,0.15))) +
  scale_fill_manual(values = c("flag" = as.character(gmri_cols("orange")), 
                                "unflag" = as.character(gmri_cols("gmri blue")))) +
  labs(x = "Annually Averaged Temperature Anomaly",
       y = NULL,
       title = str_c("Ten Hottest Years in the Gulf of Maine"),
       caption = "Anomalies calculated using 1982-2011 climatology.") +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0),
        legend.position = "none", 
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(hjust = 1, face = "bold"),
        axis.ticks.y = element_blank(),
        plot.margin = margin(t = 20, l = 10, r = 5, b = 5)) +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.1, y_npc = -0.15, logo_height = 1, height_units = "cm")

temp_rankings_p
```

## Worldwide Temperatures

It was not just the Gulf of Maine that experienced exceptional warmth in `r report_yr`, however. Globally, `r report_yr` was the `r global_report_rank`rd warmest year for sea surface temperatures, and [6th-warmest for combined land & ocean temperatures.](https://www.noaa.gov/news/2022-was-worlds-6th-warmest-year-on-record). 

@fig-global-anomalies shows annual average SST anomalies for oceans all over the world in `r report_yr`. While much of the Southern Ocean and expanses of the southeastern Pacific were anomalously cool (a feature of La Niña conditions), most of the world’s oceans experienced unusually warm temperatures in `r report_yr`. This is particularly true for northern mid-latitudes, especially the northwestern Pacific (off Japan) and along the eastern seaboard of the United States—a region that includes the Gulf of Maine.


```{r}
#| label: code-global-anom-prep

# Load and format the anomalies for the year:

# Set time limits
annual_dates <- as.Date(c(
  str_c(report_yr,"-01-01"), str_c(report_yr,"-12-31")))

# Make data window for season
global_data_window <- data.frame(lon = c(-360, 180),
                          lat = c(-90, 90),
                          time = annual_dates)

# Load data
global_anoms <- oisst_window_load(
  data_window = global_data_window, 
  anomalies = T, 
  box_location = "cloudstorage")

# Stack and tidy
global_anoms <- stack(global_anoms)
global_anoms <- setExtent(global_anoms, extent(c(0, 360, -90, 90)))
global_anoms <- rotate(global_anoms)


# Change to F
global_anoms_f <- calc(global_anoms, function(x) as_fahrenheit(x, data_type = "anomalies"))

# Get Average for the year
global_anom_mean_f <- mean(global_anoms_f, na.rm = T)


```


```{r}
#| label: fig-global-anomalies
#| fig-cap: "Map of annual average sea surface temperature anomalies for the world’s oceans in 2022."
#| fig.width: 8
#| fig.height: 6

# Make it a stars class to plot with ggplot
global_anoms_st <- st_as_stars(global_anom_mean_f)

#### 1. Map the Anomalies in Space
annual_anom_map <- ggplot() +
  geom_stars(data = global_anoms_st) +
  geom_sf(data = world_sf, fill = "gray90", linewidth = .25) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  scale_fill_distiller(palette = "RdBu", 
                       na.value = "transparent", 
                       limit = temp_limits, 
                       breaks = temp_breaks,
                       labels = temp_labels,
                       oob = scales::oob_squish) +
  labs(title = str_c(report_yr, ": Average Temperature Anomalies"),
       caption = "Anomalies calculated using 1982-2011 climatology.") +
  coord_sf(expand = FALSE) +
  map_theme(
    legend.position = "bottom",
    legend.title = element_text(angle = 0),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()) +
  guides("fill" = guide_colorbar(
    title = "Average Temperature Anomaly ",
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(3, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 

# Show figure
annual_anom_map
grid::grid.raster(lab_logo, x = 0.08, y = 0.08, just = c('left', 'bottom'), height = unit(1, "cm"))
```



## Daily Sea Surface Temperatures

```{r}
#| label: txt-extreme-events

# typical range
seasonal_cycle_min <- min(report_yr_only$seas_f)
seasonal_cycle_max <- max(report_yr_only$seas_f)
seasonal_range <- round(seasonal_cycle_max - seasonal_cycle_min, 2)

# report_yr range
report_yr_min <- min(report_yr_only$sst_f)
report_yr_min_anom <- round(min(report_yr_only$anom_f), 2)
report_yr_min_date <- format(report_yr_only$time[report_yr_only$sst_f == report_yr_min], "%b-%d")
report_yr_max <- max(report_yr_only$sst_f)
report_yr_max_anom <- round(max(report_yr_only$anom_f), 2)
report_yr_max_date <- format(report_yr_only$time[report_yr_only$sst_f == report_yr_max], "%b-%d")
report_yr_range <- round(report_yr_max - report_yr_min, 2)
report_yr_max <- round(report_yr_max, 2)
report_yr_min <- round(report_yr_min, 2)
```


The annual cycle of SST in the Gulf of Maine exhibits a familiar pattern with the lowest annual temperatures being observed in March, while the highest SSTs are observed in August (@fig-annual_hw_timeseries). The average difference between the annual maximum SST in August and the annual minimum SST in March is `r seasonal_range` °F. In `r report_yr`, the difference between the maximum SST (`r report_yr_max_date`, `r report_yr_max`°F) and minimum SST (`r report_yr_min_date`, `r report_yr_min`°F) was `r report_yr_range` °F. Daily SST anomaly values never fell below +`r report_yr_min_anom` °F and reached as high as +`r report_yr_max_anom` °F above the long-term average.


```{r}
#| label: fig-annual_hw_timeseries
#| fig.height: 4
#| fig-cap: "A timeseries of marine heatwave (MHW) conditions in the Gulf of Maine extending from January 1 through December 301 2022. Black lines representing the long-term (i.e., 1982 – 2011) average SST, the 10th percentile (i.e., cold spell threshold), and 90th percentile (i.e., heatwave threshold) for a given day in the Gulf of Maine are labeled to indicate climatological reference points. A solid line (red for marine heatwave or blue for a non-heatwave day) indicate the observed SST for each day; red and blue shading illustrates whether each day is considered part of a MHW event."
#| fig-alt: "A line chart shows the average sea surface temperature for each day of the year, and overlaid against it are notably smoother black lines representing the long-term mean, 10th percentile, and 90th percentile temperatures for those same days of the year. The area between the mean and the observed temperature is filled in with a solid color to indicate whether that day meets marine heatwave criteris. The observed temperatures for 2022 are above the 90th percentile for nearly the entire year, with the exception of two periods less than a week long in June and October."


# annual heatwave timeseries


# Show MHW Status Through the Year:
hw_temp_p <- year_hw_temps_two(year_hw_dat = report_yr_only, temp_units = "F") + 
  labs(
    title = str_c("Gulf of Maine SST: ", params$report_year))  +
  scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(add = c(0,0))) +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.9, y_npc = 1.125, logo_height = 1, height_units = "cm")
hw_temp_p

```



## Record Setting Daily Temperatures

@fig-hottest-days shows the highest SST values ever recorded for each day of the year in the Gulf of Maine. Record-setting days that occurred in `r report_yr` are denoted with dark red dots (@fig-hottest-days). Indeed, daily high SST records were set for more than half of all days in November and December. The most daily records were set in November, which experienced record high SSTs for 24 of the month's 30 days - or 80% of the month. In total, record high daily temperatures were reached on 169 days during`r report_yr`; nearly 46% of days in `r report_yr` experienced record high SSTs in the Gulf of Maine.

```{r}
#| label: code-hottest-days
#| eval: true
#| fig-cap: "Timeseries of the record high daily sea surface temperature in the Gulf of Maine for every day of the year. Record-breaking temperatures that were observed in `r report_yr` are denoted with red dots; when a record-breaking temperature for a given day was observed in any other year, that daily record is in gray."

# New hottest daily temps - maybe scratch
# Grab the hottest day for each day of the year
hottest_days <- region_hw %>% 
  group_by(doy) %>% 
  slice_max(order_by = sst_anom, n = 1) %>% 
  ungroup() 

# Total Number in report_yr
hottest_report_yr <- hottest_days %>% 
  count(yr) %>% 
  filter(yr == report_yr) %>% 
  pull(n)

# # Daily Record Temperatures
hottest_days <- hottest_days %>%
  mutate(
    record_period = case_when(
      yr == report_yr ~ as.character(report_yr),
      yr < report_yr & yr > 2000 ~ "Since 2010",
      yr < 2000 ~ "Before 2000"),
    flat_date = as.Date(str_c(report_yr, "-01-01")) + (doy - 1))
```


```{r}
#| label: fig-hottest-days
#| eval: false
#| fig-cap: "Timeseries of the record high daily sea surface temperature in the Gulf of Maine for every day of the year. Record-breaking temperatures that were observed in `r report_yr` are denoted with red dots; when a record-breaking temperature for a given day was observed in any other year, that daily record is in gray."


# Timeseries of what the temperatures are for the daily records
hottest_days_plot <- ggplot() +
  geom_point(
    data = filter(hottest_days, record_period != report_yr), 
    aes(x = flat_date, y = sst_f, color = record_period), alpha = 0.5, size = 1) +
  geom_point(
    data = filter(hottest_days, record_period == report_yr), 
    aes(x = flat_date, y = sst_f, color = record_period), size = 1.5) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(add = c(0,0))) +
  scale_y_continuous(labels = number_format(suffix = deg_f)) +
  scale_color_manual(values = c("darkred", "black", gmri_cols("orange"))) +
  labs(x = "Day of the Year",
       y = "Sea Surface Temperature",
       title = "Hottest Days on Record",
       subtitle = "Gulf of Maine",
       color = "Record Setting Year",
       caption = str_c("Temperature data covers period of 1982-", report_yr))  +
   theme(legend.position = c(0.15, 0.75),
         legend.background = element_rect(color = "black", fill = "white")) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5))  +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.1, y_npc = -0.175, logo_height = 1, height_units = "cm")


hottest_days_plot


#--------------  As geom_tile instead

{
# # New hottest daily temps - maybe scratch
# 
# # Grab the hottest day for each day of the year
# hottest_days <- region_hw %>% 
#   group_by(doy) %>% 
#   slice_max(order_by = sst_anom, n = 1) %>% 
#   ungroup() 
# 
# 
# # Total Number in report_yr
# hottest_report_yr <- hottest_days %>% 
#   count(yr) %>% 
#   filter(yr == report_yr) %>% 
#   pull(n)
# 
# 
# # # Daily Record Temperatures
# hottest_days <- hottest_days %>%
#   mutate(
#     record_period = case_when(
#       yr == report_yr ~ "New Record Daily Temperature",
#       yr < report_yr & yr > 2000 ~ "Last Decade",
#       yr < 2000 ~ "No Record Set"),
#     flat_date = as.Date(str_c(report_yr, "-01-01")) + (doy - 1))
# 
# 
# 
# hottest_days_plot <- hottest_days %>% 
#   mutate(record_period = ifelse(record_period == "New Record Daily Temperature", "New Record Daily Temperature", "No Record Set")) %>% 
#   ggplot() +
#   geom_tile(aes(x = flat_date, y = 0, fill = record_period)) +
#   scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(add = c(0,0))) +
#   scale_y_continuous(expand = expansion(add = c(0,0))) +
#   scale_fill_manual(values = c("purple", "gray90")) +
#   labs(x = "Day of the Year",
#        y = NULL,
#        title = "Record Setting Daily Temperatures of 2022",
#        subtitle = "Gulf of Maine",
#        fill = "Record Setting Year",
#        caption = NULL)  +
#    theme(legend.position = "top",
#          axis.text.y = element_blank(),
#          axis.ticks.y = element_blank()) +
#   guides(color = guide_legend(title.position = "top", title.hjust = 0.5))  
# 
# 
# hottest_days_decade <-  hottest_days %>% 
#   mutate(record_period = ifelse(record_period %in% c("Last Decade", "New Record Daily Temperature"), as.character(report_yr), "No Record Set")) %>% 
#   ggplot() +
#   geom_tile(aes(x = flat_date, y = 0, fill = record_period), show.legend = FALSE) +
#   scale_x_date(date_labels = "%b", date_breaks = "1 month", expand = expansion(add = c(0,0))) +
#   scale_y_continuous(expand = expansion(add = c(0,0))) +
#   scale_fill_manual(values = c("purple", "gray90")) +
#   labs(x = "Day of the Year",
#        y = NULL,
#        title = "Record Temperatures in the Last Decade",
#        subtitle = NULL,
#        fill = "Record Setting Year",
#        caption = str_c("Temperature data covers period of 1982-", report_yr))  +
#    theme(legend.position = "top",
#          axis.text.y = element_blank(),
#          axis.ticks.y = element_blank()) +
#   guides(color = guide_legend(title.position = "top", title.hjust = 0.5)) 
# 
# 
# hottest_days_plot / hottest_days_decade
}
```





```{r}
#| label: fig-percent-records-bars
#| fig-cap: "An illustration of the percentage of days during each month in 2022 when a record-high temperature was observed in the Gulf of Maine (e.g., 80% — 24 days — of November were new record-setting high temperatures)."
#| fig.height: 4

# Fraction of each month that were set


# What days set records in report_yr
records_report_yr <- hottest_days %>% 
  filter(yr == report_yr) %>% #count(month)
  mutate(record_temp = TRUE) %>% 
  select(yr, time, sst, seas, sst_anom, record_temp)

# What percent of days in month were records:
record_percents <- region_hw %>% 
  filter(yr == report_yr) %>% 
  full_join(records_report_yr) %>% 
  mutate(record_temp = ifelse(record_temp == TRUE, TRUE, FALSE)) %>% 
  group_by(yr, month) %>% 
  summarise(n_days = n(),
            n_records = sum(record_temp, na.rm = T),
            .groups = "drop") %>% 
  mutate(percent_records = (n_records / n_days),
         highlight_oct = ifelse(month == "Nov", "flag", "filler")) %>% 
  ggplot(aes(month, percent_records, #fill = highlight_oct)#, 
             alpha = highlight_oct)
         ) +
  geom_col() +
  geom_col(fill = as.character(gmri_cols("orange"))) +
  scale_alpha_manual(values = c("flag" = 1, "filler" = 0.4)) +
  scale_y_continuous(labels = percent_format(), limits = c(0,1), expand = expansion(add = c(0,0))) +
  theme(legend.position = "none") +
  labs(x = "Month", 
       y = "Fraction of the Month", 
       title = str_c(report_yr, ": Record Warm Daily Temperatures"),
       subtitle = "Large fraction of Nov. & Dec. dates set all-time highs for the Gulf of Maine ")   +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.1, y_npc = -0.18, logo_height = 1, height_units = "cm")


# Show plot
record_percents

```

## More Persistent, Intense Heatwaves


A marine heatwave (MHW) is defined as a period when there are five or more consecutive days when the observed SST is greater than the 90th percentile of the long-term average for that day. Subsequent MHW periods that are separated by two or fewer daily average SSTs that fall below the threshold are still considered part of a MHW event. @fig-annual_hw_timeseries shows the long-term average SST (gray line labeled "Climatological Mean"), as well as the threshold for a MHW day (gray line labeled "Heatwave Threshold"). The solid line and shaded area in the figure shows the observed SST for that day and its distance from the climatological average; when it is red, the Gulf of Maine is experiencing MHW conditions.

The Gulf of Maine met the criteria for MHW status for `r hw_criteria` days in `r report_yr`. While the warmest temperatures of the year were experienced in August (as we would expect from the climatology for the region), the largest temperature anomalies were observed during the last days of November, when they breached `r floor(report_yr_max_anom)` °F above the long-term average.


Comparing daily SST anomalies and MHW status for `r report_yr` to the long-term record (@fig-mhw-heatmap), as in the figure below — it becomes clear that the frequency, duration, and intensity of MHWs has not only increased in the past decade. In a world without human-caused climate change, we would expect, positive (warm) and negative (cool)  SST anomalies to more or less balance out over the span of several years, as various patterns of natural climate variability alternate having a dominant influence in Earth’s climate (e.g., La Niña vs El Niño). What is being observed in the Gulf of Maine (and elsewhere around the world), however, is a loss of that balance: larger fractions of recent years are experiencing above average temperatures and cold spells are becoming vanishingly rare.



```{r}
#| label: fig-mhw-heatmap
#| fig-cap: "Heat map of daily SST anomalies from the beginning of 1982 through the end of 2022. Not only do more large warm anomalies (darker reds) appear more frequently in recent years, but the frequency and duration of marine heatwave events (black lines) in the Gulf of Maine has become more pronounced in the past decade."
#| fig-alt: "A figure displays the temperatures for each day of year as a colored stripe, organized with a row for each day such that the day of year aligns vertically. The lower two-thirds has a roughly equal balance between colors for warm (red) and cold (blue) temperature anomalies. The top third of the image is almost completely red as temperature anomalies rarely fall below the long-term average. Black dots are overlaid onto days that meet the criteria for a heatwave, they are rare in the lower section and common in the red section."
#| fig.height: 8


# Heatmap: All years:
heatmap_p <- region_hw %>% 
  heatwave_heatmap_plot(temp_units = "F", start_yr = 1981, end_yr = 2022) +
  labs(title = "Gulf of Maine Heatwave Record",
       y = "Year",
       x = "Month") +
  theme(legend.position = "bottom",
        legend.title = element_text(angle = 0),
        plot.margin = margin(b = 0, t = 20)) +
  guides(fill = guide_colorbar(
    title = "Sea Surface Temperature Anomaly",
    title.position = "top", 
    title.hjust = 0.5, barwidth = unit(3.5, "in"), 
    barheight = unit(0.5, "cm"), 
    frame.colour = "black", 
    ticks.colour = "black")) +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.9, y_npc = 1.05, logo_height = 1, height_units = "cm")

heatmap_p + labs(caption = "")
```

## Warming Rate


```{r}
#| label: code-warming-rate-raster


# # 1. Warming Rates and Rankings
rates_path <- str_c(oisst_path, "warming_rates/annual_warming_rates1982to2022.nc")
rates_nc <- stack(rates_path, varname = "annual_warming_rate")
ranks_nc <- stack(rates_path, varname = "rate_percentile")



# Get the rank information that go with each area
ranks_masked <- mask_nc(ranks_nc, region_extent)
rates_masked <- mask_nc(rates_nc, region_extent)
region_ranks <- get_masked_vals(masked_ranks = ranks_masked, 
                                masked_rates = rates_masked, 
                                in_fahrenheit = T)


```



```{r}
#| label: txt-warming-rate-raster


# Prep it for text input.
avg_rank <- region_ranks$`Mean Rank` * 100
avg_rate <- region_ranks$`Mean Rate`
low_rank <- region_ranks$`Min Rank` * 100
low_rate <- region_ranks$`Min Rate`
top_rank <- region_ranks$`Max Rank` * 100
top_rank <- ifelse(top_rank == 100, "greater than or equal to 99.5", top_rank)
top_rate <- region_ranks$`Max Rate`

```



GMRI research was the first to reveal that the Gulf of Maine has been warming faster than the vast majority of the world’s ocean. @fig-warming-rank-map  updates this historical analysis by including data for 2022. Indeed, the story has not changed: the Gulf of Maine is warming faster than ~`r round(avg_rank, 2)`% of the world’s ocean. The reason why this is so important is because the rate of change can have profound consequences for the biology of individual species and for the integrated ecology of entire food webs. 




```{r}
#| label: fig-warming-rank-map
#| fig-cap: "Maps illustrating the rate of warming (from 1982 through 2022) displayed as their relative percentiles for [left] the northwest Atlantic Ocean and [right] entire global ocean surface."
#| fig.width: 8

# Map of fastest Warming Regions:
# #| column: page-inset-left

# Idea: Global Map & Highlighted Regions
# Fix Colors


# for an expanded view need to use the global ranks
world_rank_stars <- st_as_stars(rotate(ranks_nc))

# Reclassify raster into 10% Bins
world_ranks_binned <- world_rank_stars %>% 
  mutate(
    rank = Annual.Sea.Surface.Temperature.Warming.Rate.Rank,
    rank_bin = cut(rank, breaks = seq(0, 1, by = 0.1))) %>% 
  select(-c(rank, Annual.Sea.Surface.Temperature.Warming.Rate.Rank))

# # Define the number of colors you want
# nb.cols <- 10
# mycolors <- RColorBrewer::brewer.pal(n = 10, "Reds")


# Northeast Shelf
global_ranks_zoom <- ggplot() +
  geom_stars(data = world_ranks_binned) +
  geom_sf(data = new_england, color = "white", fill = "gray30", linewidth = .05) +
  geom_sf(data = canada, color = "white", fill = "gray30", linewidth = .05) +
  geom_sf(data = greenland, color = "white", fill = "gray30", linewidth = .05) +
  scale_y_continuous(breaks = seq(-90, 90, by = 5)) +
  scale_x_continuous(breaks = seq(-180, 180, by = 10))+
  scale_fill_brewer(palette = "Spectral", direction = -1, na.value = "transparent") +
  map_theme() +
  labs(title = NULL, subtitle = NULL) +
  coord_sf(
    xlim = c("xmin" = -77, "xmax" = -50),
    ylim = c("ymin" = 32, "ymax" = 48),
    expand = T) +
  guides("fill" = guide_colorsteps(
    title = str_c("Warming Rate Percentile: 1982-", report_yr),
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(4, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# Global View
global_ranks <- ggplot() +
  geom_stars(data = world_ranks_binned) +
  geom_sf(data = world_sf, color = "white", fill = "gray30", linewidth = .1) +
  scale_fill_brewer(palette = "Spectral", direction = -1, na.value = "transparent") +
  labs(title = NULL, subtitle = NULL) +
  map_theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()) +
  coord_sf(expand = F) +
  scale_y_continuous(breaks = seq(-90, 90, by = 30)) +
  guides("fill" = guide_colorsteps(
    title = str_c("Warming Rate Percentile: 1982-", report_yr),
    title.position = "top", 
    title.hjust = 0.5,
    barwidth = unit(4, "in"),
    frame.colour = "black",
    ticks.colour = "black")) 


# plot map
ranks_global_figure <- (global_ranks_zoom | global_ranks ) + 
  plot_annotation(title = "Rates of Ocean Temperature Change", 
                  caption = "Data Source: OISSTv2") +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom",
        plot.margin = margin(t = 5, r = 5, l = 10, b = 5))


# plot patchwork
ranks_global_figure
grid::grid.raster(lab_logo, x = 0.115, y = 0.115, just = c('left', 'bottom'), height = unit(1, "cm"))

```


## Concluding Thoughts:

In recent years, SSTs in the Gulf of Maine (and adjacent waters) have increased significantly and the rate of warming has accelerated. Combined, these climate-related impacts are beginning to have profound consequences for the marine ecology in the Gulf of Maine, with implications for the people and communities who rely on marine resources for their livelihoods and well-being. Research suggests there are two major factors driving these trends, one oceanographic and one atmospheric:


1. **Oceanographic:** The Gulf of Maine is influenced by two primary oceanic currents: the Labrador Current and the Gulf Stream. Historically, there was a stronger influence of the Labrador Current on the water in the Gulf of Maine. That stronger flow helped keep the heat transported northward along the U.S. East Coast by the Gulf Stream further out to sea. More recently, scientists have observed changes occurring in the interplay between the Labrador Current and Gulf Stream. More Arctic-origin freshwater from melting sea ice and land-based ice is constricting the southward flow of the Labrador Current. This is allowing the Gulf Stream—which transports warm water from lower latitudes to spread out more at latitudes around the Gulf of Maine. This, in turn, allows warmer water, more heavily influenced by the Gulf Stream to “spill over” into the Gulf of Maine.


2. **Atmospheric:** The North Atlantic Oscillation (NAO)— a large-scale pattern of natural variability in the atmosphere — has been in a “positive phase” more frequently over the past 10-15 years, meaning a more zonal (i.e., west-to-east) atmospheric pattern has dominated in the region. The relative decrease in “waviness” of the jet stream inhibits more consistent intrusions of cooler, Arctic air into the region.

While the oceanographic driver of the observed SST change in the Gulf of Maine has clear links to human-caused climate change, more research is needed to ascertain whether the NAO is experiencing significant changes as a result of anthropogenic climate change. Regardless, the atmospheric and oceanic conditions of the past several years combine to reveal an unambiguous trend: The Gulf of Maine is experiencing unprecedented warmth—and is heating up at a rate that is faster than almost anywhere else in the world’s oceans. 

---

**A Note on Data Sources**     

 * NOAA High Resolution SST data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html.    

**Recommended Citation:**    

> Gulf of Maine Research Institute. 2022. Gulf of Maine Warming Update: 2022 . https://gmri.org/stories/warming-22




## Technical Report Preview

The study of Marine heatwaves is a new area of study that began with GMRI scientists bringing public attention to the extreme temperatures in the Gulf of Maine during 2012. Since that original work, scientific debate has emerged on how "best" to identify and measure the impacts of these rare events. GMRI scientists are part of an international community that is actively studying different approaches to identifying and studying the impacts of these extreme events. 








---
# OLD ASSETS:



##  Temperature Anomaly Horizons

One way to think about the severity of these changes is to think about temperature horizons. A temperature horizon captures how long temperatures remain above certain thresholds. Each threshold is designated its own temperature, and in this way we can see how long within a year temperatures remained: 1 to 2 or as much as 4$^oC$ above normal.  



```{r}
#| label: fig-horizons-all
#| fig-cap: "Temperature anomaly horizons for all years of available satellite data (1982-2022). Horizons display how long temperatures were above certain thresholds (horizon width) and are colored by the strength of these events."
#| fig-alt: "Horizon plot of all years and their temperature anomaly horizons"
#| fig.height: 8
#| fig.width: 8

# Filter outliers beyond 25th and 75th quantile
cutpoints <- region_hw %>% 
  mutate(
    outlier = between(
      anom_f,
      quantile(anom_f, 0.25, na.rm = T)-
        1.5 * IQR(anom_f, na.rm = T),
      quantile(anom_f, 0.75, na.rm = T)+
        1.5 * IQR(anom_f, na.rm = T))) %>% 
  filter(outlier)

# Set origin
ori <- 0

# Manually pick cut points
sca <- seq(-5, 5, length.out = 9)
sca_bins <- str_c(sca[1:(length(sca)-1)], " to ", sca[2:length(sca)], "\u00b0F")
sca_labels <- rev(sca_bins)



# Plot Horizon Plot Using All Years
gom_annual_horizons <- anom_horizon_plot(
    grid_data = filter(region_hw, yr <= report_yr), 
    origin = ori, 
    scale_cutpoints = sca) +
  labs(title = "Temperature Anomaly Horizons",
       x = "Month",
       subtitle = "How deviation from the norm has become the new-normal",
       caption = "Climatatological Reference Period: 1982-2011") 


# Plot it
gom_annual_horizons

```

## Comparing the Gulf's Two Hottest Years

If we pull the horizons of our two hottest years on record it makes it easier to contrast where either one experienced acute high temperature events, and where there were sustained periods of above average temperatures.

Early into `r report_yr` it was apparent that the year was on-par with the previous title-holder for warmest year on record.


```{r}
#| label: fig-hottest-year-horizons
#| fig-cap: "Two horizon plots compare the timing and duration of elevated temperatures in the two warmest years on record, 2021 and now 2022. The width of a color horizon indicates the duration that temperatures were above that threshold."
#| fig-alt: "Single year horizon plots comparing 2021 and 2022"
#| fig.height: 6

# cut out outliers and set cutpoints
cutpoints_2 <- region_hw %>% 
  filter(year %in% c(2021, report_yr)) %>% 
  mutate(
    outlier = between(
      anom_f,
      quantile(anom_f, 0.25, na.rm=T)-
        1.5*IQR(anom_f, na.rm=T),
      quantile(anom_f, 0.75, na.rm=T)+
        1.5*IQR(anom_f, na.rm=T))) %>% 
  filter(outlier)



# 2021
hori_2021 <- region_hw %>% 
  filter(year == "2021") %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  facet_wrap(~year, strip.position = "top")  +
  labs(y = "")
  
# report_yr
hori_report_yr <- region_hw %>% filter(year == report_yr) %>% 
  anom_horizon_plot(scale_cutpoints = cutpoints_2) +
  facet_wrap(~year, strip.position = "top") +
  theme(plot.caption = element_text(colour = "gray25", size = 6)) +
  labs(x = "Month",
       y = "",
       caption = "Climatatological Reference Period: 1982-2011")

# Put them together as one figure
comparison_horizons <- hori_2021 / hori_report_yr + 
  plot_layout(guides = "collect")

# Do a bunch of formatting
comparison_horizons <- comparison_horizons &
  theme(
    plot.margin = margin(t = 0, r = 0.5, b = 0, l = 0.5),
          
    #Legend
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines")) &
   guides(fill = guide_legend(
     title = "Temperature Anomaly",
     title.hjust = 0.5, 
     title.position = "top",
     label.position = "bottom", 
     nrow = 1, 
     byrow = T, 
     reverse = T)) &
  plot_annotation(
    title = "Comparing the Gulf of Maine's Two Hottest Years", 
    subtitle = "Duration of temperature anomaly horizons, colored by their strength")




# Plot it
comparison_horizons
grid::grid.raster(lab_logo, x = 0.04, y = 0.01, just = c('left', 'bottom'), height = unit(1, "cm"))
  
```


## The Balance of Hot and Cold

Another way to visualize the climate transition that we are observing is by looking at the fraction of each year spent in different temperature ranges. Under a steady climate we would expect over the long-term to spend similar amounts of the year experiencing relatively warm & cold temperatures. These periods would balance themselves out and we would on-average have experienced something similar to the long-term climate.

What we have been seeing in the Gulf of Maine recently has lost that sense of balance. Larger fractions of the year are shared by above average temperatures & cold spell events are becoming vanishingly rare.


```{r}
#| label: fig-streamplot
#| fig-cap: "A streamplot shows the fraction of each year on record that temperature anomalies fell within ranges further from the climate reference period average. Since 2010 the Gulf of Maine's temperatures shifted rapidly outside of this expected range and a large fraction of the year is now 2 degrees or more above that average (red)."
#| fig-alt: "Streamplot tracking fraction of each year spent in varying degrees of temperature anomalies"



# Set Factor Order for Bins
bin_order <- c(
  "Less than -2", 
  "-2 to -1",
  "Within 1",
  "1 to 2", 
  "Greater than 2")

# Set Colors
bin_colors <- RColorBrewer::brewer.pal(n = length(bin_order), 
                                       name = "RdBu")
bin_colors <- rev(bin_colors)

# Count Days in Bins
day_types <- region_hw %>% 
  mutate(
    `Less than -2`   = if_else(anom_f <= -2, TRUE, FALSE),
    `-2 to -1`     = if_else(between(anom_f, -2, -1), TRUE, FALSE),
    `Within 1`     = if_else(between(anom_f, -1, 1), TRUE, FALSE),
    `1 to 2`       = if_else(between(anom_f, 1, 2), TRUE, FALSE),
    `Greater than 2` = if_else(anom_f >= 2, TRUE, FALSE)) %>% 
  group_by(yr) %>% 
  summarise(
    `Less than -2`   = sum(`Less than -2`, na.rm = T),
     `-2 to -1`    = sum(`-2 to -1`, na.rm = T),
    `Within 1`     = sum(`Within 1`, na.rm = T),
    `1 to 2`       = sum(`1 to 2`, na.rm = T),
    `Greater than 2` = sum(`Greater than 2`, na.rm = T),
    .groups = "drop") %>% 
  pivot_longer(cols = "Less than -2":"Greater than 2", 
               names_to = "Anomaly Strength", 
               values_to = "day_total", 
               values_drop_na = FALSE) %>% 
  mutate(day_total = if_else(is.na(day_total), 0L, day_total),
         `Anomaly Strength` = factor(`Anomaly Strength`, levels = bin_order))






# Plot the Streamplot
hw_streamplot <- ggplot(day_types, aes(yr, y = day_total, fill = `Anomaly Strength`)) +
  geom_stream(type = "proportion") +
  scale_x_continuous(breaks = seq(1980,2030, by = 10), 
                     minor_breaks = seq(1975, 2030, by = 5), 
                     expand = expansion(add = c(0,0))) +
  scale_y_continuous(labels = scales::percent_format(), 
                     expand = expansion(add = c(0,0))) + 
  scale_fill_manual(values = setNames(bin_colors, bin_order)) + 
  labs(y = "Percent of Year", x = "Year", fill = "", 
       title = "A Warmer Gulf of Maine",
       subtitle = "An increasing fraction of each year experiencing warm temperature anomalies",
       caption = "Anomalies calculated using 1982-2011 climatology.") +
  theme(
    legend.position = "top",
    legend.key.height = unit(.5, "lines"),
    legend.key.width = unit(3.75, "lines"),
    plot.margin = unit(c(.5,1,.3,.5), "cm"),
    plot.title.position = "plot",
    legend.margin = margin(c(7,0,0,0)),
    legend.justification = "center") +
   guides(fill = guide_legend(title = "Temperature Anomaly \u00b0F",
                             title.hjust = 0.5,
                             title.position = "left",
                             label.position = "bottom",
                             nrow = 1,
                             byrow = T)) +
  coord_cartesian(clip = "off") +
  geom_logo(x_npc = 0.1, y_npc = -0.115, logo_height = 1, height_units = "cm")


# Show figure
hw_streamplot  
```










```{r}
#| label: figure-saving
#| results: hide


####  Saving Figures  ####
if(params$save_figures){
  
  
  # Gom VS Global - annual
  ggsave(plot = temp_simplified, filename = str_c(save_location, "gom_and_global_trends_annual.jpeg"),
         height = unit(6, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
  
  # GOM vs Global - Seasons
  ggsave(plot = panel_plot, filename = str_c(save_location, "gom_and_global_trends_seasonal.jpeg"),
         height = unit(8, "in"),
         width = unit(10, "in"),
         dpi = "retina", 
         bg = "white")
  
  
  
  # Annual Temperature Anomalies Map(s)
  ggsave(plot = annual_gom_map, 
         filename = str_c(save_location, "gom_annual_anom_map.jpeg"),
         height = unit(5, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
  # seasonal Temperature Anomalies Map(s)
  ggsave(plot = season_panels_all, 
         filename = str_c(save_location, "gom_seasonal_anom_map.jpeg"),
         height = unit(5, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
  
  # Warmest Years Plot
  ggsave(plot = season_panels_all, 
         filename = str_c(save_location, "Annual_temp_ranks_plot.png"),
         height = unit(5, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
    # Global Anomaly Map
    ggsave(plot = annual_anom_map, 
           filename = str_c(save_location, "global_annual_anom_map.jpeg"),
           height = unit(5, "in"),
           width = unit(8, "in"),
           dpi = "retina", 
           bg = "white")
  
  
  
  # # Record daily temperatures plot
  # ggsave(plot = hottest_days_plot,
  #        filename = str_c(save_location, report_yr, "_hottest_days.jpeg"),
  #        height = unit(5, "in"),
  #        width = unit(8, "in"),
  #        dpi = "retina", 
  #        bg = "white")
  
  
  
  # Percentage of months that set record temperatures bars
  ggsave(plot = record_percents,
         filename = str_c(save_location, report_yr, "_percent_record_bars.jpeg"),
         height = unit(4, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
  
  
  # Full Year Temperature Timeseries
  ggsave(plot = hw_temp_p, filename = str_c(save_location, "oisst_lastyr_temps.jpeg"),
         height = unit(5, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
  
  
  # Heatwave status heatmap
  ggsave(plot = heatmap_p, filename = str_c(save_location, "oisst_gom_heatmap.jpeg"),
         height = unit(8, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
  
  
  
  
  # global warming rankings
  ggsave(plot = ranks_global_figure, filename = str_c(save_location, "global_ranks_map.jpeg"),
         height = unit(5, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  




    # All year temperature horizons
  ggsave(plot = gom_annual_horizons, filename = str_c(save_location, "oisst_gom_horizons.jpeg"),
         height = unit(8, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")


  # # report_yr | 2021 horizons
  # ggsave(plot = comparison_horizons, filename = str_c(save_location, "gom_hottest_horizons.jpeg"),
  #        dpi = "retina", bg = "white")
  
  # streamplot
  ggsave(plot = hw_streamplot, filename = str_c(save_location, "gom_anomaly_streamplot.jpeg"),
         height = unit(6, "in"),
         width = unit(8, "in"),
         dpi = "retina", 
         bg = "white")
  
  
}



  
```



```{r}
#| label: additional-figures
#| eval: false

# Load ragg for re-scaling
library(ragg)



# make a lcal save path
new_save <- here::here("local_data", report_yr)

# Re-size them for large plotting
# temp_simplified: 6x8
# panel plot: 8x10



#-------JPEG

# Gom VS Global - annual
ragg::agg_jpeg(filename = str_c(new_save, "gom_and_global_trends_annual.jpeg"), 
              height = 18, 
              width = 24, 
              units = "in",
              res = 320, 
              scaling = 2)
plot(temp_simplified)
invisible(dev.off())


# GOM vs Global - Seasons
ragg::agg_jpeg(filename = str_c(new_save, "gom_and_global_trends_seasonal.jpeg"), 
              height = 24, 
              width = 30, 
              units = "in",
              res = 320, 
              scaling = 3)
plot(panel_plot)
grid::grid.raster(lab_logo, x = 0.08, y = 0.01, just = c('left', 'bottom'), height = unit(1, "cm"))
invisible(dev.off())


```



```{r}
#| results: asis
insert_gmri_footer()
```

