---
title: "Gulf of Maine Current Year SST Report"
author: 
    first_name: "Adam"
    last_name: "Kemberling"
    url: https://github.com/adamkemberling
    affiliation: Gulf of Maine Research Institute
    affiliation_url: https://www.gmri.org
description: | 
  Temperature tracking for the Gulf of Maine
date: "Updated on: `r Sys.Date()`"
format: 
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 2
    df-print: kable
    self-contained: true
---


```{r}
#| label: setup
#| include: false
#| echo: false

# Packages
library(lubridate) # date support
library(rts) # raster time series
library(raster)  # raster support
library(rnaturalearth) # coastline polygons
library(sf) # simple feature support
library(stars) # raster plotting with ggplot
library(gmRi) # styling for GMRI
library(scales) # axis labels
library(here) # project navigation
library(patchwork) # multiplot arrangement
library(tidyverse) # data wrangling and plotting
library(climetrics)




# Support Functions

# File Paths
mills_path <- cs_path("mills")
oisst_path <- cs_path("res", "OISST/oisst_mainstays")


# Polygons for mapping
new_england <- ne_states("united states of america", returnclass = "sf")
canada      <- ne_states("canada", returnclass = "sf")
world       <- ne_countries(returnclass = "sf")
greenland   <- ne_states(country = "greenland", returnclass = "sf")
```

# Climate Velocity Vectors

How do rates of SST warming change over space? Are there corridors and chokepoints in our area that would align with the expectations of species movements?


## Purpose

This document looks into applying the [climetrics R package](https://github.com/shirintaheri/climetrics) to regional sea surface temperatures in the Gulf of Maine. This is an exploratory analysis using open-source software and public data resources. Huge credit to the authors of the climetrics package.


Climetrics is an extensible and reproducible R package to spatially quantify and explore multiple dimensions of climate change. Six widely known metrics of climate change are implemented in the package including:

 1. Standardized local anomalies (sed);   
 2. Changes in probability of local climate extremes (localExtreme);   
 3. Change in area of analogous climates (aaClimate);   
 4. Change in the distance to analogous climates (daClimate),   
 5. Novel climates (novelClimate);   
 6. Climate Change Velocity (ve; dVelocity; gVelocity).   

This Report will focus on **6. Climate Change Velocity (ve; dVelocity; gVelocity).**

```{r}
# Load annual Data
annual_sst <- raster::stack(
  str_c(oisst_path, "yearly_averages/annual_sst_1982to2021.nc"), 
  .varName = "sst")


# Crop it to a manageable area
crop_x <- c(-76, -62)
crop_y <- c(37, 46)
```


```{r}
# Expanded area as bbox
crop_area <- structure(
    c(crop_x[[1]], crop_y[[1]], crop_x[[2]], crop_y[[2]]),  
    class = "bbox",
  crs = 4326) %>% 
  st_as_sfc() %>% 
  st_as_sf()

# Crop the sst
sst_crop <- mask_nc(ras_obj = annual_sst, mask_shape = crop_area, rotate = T)


plot(sst_crop$X1982)

```


```{r}
# Make a raster time series
date_vec <- names(sst_crop) %>% str_remove_all("X") %>% str_c("-01-01") %>% as.Date()

# use rts function in the rts package to make a raster time series:

# Average SST
sst_ts <- rts(sst_crop, date_vec) 

# Minimum SST

# Maximum SST

```



### Distance Based Climate Velocity

This method is a distance-based velocity that can be quantified based on one or multiple climate variables:


From Burrows et al. 2011:
 > The velocity of climate change (in km/year) was calculated as the ratio of the long-term temperature trend (in C/year) to the 2-dimensional spatial gradient in temperature (in C/km, calculated over a 3 deg x 3 deg grid), oriented along the spatial gradient...

```{r}
dv <- dVelocity(sst_ts , 
                t1='2000/2009',
                t2='2010/2019')
plot(dv,col = hcl.colors(100), 
     main='Distance-based Velocity of Climate Change for SST\nPeriods 2000-2009 to 2010 to 2019')
```

Velocities are measured in km/decade or km/time


### Compare to Warming Rates

```{r}
rates_path <- paste0(oisst_path, "warming_rates/annual_warming_rates")
rates_stack <- stack(str_c(rates_path, "1982to2022.nc"), 
                     varname = "annual_warming_rate")

# Crop Stack
rates_crop <- mask_nc(rates_stack, mask_shape = crop_area)
plot(rates_crop, main = "Annual Warming Rate - 1982 to 2022")
```


### Map Mean SST for Periods

```{r}
mean(subset(sst_ts,"2000/2009")@raster) %>% plot(main = "2000-2009 Average SST")
```

```{r}
mean(subset(sst_ts,"2010/2019")@raster) %>% plot(main = "2010-2019 Average SST")
```



### Distance Based Climate Velocity: Multiple Metrics

```{r}

```


### Distance to Analagous Climate

This runs on multiple variables, maybe save for later

```{r}

da <- daClimate(
  precip = sst_ts, 
  tmin = tmin.t, 
  tmax = tmax.t, 
  tmean = tmean.t, 
  t1='1991/2000',t2='2010/2020')

cl <- colorRampPalette(c("MidnightBlue","blue","Turquoise","lightblue","gray","yellow","red"))

plot(da,col=cl(100),main='Changes in Distance to Analogous Climate')

```

# Compare to Warming Rates

```{r}
# # 1. Warming Rates and Rankings
rates_path <- str_c(oisst_path, "warming_rates/annual_warming_rates1982to2022.nc")
rates_nc <- stack(rates_path, varname = "annual_warming_rate")




```

