---
title: "North Sea Warming"
author: "Adam Kemberling"
url: "https://github.com/adamkemberling"
affiliation: "Gulf of Maine Research Institute"
affiliation_url: "https://www.gmri.org"
format: 
  html:
    toc: true
    self-contained: true
execute: 
  echo: false
  warning: false
  message: false
  comment: ""
  fig.align: center
---

# North Sea Warming


**Request:**

Could you take the data in the attached Excel file and help fill in the numbers for the following statement:

 > Data from the Helgoland Roads observation station show a warming trend of ?? C per decade since 1950. For the most recent decade (2015-2024), annual SSTs have averaged ????, while during an earlier decade (????-????), they averaged ??, representing an increase of ?? C.

```{r}
library(tidyverse)
library(gmRi)
library(here)


```


### Results

Surface temperature data for the North Sea shows the following changes:

```{r}
# north sea data from ICES
nsea <- read_csv(here("local_data","North_Sea_HelgolandRoads_Annual/North_Sea_HelgolandRoads_Annual.csv"), skip = 20) %>% 
  janitor::clean_names() %>% 
  mutate(decade = floor_decade(year))


ggplot(nsea, aes(year, temperature_c)) +
  geom_line() +
  theme_gmri() +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0C")) +
  labs(title = "North Sea, Helgolands Roads",
       subtitle = "Annual Temperature",
       y = "Temperature",
       x = "Year")
```


 
```{r}

# Just get the rate, multiply by ten
nsea_lm <- lm(temperature_c ~ year, data = filter(nsea, year >= 1950))
decadal_rate <- round(coef(nsea_lm)[[2]] * 10, 2)

# # Group on decades?
# nsea_decades <- nsea %>% #tail()
#   filter(year >= 1950) %>% 
#     group_by(decade) %>% 
#     summarise(across(-c(year), .fns = mean))

```


Data from the Helgoland Roads observation station show a warming trend of `r decadal_rate` C per decade since 1950.

```{r}
nsea_recent <- nsea %>% 
  filter(between(year, 2015,2024)) %>% 
  summarise(
    min_year = min(year),
    max_year = max(year),
    mean_temp = mean(temperature_c), 
    min_temp = min(temperature_c),
    max_temp = max(temperature_c))


nsea_early <- nsea %>% 
  filter(between(year, 1950,1960)) %>% 
  summarise(
    min_year = min(year),
    max_year = max(year),
    mean_temp = mean(temperature_c), 
    min_temp = min(temperature_c),
    max_temp = max(temperature_c))

# Change from early to late
temp_change <- round(nsea_recent$mean_temp - nsea_early$mean_temp, 2)


# Group on decades?
nsea_decades <- nsea %>% #tail()
  filter(year >= 1950) %>%
    group_by(decade) %>%
    summarise(
      min_year = min(year),
      max_year = max(year),
      mean_temp = mean(temperature_c), 
      min_temp = min(temperature_c),
      max_temp = max(temperature_c)
    )



```


For the most recent decade (`r nsea_recent$min_year`-`r nsea_recent$max_year`), annual SSTs have averaged `r round(nsea_recent$mean_temp, 2)`, while during an earlier decade (`r nsea_early$min_year`-`r nsea_early$max_year`), they averaged `r round(nsea_early$mean_temp, 2)`, representing an increase of `r temp_change` C.
