---
title: "Casco Baselines"
author: "Adam Kemberling"
url: "https://github.com/adamkemberling"
affiliation: "Gulf of Maine Research Institute"
affiliation_url: "https://www.gmri.org"
description: | 
  Exploring how the Temperatures of the Gulf of Maine Interact with Biological Limits
date: "Updated on: `r Sys.Date()`"
format: 
  html:
    toc: true
    self-contained: true
execute: 
  echo: false
  warning: false
  message: false
  comment: ""
  fig.align: center
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: packages
#| include: false
#| echo: false
#| warning: false

####  Packages  ####
library(lubridate)
library(raster)
library(here)
library(rnaturalearth)
library(scales)
library(stars)
library(sf)
library(heatwaveR)
library(patchwork)
library(showtext)
library(gmRi)
library(rcartocolor)
library(tidyverse)

# Support Functions
source(here("R/oisst_support_funs.R"), verbose = FALSE)
source(here("R/temp_report_support.R"), verbose = FALSE)

# Box paths
oisst_path <- cs_path("res", "OISST/oisst_mainstays")

# Polygons for mapping
new_england <- ne_states("united states of america", returnclass = "sf")



```

```{r}
#| label: style-sheet
#| results: asis

# Use GMRI style
use_gmri_style_rmd()

```

```{r}
#| label: fonts-config
#| echo: false

# # Using Raleway Font because it works
# sysfonts::font_add_google("Raleway")

# Path to the directory containing the font file (replace with your actual path)
font_dir <- paste0(system.file("stylesheets", package = "gmRi"), "/GMRI_fonts/Avenir/")

# Register the font
font_add(
  family = "Avenir",
  file.path(font_dir, "LTe50342.ttf"),
  bold = file.path(font_dir, "LTe50340.ttf"),
  italic = file.path(font_dir, "LTe50343.ttf"),
  bolditalic = file.path(font_dir, "LTe50347.ttf"))

# Load the font
showtext::showtext_auto()


```

# Objectives:

Look at what kind of seasonality changes we can track for Casco Bay,
either from buoys/moorings/models.

-   National Data Buoy Center Database & the Portland NOAA Tide Gauge
    Station
-   FVCOM Hindcast?

## Notes from the 2023 ecosystem report:

The Casco Bay ecosystem is changing as a result of ocean warming. In 10
years of monitoring, we have caught 50 unique species in our beach
seine.

1.  Our eight most common catches are green crab, Atlantic silverside,
    winter flounder, mummichog, alewife, Atlantic herring, Atlantic
    tomcod, and American sand lance.
2.  In colder years, Atlantic herring is the most commonly caught fish,
    followed by Atlantic silverside. This relationship flips in warmer
    years.
3.  There is clear evidence that warming ocean temperatures in the Gulf
    of Maine are adversely affecting cold-adapted native species,
    particularly Atlantic tomcod and Atlantic herring.
4.  We are tracking the recent appearance and increase in abundance of
    permit, crevalle jack, white mullet, and summer flounder in Casco
    Bay. 5. These species are more commonly found in warmer regions to
    the South, but are now appearing in Casco Bay.
5.  Fishermen observations play a key role in our understanding of Casco
    Bay ecosystem changes (see the On the Water section of this report).

## Water Temperatures in Casco Bay

#### National Data Buoy Center:

This data was downloaded from here:
https://www.ndbc.noaa.gov/to_station.shtml and entering Portland's
station ID: CASM1

Which will take you here:
https://www.ndbc.noaa.gov/station_history.php?station=casm1

But data can also be filtered and obtained from here if we want other
pieces of information
https://tidesandcurrents.noaa.gov/stationhome.html?id=8418150

Or we could use {noaaoceans}

```{r}
#| label: noaaoceans fail
#| eval: false

# # Load the noaaoceans
# # Wont work for me because of install issues b/c python conflicts
# library(noaaoceans)
# 
# # Get a list of all the stations.
# station_df <- list_coops_stations()
# 
# # Filter to stations in Washington with Water Temp Sensor
# station_df %>% 
#     filter(station_state == 'ME', station_names == "Portland")
# 
# # Create an empty storage data frame. 
# water_temp <- data.frame()
# 
# # Call the API for the Portland station
# water_temp <- query_coops_data(
#   station_id = "8418150",
#   start_date = '20181201',
#   end_date = '20181231',
#   data_product = 'water_temperature',
#   interval = 'h') # hourly readings
#     

```

```{r}
# Read yearly tables - from ndbc station history archive
portland_buoy <- list.files(here::here("local_data/portland_buoy"), full.names = T, pattern = ".txt") %>%
  map_dfr(~read_table(
    .x , 
    col_types = cols(
      `#YY` = col_character(), MM    = col_character(), DD    = col_character(),
      hh    = col_character(), mm    = col_character(), WDIR  = col_character(),
      WSPD  = col_character(), GST   = col_character(), WVHT  = col_character(),
      DPD   = col_character(), APD   = col_character(), MWD   = col_character(),
      PRES  = col_character(), ATMP  = col_character(), WTMP  = col_character(),
      DEWP  = col_character(), VIS   = col_character(), TIDE  = col_character())))



# Clean up the inconsistent column names
portland_buoy <- portland_buoy %>%
  mutate(
    `#YY` = as.character(`#YY`),
    PRES  = as.character(PRES),
    WDIR  = as.character(WDIR)) %>%
  mutate(
    YYYY = ifelse(is.na(YYYY) & !is.na(`#YY`),`#YY`, YYYY),
    BAR = ifelse(is.na(BAR) & !is.na(PRES),PRES, BAR),
    WD = ifelse(is.na(WD) & !is.na(WDIR), WDIR, WD),
    MM = ifelse(is.na(MM) & !is.na(mm), mm, MM),
    date = as.Date(str_c(YYYY, MM, DD, sep = "-")))  %>%
  select(-c(`#YY`, PRES, WDIR, mm)) %>% 
  filter(YYYY != "#yr")%>% 
  mutate(
    YYYY = as.numeric(YYYY),
    WSPD  = as.numeric(WSPD),
    GST   = as.numeric(GST),
    WVHT  = as.numeric(WVHT),
    DPD   = as.numeric(DPD),
    APD   = as.numeric(APD),
    MWD   = as.numeric(MWD),
    ATMP  = as.numeric(ATMP),
    WTMP  = as.numeric(WTMP),
    DEWP  = as.numeric(DEWP),
    VIS   = as.numeric(VIS),
    TIDE  = as.numeric(TIDE),
    # So apparently 9999 etc is used as NA
    # Make 99's actual NA's
    WD    = ifelse(WD == 999, NA, WD),
    WSPD  = ifelse(WSPD == 99, NA, WSPD),
    GST   = ifelse(GST == 99, NA, GST),
    WVHT  = ifelse(WVHT == 99, NA, WVHT),
    DPD   = ifelse(DPD == 99, NA, DPD),
    APD   = ifelse(APD == 99, NA, APD),
    MWD   = ifelse(MWD == 999, NA, MWD),
    BAR   = ifelse(BAR == 9999, NA, BAR),
    ATMP  = ifelse(ATMP == 999, NA, ATMP),
    WTMP  = ifelse(WTMP == 999, NA, WTMP),
    DEWP  = ifelse(DEWP == 999, NA, DEWP),
    VIS   = ifelse(VIS == 99, NA, VIS),
    TIDE  = ifelse(TIDE == 99, NA, TIDE))

# Load 2024 Since its needed to be downloaded elsewhere
portland_buoy_24 <- list.files(here::here("local_data/portland_buoy/tidesandcurrents_portal"), full.names = T, pattern = ".csv") %>%
  map_dfr(~read_csv(.x)) %>% 
  select(date = `Date`,
         WTMP = `Water Temp (°C)`) %>% 
  mutate(YYYY = 2024,
         MM = str_pad(lubridate::month(date), side = "left", pad = "0", width = 2),
         DD = str_pad(lubridate::day(date), side = "left", pad = "0", width = 2))

# Combine
portland_buoy <- bind_rows(portland_buoy, portland_buoy_24)

# # Plot the Seasonal Cycle w/ Hourly Data
# ggplot(portland_buoy) +
#   geom_line(aes(date, WTMP))
```

### Water Temperature in Portland Harbor

```{r}
# Distibution of temperatures for the period we have CBASS data
portland_buoy %>% 
  # filter(YYYY > 2013) %>% 
  ggplot() +
  geom_boxplot(aes(YYYY, WTMP, group = YYYY)) +
  scale_x_continuous(breaks = seq(1990,2026, 2))
```

# Summer Sampling Period Temperatures

May - September CBASS study period Usually second week in June, but
sometimes in late may

```{r}
library(mgcv)
seasonal_mod <- gam()
```

```{r}
# Plot where 2024 sits on the CBASS Months
portland_buoy %>% 
  filter(!(YYYY==2011 & MM == "06")) %>% 
  group_by(YYYY, MM) %>% 
  summarise(
    min = min(WTMP, na.rm = T),
    max = max(WTMP, na.rm = T),
    mean  = mean(WTMP, na.rm = T), 
    .groups= "drop") %>% 
  pivot_longer(
    cols = -c(YYYY, MM), 
    names_to = "metric", 
    values_to = "temp") %>% 
  mutate(date = as.Date(
    str_c(YYYY, MM, "15", sep = "-"))) %>% 
  dplyr::filter(
    YYYY >= 2013,
    MM %in% c("04", "05", "06", "07", "08")) %>% 
  ggplot() +
  geom_line(aes(date, temp, color = metric), linewidth = 1) +
  facet_wrap(~MM) +
  scale_color_gmri()
```

## Current (2023) Baseline

-   Herring Population Trends
-   Lobster
-   Stratification Trends
-   Maine Coastal Current Behavior

8 most frequent species in CBASS herring, alewife, green crab,
silverside, mummichog, winter flounder

## How to help Katie

Could do some feature engineering of water temperature information to
make it species specific: - Growth degree days - days within species
preferred, days outside -

```{r}

```

# FVCOM

Don't have FVCOM through 2024 so we can't really use it here.
